---
title: "Step 09: Integration Preview v3 - Focused Metadata with Cell Type Labels"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 10,
  fig.height = 8
)

# Load required libraries
suppressPackageStartupMessages({
  library(Seurat)
  library(ggplot2)
  library(dplyr)
  library(presto)
  library(patchwork)
  library(viridis)
  library(yaml)
})

cat("✓ Libraries loaded\n")
```

# Setup Project Root

```{r project_root}
# Use here package to navigate to preprocessing directory
if (!require("here", quietly = TRUE)) {
  stop("Install 'here' package: install.packages('here')")
}

# PROJECT_ROOT is set by the wrapper from CFG_PROJECT_ROOT. The here::here
# call is a legacy fallback for interactive reruns outside the wrapper and
# is not exercised by the canonical pipeline.
PROJECT_ROOT <- here::here("project", "01_Preprocessing")

cat("Project root:", PROJECT_ROOT, "\n")

# Verify it's the right directory (should have config/ and outputs/)
if (!dir.exists(file.path(PROJECT_ROOT, "config"))) {
  stop("PROJECT_ROOT doesn't have config/ directory. Current: ", PROJECT_ROOT)
}

cat("✓ Project root verified\n")
```

# Load Configuration Files

```{r load_configs}
# Load module configurations
config_path <- file.path(PROJECT_ROOT, "config/module_configs.yaml")
if (!file.exists(config_path)) {
  stop("Config file not found: ", config_path)
}
config <- yaml::read_yaml(config_path)

# FIX: Correct schema path (config$modules$step_09_...)
step_09_config <- config$modules$step_09_integration_preview

if (is.null(step_09_config)) {
  stop("step_09_integration_preview not found in config$modules")
}

cat("✓ Loaded module_configs.yaml\n")

# Load preprocessing manifest
manifest_path <- file.path(PROJECT_ROOT, "config/preprocessing_manifest.yaml")
if (!file.exists(manifest_path)) {
  stop("Preprocessing manifest not found: ", manifest_path)
}
manifest <- yaml::read_yaml(manifest_path)

cat("✓ Loaded preprocessing_manifest.yaml\n")

# Load raw data manifest (up one level from preprocessing)
raw_manifest_path <- file.path(dirname(PROJECT_ROOT), "config/raw_data_manifest.yaml")
if (!file.exists(raw_manifest_path)) {
  stop("Raw data manifest not found: ", raw_manifest_path)
}
raw_manifest <- yaml::read_yaml(raw_manifest_path)

cat("✓ Loaded raw_data_manifest.yaml\n")
```

# Extract Step 08 Outputs

```{r extract_step08}
# Get Step 08 outputs location
step_08 <- manifest$integrated_objects$unfiltered$all_cells

if (is.null(step_08)) {
  stop("Step 08 outputs not found in manifest. Run Step 08 first.")
}

cat("✓ Step 08 outputs found in manifest\n")

# Extract paths
latent_path <- step_08$outputs$embeddings$latent
umap_path <- step_08$outputs$embeddings$umap
cluster_files <- unlist(step_08$outputs$clusters)

# Validate files exist
if (!file.exists(latent_path)) {
  stop("Latent embeddings not found: ", latent_path)
}
if (!file.exists(umap_path)) {
  stop("UMAP coordinates not found: ", umap_path)
}

cat("✓ Step 08 file paths validated\n")

# Get default resolution from config
default_resolution <- step_09_config$algorithm_params$default_resolution
cat("✓ Default resolution:", default_resolution, "\n")
```

# Load Step 08 Embeddings

```{r load_embeddings}
cat("\n=== Loading Step 08 Embeddings ===\n\n")

# PROVEN_PATTERN: LOGIC-09-001
# SOURCE: hbca_analysis/.../Epithelial_CompartmentIntegration_Filtered_20250814.Rmd:115-117
# Load latent embeddings
latent_embedding <- read.csv(latent_path, row.names = 1, check.names = FALSE)
cat("  Latent:", nrow(latent_embedding), "cells ×", ncol(latent_embedding), "dims\n")

# PROVEN_PATTERN: LOGIC-09-002
# SOURCE: hbca_analysis/.../Epithelial_CompartmentIntegration_Filtered_20250814.Rmd:119-121
# Load UMAP coordinates
umap_embedding <- read.csv(umap_path, row.names = 1, check.names = FALSE)
cat("  UMAP:", nrow(umap_embedding), "cells ×", ncol(umap_embedding), "dims\n")

# Validate alignment
if (!identical(rownames(latent_embedding), rownames(umap_embedding))) {
  stop("⚠️  Cell IDs don't match between latent and UMAP!")
}

cat("\n✓ Embeddings loaded\n")
```

# Define QC-Passing Cells

```{r qc_cells}
# All cells in Step 08 embeddings already passed QC (Steps 02+03)
passing_cells <- rownames(latent_embedding)

cat("✓ QC-passing cells:", length(passing_cells), "\n")
```

# Load Step 08 Integrated Metadata

```{r load_step08_metadata}
cat("\n=== Loading Step 08 Cell Metadata ===\n\n")

# NOVEL_LOGIC: GAP-001 (Step 08 metadata loading)
# SCHEMA_VALIDATED: .scratch/09/.schema/latent_50d.csv.schema.yaml confirms cell_id format

# Get Step 08 cell metadata path from manifest
step_08_metadata_path <- step_08$outputs$metadata$cell_metadata

if (is.null(step_08_metadata_path) || !file.exists(step_08_metadata_path)) {
  stop("Step 08 cell metadata not found: ", step_08_metadata_path)
}

# Load Step 08 metadata (contains sample_id for each cell)
step_08_cell_metadata <- read.csv(step_08_metadata_path, stringsAsFactors = FALSE)
step_08_cell_metadata$cell_id <- step_08_cell_metadata$X
# Validate cell_id column exists
if (!"cell_id" %in% colnames(step_08_cell_metadata)) {
  stop("cell_id column not found in Step 08 metadata")
}

cat("  Loaded Step 08 metadata:", nrow(step_08_cell_metadata), "cells\n")
cat("  Columns:", paste(colnames(step_08_cell_metadata), collapse = ", "), "\n")

# Extract sample list
samples_needed <- unique(step_08_cell_metadata$sample_id)
cat("\n  Samples detected:", length(samples_needed), "\n")
cat("  ", paste(samples_needed, collapse = ", "), "\n")

cat("\n✓ Step 08 metadata loaded\n")
```

# Extract Cluster Column Names from Step 08 Metadata

```{r define_clustering_columns}
cat("\n=== Extracting Clustering Column Names ===\n\n")

# FIX: Use clustering columns directly from Step 08 metadata (leiden_scvi_*)
# Step 08 metadata already contains all clustering resolutions
# No need to load separate cluster CSV files

# Extract clustering columns from Step 08 metadata
clustering_columns <- grep("^leiden_scvi_", colnames(step_08_cell_metadata), value = TRUE)

cat("  Found", length(clustering_columns), "clustering columns in Step 08 metadata:\n")
for (col in clustering_columns) {
  cat("    -", col, "\n")
}

# Define default cluster column using leiden_scvi_* format
default_cluster_col <- sprintf("leiden_scvi_%.1f", default_resolution)

# Verify it exists
if (!(default_cluster_col %in% clustering_columns)) {
  stop("Default cluster column '", default_cluster_col, "' not found!\n",
       "  Available: ", paste(clustering_columns, collapse = ", "))
}

cat("\n✓ Default cluster column:", default_cluster_col, "\n")
```

# Load Step 04 Label Transfer

```{r load_label_metadata}
cat("\n=== Loading Step 04 Cell Type Labels ===\n\n")

# NOVEL_LOGIC: Load reconciled HBCA labels from Step 04 label transfer
# All cells in Step 08 are singlets (doublet-filtered), so we only need cell type annotations

# Load label metadata per sample
label_list <- list()

for (sample_id in samples_needed) {
  label_file <- file.path(PROJECT_ROOT, "outputs/04_LabelTransfer/cell_metadata",
                          paste0(sample_id, "_labels.csv"))

  if (!file.exists(label_file)) {
    warning("Label file not found for ", sample_id, ": ", label_file)
    next
  }

  # Load label data
  label_data <- read.csv(label_file, stringsAsFactors = FALSE)

  # FIX: First column is ALREADY PREFIXED cell_id (e.g., "Pat1_P1_AAACCCAAGAAACCAT-1")
  # CSV schema: unnamed first column contains prefixed cell_id from Step 04 R output
  # DO NOT add prefix again (would create double-prefixed IDs)
  cell_id_col <- colnames(label_data)[1]
  label_data$cell_id <- label_data[[cell_id_col]]  # Use directly, NO prefix addition!

  # Select only HBCATransferredLabels.Kumar_2023 (reconciled labels)
  # Exclude: prediction scores, SC/SN predictions, cluster_coarse (not applicable at integrated context)
  label_list[[sample_id]] <- label_data %>%
    select(cell_id, HBCATransferredLabels.Kumar_2023)

  cat("  Loaded", nrow(label_data), "cells from", sample_id, "\n")
}

# Combine all label data
label_metadata <- bind_rows(label_list)
cat("\n✓ Total label metadata:", nrow(label_metadata), "cells\n")
```

# Load Step 03 Doublet Scores (Seurat handles QC metrics natively)

```{r load_step03_doublets}
cat("\n=== Loading Step 03 Doublet Scores ===\n\n")

# Load doublet scores per sample (same pattern as label loading)
doublet_list_v2 <- list()

for (sample_id in samples_needed) {
  doublet_file <- file.path(PROJECT_ROOT, "outputs/03_DoubletDetection/metadata",
                            paste0(sample_id, ".csv"))

  if (!file.exists(doublet_file)) {
    warning("Doublet file not found for ", sample_id, ": ", doublet_file)
    next
  }

  # Load doublet data
  doublet_data <- read.csv(doublet_file, stringsAsFactors = FALSE)

  # First column is cell_barcode (already prefixed)
  cell_id_col <- colnames(doublet_data)[1]
  doublet_data$cell_id <- doublet_data[[cell_id_col]]

  # Select only doublet score (all cells here passed filters, so class not needed)
  doublet_list_v2[[sample_id]] <- doublet_data %>%
    select(cell_id, scDblFinder_score, doublet_filter_pass)

  cat("  Loaded", nrow(doublet_data), "cells from", sample_id, "\n")
}

# Combine all doublet data
step_03_doublets <- bind_rows(doublet_list_v2)
cat("\n✓ Total doublet scores:", nrow(step_03_doublets), "cells\n")
```

# Map Anatomical Positions

```{r map_anatomical_positions}
cat("\n=== Mapping Anatomical Positions ===\n\n")

# NOVEL_LOGIC: GAP-002 (Anatomical position mapping)
# DIFFICULTY: LOW, PROJECT_IMPACT: LOW
# ADAPTS: raw_data_manifest.yaml anatomical structure

# Build position mapping from raw_data_manifest
position_map <- list()

for (patient_name in names(raw_manifest$patients)) {
  patient <- raw_manifest$patients[[patient_name]]

  for (pos_name in names(patient$positions)) {
    pos <- patient$positions[[pos_name]]
    anatomical <- pos$anatomical

    # Defensive check: anatomical section exists
    if (is.null(anatomical)) {
      warning("No anatomical data for ", patient_name, "/", pos_name)
      next
    }

    # Handle multi-valued positions (P1, some P2 positions have Upper+Lower)
    midline <- anatomical$midline_position

    # Vertical position handling
    vertical <- if (length(anatomical$vertical_position) > 1) {
      "Upper+Lower"  # P1 spans both
    } else {
      anatomical$vertical_position
    }

    # Horizontal position handling
    horizontal <- if (length(anatomical$horizontal_position) > 1) {
      "Inner+Outer"  # P1 spans both
    } else {
      anatomical$horizontal_position
    }

    # Store mapping (sample_id constructed from patient + position)
    sample_id <- paste0(patient_name, "_", pos_name)
    position_map[[sample_id]] <- list(
      midline_position = midline,
      vertical_position = vertical,
      horizontal_position = horizontal
    )
  }
}

# Convert to dataframe for merging
position_df <- bind_rows(lapply(names(position_map), function(sid) {
  data.frame(
    sample_id = sid,
    midline_position = position_map[[sid]]$midline_position,
    vertical_position = position_map[[sid]]$vertical_position,
    horizontal_position = position_map[[sid]]$horizontal_position,
    stringsAsFactors = FALSE
  )
}))

cat("✓ Mapped", nrow(position_df), "positions\n")
cat("\nPosition summary:\n")
print(position_df)
```

# Assemble Unified Metadata

```{r assemble_unified_metadata}
cat("\n=== Assembling Unified Metadata ===\n\n")

# NOVEL_LOGIC: GAP-003 (Metadata orchestration)
# ORCHESTRATION: Merge all metadata sources

# Start with Step 08 integrated metadata (already loaded)
unified_metadata <- step_08_cell_metadata

cat("Starting with Step 08 metadata:", nrow(unified_metadata), "cells\n")

# Merge Step 04 cell type labels
unified_metadata <- unified_metadata %>%
  left_join(label_metadata, by = "cell_id")

cat("After label merge:", nrow(unified_metadata), "cells\n")
cat("  Cells with cell type labels:", sum(!is.na(unified_metadata$HBCATransferredLabels.Kumar_2023)), "\n")

# DIAGNOSTIC: Check for missing Step 04 data
n_missing_labels <- sum(is.na(unified_metadata$HBCATransferredLabels.Kumar_2023))
if (n_missing_labels > 0) {
  cat("\n⚠️  WARNING:", n_missing_labels, "cells missing Step 04 label data!\n")

  # Identify which samples are affected
  missing_label_cells <- unified_metadata %>%
    filter(is.na(HBCATransferredLabels.Kumar_2023)) %>%
    select(cell_id, sample_id, patient_id)

  missing_label_samples <- unique(missing_label_cells$sample_id)
  cat("  Affected samples:", paste(missing_label_samples, collapse = ", "), "\n")

  # Check if these samples are in label_metadata
  for (samp in missing_label_samples) {
    n_in_labels <- sum(grepl(paste0("^", samp, "_"), label_metadata$cell_id))
    cat("  ", samp, "- Step 04 records:", n_in_labels, "\n")
  }
}

# Merge Step 03 doublet scores
unified_metadata <- unified_metadata %>%
  left_join(step_03_doublets, by = "cell_id")

cat("After doublet score merge:", nrow(unified_metadata), "cells\n")
cat("  Cells with doublet scores:", sum(!is.na(unified_metadata$scDblFinder_score)), "\n")

# Merge anatomical position data
unified_metadata <- unified_metadata %>%
  left_join(position_df, by = "sample_id")

cat("After position merge:", nrow(unified_metadata), "cells\n")
cat("  Cells with anatomical data:", sum(!is.na(unified_metadata$midline_position)), "\n")

# Validation
cat("\n=== Metadata Assembly Validation ===\n")
cat("  Total cells:", nrow(unified_metadata), "\n")
cat("  Cells with cell type labels:", sum(!is.na(unified_metadata$HBCATransferredLabels.Kumar_2023)), "\n")
cat("  Cells with doublet scores:", sum(!is.na(unified_metadata$scDblFinder_score)), "\n")
cat("  Cells with anatomical data:", sum(!is.na(unified_metadata$midline_position)), "\n")

# Check for unexpected NAs
na_counts <- sapply(unified_metadata, function(x) sum(is.na(x)))
if (any(na_counts > 0)) {
  cat("\n⚠️  NA counts per column:\n")
  print(na_counts[na_counts > 0])
}

# Save unified metadata
metadata_output <- file.path(PROJECT_ROOT, "outputs/09_IntegrationPreview/metadata")
dir.create(metadata_output, recursive = TRUE, showWarnings = FALSE)
write.csv(unified_metadata,
          file.path(metadata_output, "unified_metadata.csv"),
          row.names = FALSE)

cat("\n✓ Unified metadata saved to:", file.path(metadata_output, "unified_metadata.csv"), "\n")
```

# Load Raw Counts Per Sample

```{r load_counts}
cat("\n=== Loading Raw Counts Per Sample ===\n\n")

# PROVEN_PATTERN: LOGIC-09-006, LOGIC-09-007
# SOURCE: step_04_post_qc_vfs/source/post_qc_vfs.R:465-493

# Initialize
seurat_list <- list()
manifest_metadata <- data.frame(
  sample_id = character(),
  patient_id = character(),
  position = character(),
  stringsAsFactors = FALSE
)

# Define patients to process
patients <- c("Pat1", "Pat2", "UCI604", "UCI220228")

for (patient_id in patients) {
  patient_data <- raw_manifest$patients[[patient_id]]

  if (is.null(patient_data)) {
    warning("Patient ", patient_id, " not found in raw manifest")
    next
  }

  positions <- names(patient_data$positions)

  for (pos_name in positions) {
    pos_data <- patient_data$positions[[pos_name]]

    # Check for chromium data
    if (is.null(pos_data$chromium)) {
      next  # Skip silently if no chromium data
    }

    sample_id <- pos_data$chromium$sample_id
    h5_path <- pos_data$chromium$filtered_data_path

    if (is.null(sample_id) || is.null(h5_path)) {
      warning("Missing sample_id or h5_path for ", patient_id, "/", pos_name)
      next
    }

    if (!file.exists(h5_path)) {
      warning("H5 file not found: ", h5_path)
      next
    }

    # Load raw counts (LOGIC-09-006)
    counts_raw <- Seurat::Read10X_h5(h5_path)

    # Add sample_id prefix to match Step 08 format
    colnames(counts_raw) <- paste0(sample_id, "_", colnames(counts_raw))

    # Filter to QC-passing cells (LOGIC-09-007)
    sample_passing_cells <- passing_cells[grepl(paste0("^", sample_id, "_"), passing_cells)]
    cells_to_keep <- intersect(sample_passing_cells, colnames(counts_raw))

    if (length(cells_to_keep) == 0) {
      warning("No passing cells found for ", sample_id)
      next
    }

    # EXCEPTION-BASED LOGGING: Only report mismatches
    expected_cells <- length(sample_passing_cells)
    if (length(cells_to_keep) != expected_cells) {
      cat("⚠️  ", sample_id, ": ", length(cells_to_keep), "/", expected_cells,
          " cells (", expected_cells - length(cells_to_keep), " missing from H5)\n", sep = "")
    }

    # Subset counts
    counts_filtered <- counts_raw[, cells_to_keep]

    # Create Seurat object (LOGIC-09-009)
    seurat_obj <- CreateSeuratObject(
      counts = counts_filtered,
      project = sample_id,
      min.cells = 0,
      min.features = 0
    )

    # Store in list
    seurat_list[[sample_id]] <- seurat_obj

    # Track metadata
    manifest_metadata <- rbind(manifest_metadata, data.frame(
      sample_id = sample_id,
      patient_id = patient_id,
      position = pos_name,
      stringsAsFactors = FALSE
    ))
  }
}

cat("\n✓ Loaded", length(seurat_list), "samples\n")
```

# Merge Seurat Objects

```{r merge_objects}
cat("\n=== Merging Seurat Objects ===\n\n")

# NOVEL_LOGIC: GAP-09-001 (Seurat merge)
# DIFFICULTY: LOW, PROJECT_IMPACT: MEDIUM

if (length(seurat_list) == 0) {
  stop("No Seurat objects to merge!")
}

# Get first object
first_obj <- seurat_list[[1]]

# Merge remaining objects
if (length(seurat_list) > 1) {
  remaining_objs <- seurat_list[2:length(seurat_list)]

  seurat_merged <- merge(
    x = first_obj,
    y = remaining_objs,
    add.cell.ids = NULL,  # Cell IDs already prefixed
    project = "integrated"
  )
} else {
  seurat_merged <- first_obj
}

cat("✓ Merged:", ncol(seurat_merged), "cells ×", nrow(seurat_merged), "genes\n")

# EXCEPTION-BASED: Validate total cells
expected_cells <- sum(sapply(seurat_list, ncol))
if (ncol(seurat_merged) != expected_cells) {
  warning("⚠️  Cell count mismatch! Expected: ", expected_cells, ", Got: ", ncol(seurat_merged))
}
```

# Add Patient/Position Metadata

```{r add_metadata}
cat("\n=== Adding Patient/Position Metadata ===\n\n")

# FIX: Extract sample_id from cell barcodes instead of using orig.ident
# orig.ident contains only patient ID after merge, not full sample_id
# Cell format: {sample_id}_{barcode}, e.g., "Pat1_P1_AAACCCAAGAAACCAT-1"

extract_sample_id <- function(cell_names) {
  # Remove the trailing barcode pattern (ACGT-1)
  gsub("_[ACGT]+-[0-9]+$", "", cell_names)
}

cell_sample_ids <- extract_sample_id(colnames(seurat_merged))

# Add patient_id using extracted sample_ids
seurat_merged$patient_id <- manifest_metadata$patient_id[
  match(cell_sample_ids, manifest_metadata$sample_id)
]

# Add position using extracted sample_ids
seurat_merged$position <- manifest_metadata$position[
  match(cell_sample_ids, manifest_metadata$sample_id)
]

# EXCEPTION-BASED: Check for NAs
n_na_patient <- sum(is.na(seurat_merged$patient_id))
n_na_position <- sum(is.na(seurat_merged$position))

if (n_na_patient > 0) {
  warning("⚠️  ", n_na_patient, " cells have NA patient_id")
}
if (n_na_position > 0) {
  warning("⚠️  ", n_na_position, " cells have NA position")
}

cat("✓ Metadata added\n")
```

# Add Unified Metadata to Seurat Object

```{r add_unified_metadata}
cat("\n=== Adding Unified Metadata to Seurat Object ===\n\n")

# NOVEL_LOGIC: GAP-003 continuation (add assembled metadata to Seurat)

# Match cells between unified_metadata and Seurat object
seurat_cells <- colnames(seurat_merged)
metadata_cells <- unified_metadata$cell_id

# Find matching cells
matching_cells <- intersect(seurat_cells, metadata_cells)
cat("  Matching cells:", length(matching_cells), "/", length(seurat_cells), "\n")

# FIX: Add columns from unified_metadata
# INCLUDE: sample_id (authoritative source from Step 08)
# EXCLUDE: cell_id (Seurat rownames), X (duplicate of cell_id), position_id (redundant with position)
# EXCLUDE: patient_id, position (already added to Seurat at lines 648-655)
cols_to_add <- setdiff(colnames(unified_metadata), c("cell_id", "X", "patient_id", "position", "position_id"))

for (col in cols_to_add) {
  # Match order of Seurat cells
  seurat_merged@meta.data[[col]] <- unified_metadata[[col]][
    match(seurat_cells, unified_metadata$cell_id)
  ]

  # Report NA counts
  n_na <- sum(is.na(seurat_merged@meta.data[[col]]))
  if (n_na > 0) {
    cat("  ", col, ":", n_na, "NAs\n")
  }
}

cat("\n✓ Added", length(cols_to_add), "metadata columns to Seurat object\n")
cat("  New columns:", paste(cols_to_add[1:min(5, length(cols_to_add))], collapse = ", "))
if (length(cols_to_add) > 5) cat(", ...")
cat("\n")
```

# Add Step 08 Embeddings

```{r add_embeddings}
cat("\n=== Adding Step 08 Embeddings ===\n\n")

# PROVEN_PATTERN: LOGIC-09-004
# SOURCE: hbca_analysis/.../Epithelial_CompartmentIntegration_Filtered_20250814.Rmd:189-198

# Subset embeddings to match merged Seurat cells
merged_cells <- colnames(seurat_merged)
latent_subset <- latent_embedding[merged_cells, , drop = FALSE]
umap_subset <- umap_embedding[merged_cells, , drop = FALSE]

# Add as dimensional reductions
seurat_merged[["scvi"]] <- CreateDimReducObject(
  embeddings = as.matrix(latent_subset),
  key = "scvi_",
  assay = "RNA"
)

seurat_merged[["umap"]] <- CreateDimReducObject(
  embeddings = as.matrix(umap_subset),
  key = "UMAP_",
  assay = "RNA"
)

cat("✓ Added scvi (", ncol(latent_subset), "D) and umap (2D) reductions\n", sep = "")
```

# Verify Cluster Assignments

```{r verify_clusters}
cat("\n=== Verifying Cluster Assignments ===\n\n")

# FIX: Clustering columns already added from unified_metadata (Step 08 leiden_scvi_*)
# No need to add separately - just verify they exist

# Verify all clustering columns are present
missing_cols <- setdiff(clustering_columns, colnames(seurat_merged@meta.data))
if (length(missing_cols) > 0) {
  stop("Missing clustering columns in Seurat metadata: ", paste(missing_cols, collapse = ", "))
}

cat("✓ All", length(clustering_columns), "clustering columns present in metadata:\n")
for (col in clustering_columns) {
  n_clusters <- length(unique(seurat_merged@meta.data[[col]]))
  cat("  -", col, ":", n_clusters, "clusters\n")
}

# Verify default cluster column
if (!(default_cluster_col %in% colnames(seurat_merged@meta.data))) {
  stop("Default cluster column '", default_cluster_col, "' not found in metadata!")
}

cat("\n✓ Default cluster column verified:", default_cluster_col, "\n")
```

# Join Layers (CRITICAL for Seurat v5)

```{r join_layers}
cat("\n=== Normalizing and Joining Layers ===\n\n")

# PROVEN_PATTERN: LOGIC-09-010 (CRITICAL)
# SOURCE: Documented requirement for Seurat v5 marker analysis
# CRITICAL: MUST execute after merge, before wilcoxauc()

# Memory cleanup before major transformation
cat("Running gc() before normalization...\n")
gc()

# Normalize
cat("Normalizing data...\n")
seurat_merged <- NormalizeData(seurat_merged, verbose = FALSE)
cat("  ✓ Normalization complete\n\n")

# Memory cleanup before JoinLayers
cat("Running gc() before JoinLayers...\n")
gc()

# Join layers (CRITICAL for Seurat v5)
cat("Joining layers...\n")
seurat_merged <- JoinLayers(seurat_merged)
cat("  ✓ JoinLayers complete\n\n")

# Memory cleanup after JoinLayers
cat("Running gc() after JoinLayers...\n")
gc()

cat("\n✓ Layers joined successfully\n")
```

# Quick Visualization Test

```{r quick_viz}
cat("\n=== Quick Visualization Test ===\n\n")

# PROVEN_PATTERN: LOGIC-09-011, LOGIC-09-012
# SOURCE: hbca_analysis/.../Epithelial_CompartmentIntegration_Filtered_20250814.Rmd:242-252

# UMAP by patient
p1 <- DimPlot(
  seurat_merged,
  reduction = "umap",
  group.by = "patient_id",
  pt.size = 0.1,
  raster = FALSE
) + ggtitle("UMAP by Patient")

print(p1)

# UMAP by cluster (default resolution)
p2 <- DimPlot(
  seurat_merged,
  reduction = "umap",
  group.by = default_cluster_col,
  label = TRUE,
  pt.size = 0.1,
  raster = FALSE
) +
  ggtitle(paste("UMAP by", default_cluster_col)) +
  NoLegend()

print(p2)

# UMAP by cluster (resolution 1.0)
p3 <- DimPlot(
  seurat_merged,
  reduction = "umap",
  group.by = "leiden_scvi_1.0",
  label = TRUE,
  pt.size = 0.1,
  raster = FALSE
) + ggtitle(paste("UMAP by leiden_scvi_1.0")) +
  NoLegend()
print(p3)

# UMAP by cluster (resolution 5.0)
p4 <- DimPlot(
  seurat_merged,
  reduction = "umap",
  group.by = "leiden_scvi_5.0",
  label = TRUE,
  pt.size = 0.1,
  raster = FALSE
) + ggtitle(paste("UMAP by leiden_scvi_5.0")) +
  NoLegend()
print(p4)

# UMAP by cell type labels (if available)
if ("HBCATransferredLabels.Kumar_2023" %in% colnames(seurat_merged@meta.data)) {
  cat("\nCreating UMAP by cell type labels...\n")
  p5 <- DimPlot(
    seurat_merged,
    reduction = "umap",
    group.by = "HBCATransferredLabels.Kumar_2023",
    label = TRUE,
    repel = TRUE,
    pt.size = 0.1,
    alpha = 0.3,
    raster = FALSE
  ) +
    ggtitle("UMAP by Cell Type (HBCA Kumar 2023)") +
    theme(legend.position = "right")

  print(p5)
} else {
  cat("\n⚠️  HBCATransferredLabels.Kumar_2023 not found - skipping cell type UMAP\n")
}

cat("\n✓ Visualizations created successfully\n")
```

# QC Metric Overlays

```{r qc_overlays}
cat("\n=== QC Metric Overlays ===\n\n")

# Pre-calculate percentile cutoffs with na.rm=TRUE to handle missing values
q95_counts <- quantile(seurat_merged$nCount_RNA, 0.95, na.rm = TRUE)

q5_features <- quantile(seurat_merged$nFeature_RNA, 0.05, na.rm = TRUE)
q95_features <- quantile(seurat_merged$nFeature_RNA, 0.95, na.rm = TRUE)

cat("Percentile cutoffs calculated:\n")
cat("q95 =", round(q95_counts, 2), "\n")
cat("  nFeature_RNA: q5 =", round(q5_features, 2), ", q95 =", round(q95_features, 2), "\n\n")

# FeaturePlot 1: nCount_RNA (log10 scale, q5-q95 cutoffs)
cat("Creating nCount_RNA overlay...\n")

# Order cells by nCount_RNA descending (highest first in vector = lowest on top visually)
cell_order_counts <- colnames(seurat_merged)[order(seurat_merged$nCount_RNA, decreasing = TRUE)]

p_counts <- FeaturePlot(
  seurat_merged,
  features = "nCount_RNA",
  max.cutoff = q95_counts,
  pt.size = 0.2,
  alpha = 0.3,
  raster = FALSE
) +
  scale_color_viridis(option = "viridis", direction = -1) +
  ggtitle("UMAP: UMI Counts per Cell (lowest on top, q5-q95)")

print(p_counts)

# FeaturePlot 2: nFeature_RNA (q5-q95 cutoffs)
cat("Creating nFeature_RNA overlay...\n")
p_features <- FeaturePlot(
  seurat_merged,
  features = "nFeature_RNA",
  max.cutoff = q95_features,
  min.cutoff = q5_features,
  order = TRUE,
  raster = FALSE
) +
  scale_color_viridis(option = "viridis") +
  ggtitle("UMAP: Genes Detected per Cell (q5-q95)")

print(p_features)

# FeaturePlot 3: percent.mt (if available)
if ("percent.mt" %in% colnames(seurat_merged@meta.data)) {
  cat("Creating percent.mt overlay...\n")

  # Calculate percentiles with na.rm
  q5_mito <- quantile(seurat_merged$percent.mt, 0.05, na.rm = TRUE)
  q95_mito <- quantile(seurat_merged$percent.mt, 0.95, na.rm = TRUE)

  cat("  percent.mt: q5 =", round(q5_mito, 2), ", q95 =", round(q95_mito, 2), "\n")

  p_mito <- FeaturePlot(
    seurat_merged,
    features = "percent.mt",
    max.cutoff = q95_mito,
    min.cutoff = q5_mito,
    order = TRUE,
    raster = FALSE
  ) +
    scale_color_viridis(option = "viridis") +
    ggtitle("UMAP: Mitochondrial Percentage (q5-q95)")

  print(p_mito)
}

# FeaturePlot 4: scDblFinder_score (if available)
if ("scDblFinder_score" %in% colnames(seurat_merged@meta.data)) {
  cat("Creating scDblFinder_score overlay...\n")

  p_doublet <- FeaturePlot(
    seurat_merged,
    features = "scDblFinder_score",
    order = TRUE,  # Highest scores on top
    raster = FALSE
  ) +
    scale_color_viridis(option = "magma")
  print(p_doublet)
}

cat("\n✓ QC metric overlays complete\n")
```

# QC Violin Plots

```{r qc_violins}
cat("\n=== QC Violin Plots ===\n\n")

# Calculate percentiles for horizontal lines
pct1_counts <- quantile(seurat_merged$nCount_RNA, 0.01, na.rm = TRUE)
pct99_counts <- quantile(seurat_merged$nCount_RNA, 0.99, na.rm = TRUE)

pct1_features <- quantile(seurat_merged$nFeature_RNA, 0.01, na.rm = TRUE)
pct99_features <- quantile(seurat_merged$nFeature_RNA, 0.99, na.rm = TRUE)

cat("Percentiles calculated:\n")
cat("  nCount_RNA: 1st =", round(pct1_counts, 2), ", 99th =", round(pct99_counts, 2), "\n")
cat("  nFeature_RNA: 1st =", round(pct1_features, 2), ", 99th =", round(pct99_features, 2), "\n\n")

# Violin 1: nCount_RNA (log scale)
cat("Creating nCount_RNA violin...\n")
p_violin_counts <- VlnPlot(
  seurat_merged,
  features = "nCount_RNA",
  group.by = default_cluster_col,
  pt.size = 0
) +
  scale_y_log10() +
  geom_hline(yintercept = pct1_counts, linetype = "dashed", color = "red", alpha = 0.5) +
  geom_hline(yintercept = pct99_counts, linetype = "dashed", color = "red", alpha = 0.5) +
  ggtitle(paste("UMI Counts by", default_cluster_col, "(log10 scale)")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_violin_counts)

# Violin 2: nFeature_RNA
cat("Creating nFeature_RNA violin...\n")
p_violin_features <- VlnPlot(
  seurat_merged,
  features = "nFeature_RNA",
  group.by = default_cluster_col,
  pt.size = 0
) +
  geom_hline(yintercept = pct1_features, linetype = "dashed", color = "red", alpha = 0.5) +
  geom_hline(yintercept = pct99_features, linetype = "dashed", color = "red", alpha = 0.5) +
  ggtitle(paste("Genes Detected by", default_cluster_col)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_violin_features)

cat("\n✓ QC violin plots complete\n")
```

# Anatomical UMAPs

```{r anatomical_umaps}
cat("\n=== Anatomical UMAPs ===\n\n")

# PROVEN_PATTERN: LOGIC-09-014 (adapted for anatomical metadata)

# UMAP 1: Colored by position (all patients)
cat("Creating UMAP by position (all patients)...\n")
p_position <- DimPlot(
  seurat_merged,
  reduction = "umap",
  group.by = "position",
  pt.size = 0.1,
  raster = FALSE
) +
  ggtitle("UMAP by Tissue Position") +
  scale_color_viridis_d(option = "turbo")

print(p_position)

# UMAP 2: By position, split by patient (ncol=2)
cat("Creating UMAP by position split by patient...\n")
p_position_split <- DimPlot(
  seurat_merged,
  reduction = "umap",
  group.by = "position",
  split.by = "patient_id",
  ncol = 2,
  pt.size = 0.1,
  raster = FALSE
) +
  ggtitle("UMAP by Position (per Patient)") +
  scale_color_viridis_d(option = "turbo")

print(p_position_split)

# UMAP 3: By midline position (if available)
if ("midline_position" %in% colnames(seurat_merged@meta.data)) {
  cat("Creating UMAP by midline position...\n")
  p_midline <- DimPlot(
    seurat_merged,
    reduction = "umap",
    group.by = "midline_position",
    pt.size = 0.1,
    raster = FALSE
  ) +
    ggtitle("UMAP by Midline Position") +
    scale_color_viridis_d(option = "turbo")

  print(p_midline)
}

# UMAP 4: By vertical position (if available)
if ("vertical_position" %in% colnames(seurat_merged@meta.data)) {
  cat("Creating UMAP by vertical position...\n")
  p_vertical <- DimPlot(
    seurat_merged,
    reduction = "umap",
    group.by = "vertical_position",
    pt.size = 0.1,
    raster = FALSE
  ) +
    ggtitle("UMAP by Vertical Position (Upper/Lower)") +
    scale_color_viridis_d(option = "turbo")

  print(p_vertical)
}

# UMAP 5: By horizontal position (if available)
if ("horizontal_position" %in% colnames(seurat_merged@meta.data)) {
  cat("Creating UMAP by horizontal position...\n")
  p_horizontal <- DimPlot(
    seurat_merged,
    reduction = "umap",
    group.by = "horizontal_position",
    pt.size = 0.1,
    raster = FALSE
  ) +
    ggtitle("UMAP by Horizontal Position (Inner/Outer)") +
    scale_color_viridis_d(option = "turbo")

  print(p_horizontal)
}

cat("\n✓ Anatomical UMAPs complete\n")
```

# Multi-Resolution Marker Analysis

```{r multi_res_markers, fig.width=14, fig.height=10}
cat("\n=== Multi-Resolution Marker Analysis ===\n\n")

# PROVEN_PATTERN: LOGIC-09-015, LOGIC-09-016, LOGIC-09-017
# SOURCE: hbca_analysis/.../Epithelial_CompartmentIntegration_Filtered_20250814.Rmd:390-450

cat("Running marker analysis across", length(clustering_columns), "resolutions...\n\n")

# Memory cleanup before intensive marker analysis
cat("Running gc() before marker analysis...\n")
gc()

# Initialize storage list
all_resolution_markers <- list()

# Loop over all clustering resolutions
for (i in seq_along(clustering_columns)) {
  res_col <- clustering_columns[i]

  cat("---\n")
  cat("Resolution", i, "of", length(clustering_columns), ":", res_col, "\n")

  # Set identity to current resolution
  Idents(seurat_merged) <- seurat_merged@meta.data[[res_col]]

  n_clusters <- length(unique(Idents(seurat_merged)))
  cat("  Clusters:", n_clusters, "\n")

  # Run wilcoxauc (presto differential expression)
  cat("  Running presto::wilcoxauc()...\n")
  markers <- presto::wilcoxauc(seurat_merged, assay = "data")

  cat("  Found", nrow(markers), "marker tests\n")

  # Filter for significant markers
  sig_markers <- markers %>%
    filter(padj < 0.05, logFC > 0) %>%
    arrange(group, desc(logFC))

  cat("  Significant markers (padj < 0.05, logFC > 0):", nrow(sig_markers), "\n")

  # Get top markers per cluster for visualization
  top_markers <- sig_markers %>%
    group_by(group) %>%
    slice_max(order_by = logFC, n = 5) %>%
    ungroup()

  cat("  Top 5 markers per cluster:", nrow(top_markers), "genes\n")

  # Store full results
  all_resolution_markers[[res_col]] <- markers

  # Memory cleanup after marker calculation
  gc()

  # Generate heatmaps (if we have markers)
  if (nrow(top_markers) > 0) {
    cat("  Generating heatmaps...\n")

    # Approach 1: Manual subsample (fixed total cells)
    tryCatch({
      cat("    - Subsampled heatmap (max 10K cells)...\n")
      cells_for_heatmap <- sample(colnames(seurat_merged),
                                   size = min(10000, ncol(seurat_merged)))

      p_heatmap_subsample <- DoHeatmap(
        seurat_merged[, cells_for_heatmap],
        features = unique(top_markers$feature),
        group.by = res_col,
        size = 3,
        angle = 90,
        slot = "data"
      ) +
        scale_fill_viridis(option = "magma") +
        ggtitle(paste("Top Markers Heatmap (Subsampled) -", res_col)) +
        theme(axis.text.y = element_text(size = 6))

      print(p_heatmap_subsample)
    }, error = function(e) {
      cat("    ⚠️  Subsampled heatmap failed:", e$message, "\n")
    })

    # Approach 2: Downsample by cluster (balanced representation)
    tryCatch({
      cat("    - Downsampled heatmap (1K cells per cluster)...\n")

      p_heatmap_downsample <- DoHeatmap(
        subset(seurat_merged, downsample = 1000),
        features = unique(top_markers$feature),
        group.by = res_col,
        size = 3,
        angle = 90,
        slot = "data"
      ) +
        scale_fill_viridis(option = "magma") +
        ggtitle(paste("Top Markers Heatmap (Downsampled per Cluster) -", res_col)) +
        theme(axis.text.y = element_text(size = 6))

      print(p_heatmap_downsample)
    }, error = function(e) {
      cat("    ⚠️  Downsampled heatmap failed:", e$message, "\n")
    })

    # Generate dotplot
    cat("  Generating dotplot...\n")

    tryCatch({
      p_dotplot <- DotPlot(
        seurat_merged,
        features = unique(top_markers$feature),
        group.by = res_col
      ) +
        coord_flip() +
        scale_color_viridis(option = "viridis") +
        ggtitle(paste("Top Markers -", res_col)) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
              axis.text.y = element_text(size = 6))

      print(p_dotplot)
    }, error = function(e) {
      cat("  ⚠️  Dotplot generation failed:", e$message, "\n")
    })
  } else {
    cat("  No significant markers found - skipping visualizations\n")
  }

  # Memory cleanup at end of each resolution
  cat("  Running gc() after resolution", res_col, "...\n")
  gc()

  cat("\n")
}

cat("✓ Multi-resolution marker analysis complete\n")
cat("  Total resolutions analyzed:", length(all_resolution_markers), "\n")
```

# Save Per-Resolution Markers

```{r save_markers}
cat("\n=== Saving Per-Resolution Markers ===\n\n")

# Check if markers were generated
if (length(all_resolution_markers) == 0) {
  cat("⚠️  No markers to save - skipping\n")
} else {
  # Create output directory
  output_dir <- file.path(PROJECT_ROOT, "outputs/09_IntegrationPreview/markers_by_resolution")

  if (!dir.exists(output_dir)) {
    cat("Creating output directory:", output_dir, "\n")
    dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
  }

  cat("Output directory:", output_dir, "\n\n")

  # Save per-resolution markers
  cat("Saving per-resolution marker files...\n")

  for (res_col in names(all_resolution_markers)) {
    # Extract resolution value from column name (e.g., "leiden_scvi_0.3" -> "0.3")
    res_value <- gsub("leiden_scvi_([0-9.]+)$", "\\1", res_col)

    # Construct filename
    filename <- paste0("markers_res_", res_value, ".csv")
    filepath <- file.path(output_dir, filename)

    # Save CSV
    tryCatch({
      write.csv(all_resolution_markers[[res_col]], file = filepath, row.names = FALSE)
      cat("  ✓", filename, "-", nrow(all_resolution_markers[[res_col]]), "rows\n")
    }, error = function(e) {
      cat("  ⚠️  Failed to save", filename, ":", e$message, "\n")
    })
  }

  # Save combined file with all resolutions
  cat("\nCreating combined marker file...\n")

  combined_markers <- bind_rows(
    lapply(names(all_resolution_markers), function(res_col) {
      res_value <- gsub(".*res_([0-9.]+)$", "\\1", res_col)
      all_resolution_markers[[res_col]] %>%
        mutate(resolution = res_value, .before = 1)
    })
  )

  combined_filepath <- file.path(output_dir, "all_markers_all_resolutions.csv")

  tryCatch({
    write.csv(combined_markers, file = combined_filepath, row.names = FALSE)
    cat("  ✓ all_markers_all_resolutions.csv -", nrow(combined_markers), "rows\n")
  }, error = function(e) {
    cat("  ⚠️  Failed to save combined file:", e$message, "\n")
  })

  cat("\n✓ Marker files saved successfully\n")
  cat("  Output directory:", output_dir, "\n")
  cat("  Files saved:", length(all_resolution_markers) + 1, "\n")
}
```

# Generate Output Schemas

```{r generate_schemas}
cat("\n=== Generating Output Schemas ===\n\n")

# Generate schema for unified_metadata.csv
cat("Generating unified_metadata schema...\n")

metadata_schema <- list(
  file = "unified_metadata.csv",
  generated = format(Sys.time(), "%Y-%m-%dT%H:%M:%S"),
  type = "csv_metadata",
  description = "Unified cell metadata from Step 08 integration + Step 02 QC metrics + Step 04 cell type labels + anatomical positions (QC-passing singlets only)",
  dimensions = list(
    n_rows = nrow(unified_metadata),
    n_cols = ncol(unified_metadata)
  ),
  columns = colnames(unified_metadata),
  dtypes = sapply(unified_metadata, function(x) class(x)[1]),
  sample_rows = lapply(1:min(3, nrow(unified_metadata)), function(i) {
    as.list(unified_metadata[i, ])
  })
)

metadata_output <- file.path(PROJECT_ROOT, "outputs/09_IntegrationPreview/metadata")
schema_path <- file.path(metadata_output, "unified_metadata.schema.yaml")
yaml::write_yaml(metadata_schema, schema_path)
cat("  ✓ unified_metadata.schema.yaml saved\n")

# Generate schema for marker files
if (length(all_resolution_markers) > 0) {
  cat("\nGenerating markers schema...\n")

  first_markers <- all_resolution_markers[[1]]

  markers_schema <- list(
    file = "markers_res_*.csv",
    generated = format(Sys.time(), "%Y-%m-%dT%H:%M:%S"),
    type = "presto_wilcoxauc",
    description = "Differential expression results from presto::wilcoxauc per cluster resolution",
    columns = list(
      feature = "Gene name",
      group = "Cluster ID (numeric)",
      avgExpr = "Average expression in cluster",
      logFC = "Log2 fold-change vs other clusters",
      statistic = "Wilcoxon test statistic",
      auc = "Area under ROC curve (0.5-1.0)",
      pval = "P-value",
      padj = "Adjusted p-value (Bonferroni)",
      pct_in = "Percent cells expressing in cluster",
      pct_out = "Percent cells expressing outside cluster"
    ),
    dtypes = sapply(first_markers, function(x) class(x)[1]),
    sample_rows = lapply(1:min(3, nrow(first_markers)), function(i) {
      as.list(first_markers[i, ])
    }),
    resolutions_available = names(all_resolution_markers)
  )

  markers_output <- file.path(PROJECT_ROOT, "outputs/09_IntegrationPreview/markers_by_resolution")
  markers_schema_path <- file.path(markers_output, "markers.schema.yaml")
  yaml::write_yaml(markers_schema, markers_schema_path)
  cat("  ✓ markers.schema.yaml saved\n")
} else {
  cat("\nNo markers to schema - skipping markers.schema.yaml\n")
}

# Generate schema for Seurat object structure (formulaic extraction)
cat("\nGenerating Seurat object schema...\n")

seurat_schema <- list(
  object = "seurat_merged",
  generated = format(Sys.time(), "%Y-%m-%dT%H:%M:%S"),
  type = "Seurat",
  package = "SeuratObject",
  version = as.character(seurat_merged@version),

  # Dimensions
  dimensions = list(
    n_cells = ncol(seurat_merged),
    n_genes = nrow(seurat_merged)
  ),

  # Assays
  assays = list(
    active = seurat_merged@active.assay,
    available = names(seurat_merged@assays),
    details = lapply(names(seurat_merged@assays), function(assay_name) {
      assay_obj <- seurat_merged@assays[[assay_name]]
      list(
        class = class(assay_obj)[1],
        layers = if ("Assay5" %in% class(assay_obj)) names(assay_obj@layers) else "legacy",
        n_features = nrow(assay_obj),
        n_cells = ncol(assay_obj)
      )
    }) %>% setNames(names(seurat_merged@assays))
  ),

  # Metadata
  metadata = list(
    n_columns = ncol(seurat_merged@meta.data),
    columns = colnames(seurat_merged@meta.data),
    dtypes = sapply(seurat_merged@meta.data, function(x) class(x)[1]),
    sample_rows = lapply(1:min(3, ncol(seurat_merged)), function(i) {
      as.list(seurat_merged@meta.data[i, ])
    })
  ),

  # Reductions
  reductions = list(
    available = names(seurat_merged@reductions),
    details = lapply(names(seurat_merged@reductions), function(red_name) {
      red_obj <- seurat_merged@reductions[[red_name]]
      list(
        class = class(red_obj)[1],
        dimensions = ncol(red_obj),
        n_cells = nrow(red_obj),
        key = red_obj@key,
        global = red_obj@global,
        assay_used = red_obj@assay.used
      )
    }) %>% setNames(names(seurat_merged@reductions))
  ),

  # Clustering
  clusters = list(
    active_ident = list(
      levels = levels(seurat_merged@active.ident),
      n_levels = length(levels(seurat_merged@active.ident)),
      class = class(seurat_merged@active.ident)[1]
    ),
    cluster_columns = clustering_columns,
    n_resolutions = length(clustering_columns)
  ),

  # Graphs and neighbors
  graphs = list(
    available = names(seurat_merged@graphs),
    n_graphs = length(seurat_merged@graphs)
  ),

  neighbors = list(
    available = names(seurat_merged@neighbors),
    n_neighbors = length(seurat_merged@neighbors)
  ),

  # Project metadata
  project = list(
    name = seurat_merged@project.name,
    misc = names(seurat_merged@misc),
    commands = names(seurat_merged@commands),
    tools = names(seurat_merged@tools)
  )
)

seurat_schema_path <- file.path(metadata_output, "seurat_object.schema.yaml")
yaml::write_yaml(seurat_schema, seurat_schema_path)
cat("  ✓ seurat_object.schema.yaml saved\n")

cat("\n✓ All schemas generated\n")
cat("  unified_metadata.schema.yaml\n")
if (length(all_resolution_markers) > 0) {
  cat("  markers.schema.yaml\n")
}
cat("  seurat_object.schema.yaml\n")
```

# Save Seurat Object

```{r save_seurat}
cat("\n=== Saving Seurat Object ===\n\n")

# Create seurat_objects output directory
seurat_output <- file.path(PROJECT_ROOT, "outputs/09_IntegrationPreview/seurat_objects")
dir.create(seurat_output, recursive = TRUE, showWarnings = FALSE)

# Save complete Seurat object
seurat_path <- file.path(seurat_output, "integrated_seurat.rds")
saveRDS(seurat_merged, file = seurat_path)

cat("✓ Seurat object saved\n")
cat("  Path:", seurat_path, "\n")
cat("  Size:", format(file.size(seurat_path) / 1024^3, digits = 2), "GB\n")
cat("  Cells:", ncol(seurat_merged), "\n")
cat("  Genes:", nrow(seurat_merged), "\n")
```

# Summary

```{r summary}
cat("\n=== INTERACTIVE SESSION SUMMARY ===\n\n")

cat("Seurat Object:\n")
cat("  Cells:", ncol(seurat_merged), "\n")
cat("  Genes:", nrow(seurat_merged), "\n")
cat("  Patients:", length(unique(seurat_merged$patient_id)), "\n")
cat("  Positions:", length(unique(seurat_merged$position)), "\n")
cat("  Samples:", length(unique(seurat_merged$orig.ident)), "\n")

cat("\nDimensional Reductions:\n")
print(names(seurat_merged@reductions))

cat("\nCluster Columns:\n")
cat(" ", paste(clustering_columns, collapse = ", "), "\n")

cat("\nRNA Assay Layers:\n")
cat(" ", paste(names(seurat_merged[["RNA"]]@layers), collapse = ", "), "\n")

cat("\nMetadata Columns (first 20):\n")
cat(" ", paste(head(colnames(seurat_merged@meta.data), 20), collapse = ", "), "\n")

# Check anatomical metadata availability
cat("\nAnatomical Metadata:\n")
if ("midline_position" %in% colnames(seurat_merged@meta.data)) {
  cat("  Midline positions:", paste(unique(seurat_merged$midline_position), collapse = ", "), "\n")
}
if ("vertical_position" %in% colnames(seurat_merged@meta.data)) {
  cat("  Vertical positions:", paste(unique(seurat_merged$vertical_position), collapse = ", "), "\n")
}
if ("horizontal_position" %in% colnames(seurat_merged@meta.data)) {
  cat("  Horizontal positions:", paste(unique(seurat_merged$horizontal_position), collapse = ", "), "\n")
}

cat("\n✓ All chunks executed successfully!\n")
cat("\nEnhancements in v3:\n")
cat("  - Focused metadata assembly (Step 02 QC + Step 04 labels + anatomical)\n")
cat("  - Added HBCA cell type annotations (HBCATransferredLabels.Kumar_2023)\n")
cat("  - Removed doublet/MAD columns (all cells are QC-passing singlets)\n")
cat("  - Anatomical position mapping (midline, vertical, horizontal)\n")
cat("  - QC overlays for UMI, genes, and mitochondrial percentage\n")
cat("  - Multi-resolution clustering and marker analysis\n")
```
