---
title: "SingleR L1 Classification — FLEX `r params$compartment` vs iHBCA L1 Reference"
subtitle: "11-type iHBCA L1 pseudobulk reference, restricted to FLEX panel"
params:
  compartment: "Epithelial"
  seurat_rds: ""
  reference_csv: ""
  outdir: ""
  dry_run: false
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 7)
suppressPackageStartupMessages({
  library(Seurat)
  library(SingleR)
  library(SingleCellExperiment)
  library(SummarizedExperiment)
  library(ggplot2)
  library(dplyr)
})

COMPARTMENT   <- params$compartment
SEURAT_RDS    <- params$seurat_rds
REFERENCE_CSV <- params$reference_csv
OUTDIR        <- params$outdir
DRY_RUN       <- isTRUE(params$dry_run)
dir.create(OUTDIR, recursive = TRUE, showWarnings = FALSE)
```

# Parameters

```{r params_echo}
cat("Compartment:   ", COMPARTMENT, "\n")
cat("Seurat RDS:    ", SEURAT_RDS, "\n")
cat("Reference CSV: ", REFERENCE_CSV, "\n")
cat("Output dir:    ", OUTDIR, "\n")
cat("Dry run:       ", DRY_RUN, "\n")
```

# Validation Preamble

```{r validate}
required <- list(
  seurat_rds    = SEURAT_RDS,
  reference_csv = REFERENCE_CSV
)
missing_items <- character(0)
for (nm in names(required)) {
  f <- required[[nm]]
  ok <- file.exists(f)
  cat(sprintf("  [%s] %-15s %s\n", if (ok) "OK " else "MISS", nm, f))
  if (!ok) missing_items <- c(missing_items, nm)
}
if (length(missing_items) > 0) {
  stop("Missing required inputs: ", paste(missing_items, collapse = ", "))
}
if (DRY_RUN) {
  cat("\nVALIDATION PASSED (dry run) — exiting before heavy computation\n")
  knitr::knit_exit()
}
```

# Load L1 Pseudobulk Reference

Pattern source: `singler_l1.Rmd:36-44`. Reference is 11 L1 types x ~18K FLEX
panel genes (rebuilt by `09_build_l1_reference.py`).

```{r load-reference}
cat("Loading pseudobulk reference:", REFERENCE_CSV, "\n")
pb_df <- read.csv(REFERENCE_CSV, row.names = 1, check.names = FALSE)
cat("Reference:", nrow(pb_df), "types x", ncol(pb_df), "genes\n")
cat("\nL1 types:\n")
cat(paste(" ", rownames(pb_df)), sep = "\n")

pb_mat <- as.matrix(pb_df)
```

# Load FLEX Seurat

Pattern source: `singler_l1.Rmd:48-59`.

```{r load-seurat}
cat("Loading Seurat:", SEURAT_RDS, "\n")
srt <- readRDS(SEURAT_RDS)
srt <- UpdateSeuratObject(srt)
cat("Cells:", ncol(srt), "\nGenes:", nrow(srt), "\n")

has_data <- !is.null(tryCatch(GetAssayData(srt, layer = "data"), error = function(e) NULL))
if (!has_data || length(GetAssayData(srt, layer = "data")@x) == 0) {
  cat("Normalizing...\n")
  srt <- NormalizeData(srt, verbose = FALSE)
}
```

# Gene Matching

Pattern source: `singler_l1.Rmd:63-74`. With the rebuilt reference, we expect
~17,500+ shared genes (vs 5,970 under the old reference).

```{r gene-matching}
ref_genes <- colnames(pb_mat)
query_genes <- rownames(srt)
shared_genes <- intersect(ref_genes, query_genes)

cat("Reference genes:", length(ref_genes), "\n")
cat("Query genes:   ", length(query_genes), "\n")
cat("Shared genes:  ", length(shared_genes), "\n")
cat("Coverage:      ", sprintf("%.1f%%", 100 * length(shared_genes) / length(ref_genes)), "\n")

stopifnot(length(shared_genes) >= 100)
if (length(shared_genes) < 15000) {
  cat("WARNING: shared gene count is lower than expected (~17,500) for rebuilt reference\n")
}
```

# Run SingleR

Pattern source: `singler_l1.Rmd:78-105`. Uses `de.method = "classic"` to match
the validated L1 pipeline; the reference is pseudobulk-mean so we log-transform
once here.

```{r run-singler}
ref_mat <- t(pb_mat[, shared_genes])
ref_log <- log1p(ref_mat)

ref_sce <- SummarizedExperiment(
  assays = list(logcounts = ref_log),
  colData = DataFrame(label = rownames(pb_mat))
)

query_sce <- SingleCellExperiment(
  assays = list(logcounts = GetAssayData(srt, layer = "data")[shared_genes, ])
)

cat("Running SingleR (de.method='classic', 11 L1 types)...\n")
t0 <- proc.time()

results <- SingleR(
  test = query_sce,
  ref = ref_sce,
  labels = rownames(pb_mat),
  de.method = "classic"
)

elapsed <- (proc.time() - t0)["elapsed"]
cat(sprintf("Completed in %.1f seconds\n", elapsed))
```

# Save Results

Pattern source: `singler_l1.Rmd:109-131`, with added `raw_label` column (the
non-pruned assignment — useful for the confusogram downstream).

```{r save}
max_scores <- apply(results$scores, 1, max)
deltas <- results$delta.next

# raw = argmax assignment (labels column). pruned = pruned.labels (NA if
# below-threshold). Both are reported; downstream scripts consume pruned_label
# as the authoritative SingleR L1.
raw_labels <- results$labels

n_pruned <- sum(is.na(results$pruned.labels))
cat(sprintf("Total: %d, Pruned: %d (%.1f%%)\n",
            ncol(srt), n_pruned, 100 * n_pruned / ncol(srt)))

out_df <- data.frame(
  cell_id       = colnames(srt),
  pruned_label  = results$pruned.labels,
  raw_label     = raw_labels,
  score         = max_scores,
  delta_next    = deltas,
  stringsAsFactors = FALSE
)

out_csv <- file.path(OUTDIR, sprintf("singler_l1_predictions_%s.csv", COMPARTMENT))
write.csv(out_df, out_csv, row.names = FALSE)
cat("Saved:", out_csv, "\n")

cat("\nL1 pruned label distribution:\n")
print(sort(table(out_df$pruned_label, useNA = "ifany"), decreasing = TRUE))

cat("\nL1 raw label distribution:\n")
print(sort(table(out_df$raw_label), decreasing = TRUE))
```

# Score Summary

```{r score-summary, fig.width=10, fig.height=6}
score_by_type <- out_df %>%
  filter(!is.na(pruned_label)) %>%
  group_by(pruned_label) %>%
  summarise(
    n            = n(),
    median_score = median(score),
    median_delta = median(delta_next),
    .groups      = "drop"
  ) %>%
  arrange(desc(n))

print(score_by_type)
```

```{r session-info}
cat("\nSingleR:", as.character(packageVersion("SingleR")), "\n")
cat("Seurat: ", as.character(packageVersion("Seurat")), "\n")
```
