---
title: "L1 Harmonization — `r params$compartment`"
subtitle: "SingleR L1 + cluster_annotation overrides + Leiden vote consolidation"
params:
  compartment: "Epithelial"
  singler_csv: ""
  seurat_rds: ""
  outdir: ""
  dry_run: false
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 12, fig.height = 8)
suppressPackageStartupMessages({
  library(Seurat)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
})

COMPARTMENT <- params$compartment
SINGLER_CSV <- params$singler_csv
SEURAT_RDS  <- params$seurat_rds
OUTDIR      <- params$outdir
DRY_RUN     <- isTRUE(params$dry_run)
dir.create(OUTDIR, recursive = TRUE, showWarnings = FALSE)
```

# Parameters

```{r params_echo}
cat("Compartment:   ", COMPARTMENT, "\n")
cat("SingleR CSV:   ", SINGLER_CSV, "\n")
cat("Seurat RDS:    ", SEURAT_RDS, "\n")
cat("Output dir:    ", OUTDIR, "\n")
cat("Dry run:       ", DRY_RUN, "\n")
```

# Validation

```{r validate}
required <- list(singler_csv = SINGLER_CSV, seurat_rds = SEURAT_RDS)
missing_items <- character(0)
for (nm in names(required)) {
  f <- required[[nm]]
  ok <- file.exists(f)
  cat(sprintf("  [%s] %-15s %s\n", if (ok) "OK " else "MISS", nm, f))
  if (!ok) missing_items <- c(missing_items, nm)
}
if (length(missing_items) > 0) {
  stop("Missing required inputs: ", paste(missing_items, collapse = ", "))
}
if (DRY_RUN) {
  cat("\nVALIDATION PASSED (dry run) — exiting before heavy computation\n")
  knitr::knit_exit()
}
```

# Configuration

Override types are populations present in FLEX v2 `cluster_annotation` that
are NOT in the 11-type iHBCA L1 vocabulary. These cells get their
`cluster_annotation` identity instead of the (broader) SingleR label.

The vote column is fixed at `leiden_n30_r1.0` — the rebuilt patched cluster
column selected in the preceding integration phase. No resolution scan (the
v1 pipeline scanned multiple resolutions; here we take the patched column as
authoritative).

```{r config}
# Override types matching the Spatial_HBCA authoritative list (l1_harmonize.Rmd):
# populations in cluster_annotation NOT in the 11-type iHBCA L1 vocabulary.
# Keratinocyte is handled by SingleR (present in the iHBCA reference) and is
# not overridden here.
OVERRIDE_TYPES <- c("Adipocyte", "pDC", "PS_Epithelial", "RBC", "Mast", "Neutrophil")
VOTE_COL       <- "leiden_n30_r1.0"

cat("Override types:", paste(OVERRIDE_TYPES, collapse = ", "), "\n")
cat("Vote column:   ", VOTE_COL, "\n")
```

# Load Data

## Seurat (cluster + cluster_annotation)

Pattern source: `l1_harmonize.Rmd:37-58`, simplified — no UMAP join and no
Leiden resolution scan.

```{r load-seurat}
cat("Loading Seurat:", SEURAT_RDS, "\n")
srt <- readRDS(SEURAT_RDS)
srt <- UpdateSeuratObject(srt)
cat("Cells:", ncol(srt), "\n")

if (!VOTE_COL %in% colnames(srt@meta.data)) {
  stop("Expected vote column '", VOTE_COL, "' not in Seurat metadata")
}

meta <- srt@meta.data
meta$cell_id <- rownames(meta)
rm(srt); gc()

cat(VOTE_COL, "clusters:", length(unique(na.omit(meta[[VOTE_COL]]))), "\n")
```

## SingleR L1 Predictions

Pattern source: `l1_harmonize.Rmd:62-70`. Column names updated to match the
output schema of `09b_singler_l1.Rmd` (`pruned_label`, `score`, `delta_next`).

```{r load-singler}
cat("Loading SingleR L1 predictions:", SINGLER_CSV, "\n")
singler <- read.csv(SINGLER_CSV, stringsAsFactors = FALSE)
cat("Rows:", nrow(singler), "\n")
cat("NA pruned_label count:", sum(is.na(singler$pruned_label)), "\n\n")
print(sort(table(singler$pruned_label, useNA = "ifany"), decreasing = TRUE))
```

## Check cluster_annotation

```{r check-cluster-annotation}
stopifnot("cluster_annotation" %in% colnames(meta))
cat("cluster_annotation distribution:\n")
print(sort(table(meta$cluster_annotation, useNA = "ifany"), decreasing = TRUE))
```

## Join

```{r join}
df <- merge(
  meta,
  singler[, c("cell_id", "pruned_label", "score", "delta_next")],
  by = "cell_id", all.x = TRUE
)
cat("Joined:", nrow(df), "cells\n")
cat("With SingleR pruned_label:", sum(!is.na(df$pruned_label)), "\n")
cat("With cluster_annotation:  ", sum(!is.na(df$cluster_annotation)), "\n")
```

# Apply Overrides

Pattern source: `l1_harmonize.Rmd:99-125`.

```{r overrides}
# Start with the SingleR L1 prediction
df$l1_label <- df$pruned_label

n_overrides <- 0
for (otype in OVERRIDE_TYPES) {
  mask <- !is.na(df$cluster_annotation) & df$cluster_annotation == otype
  n <- sum(mask)
  if (n > 0) {
    df$l1_label[mask] <- otype
    n_overrides <- n_overrides + n
    # What SingleR originally called these cells
    singler_top <- sort(table(df$pruned_label[mask], useNA = "ifany"),
                        decreasing = TRUE)
    singler_top_str <- paste(names(head(singler_top, 3)), collapse = ", ")
    cat(sprintf("  Override %s: %d cells (was: %s)\n",
                otype, n, singler_top_str))
  } else {
    cat(sprintf("  Override %s: 0 cells (not found in cluster_annotation)\n", otype))
  }
}

cat(sprintf("\nTotal overrides: %d / %d cells (%.1f%%)\n",
            n_overrides, nrow(df), 100 * n_overrides / nrow(df)))
cat("\nPost-override distribution:\n")
print(sort(table(df$l1_label, useNA = "ifany"), decreasing = TRUE))
```

# Leiden Vote Consolidation

Pattern source: `l1_harmonize.Rmd:131-204`, simplified to use a fixed vote
column (`leiden_n30_r1.0`). Purity is still reported.

```{r vote-purity}
compute_vote_purity <- function(df, leiden_col) {
  cluster_stats <- df %>%
    filter(!is.na(.data[[leiden_col]])) %>%
    group_by(cluster = .data[[leiden_col]]) %>%
    summarise(
      n_cells      = n(),
      label_counts = list(table(l1_label, useNA = "ifany")),
      .groups      = "drop"
    ) %>%
    rowwise() %>%
    mutate(
      majority_label = names(which.max(label_counts)),
      majority_count = max(label_counts),
      vote_purity    = majority_count / n_cells
    ) %>%
    ungroup()

  cluster_to_label  <- setNames(cluster_stats$majority_label, cluster_stats$cluster)
  cluster_to_purity <- setNames(cluster_stats$vote_purity,    cluster_stats$cluster)

  list(
    stats             = cluster_stats,
    mean_purity       = mean(cluster_stats$vote_purity, na.rm = TRUE),
    median_purity     = median(cluster_stats$vote_purity, na.rm = TRUE),
    cluster_to_label  = cluster_to_label,
    cluster_to_purity = cluster_to_purity
  )
}

vp <- compute_vote_purity(df, VOTE_COL)
cat(sprintf("%s: %d clusters, mean purity=%.3f, median=%.3f\n",
            VOTE_COL, nrow(vp$stats), vp$mean_purity, vp$median_purity))
print(vp$stats %>% select(cluster, n_cells, majority_label, majority_count, vote_purity))
```

# Apply Voted Labels

Pattern source: `l1_harmonize.Rmd:189-207`.

```{r apply-voted}
df$voted_l1_label <- vp$cluster_to_label[as.character(df[[VOTE_COL]])]
df$vote_purity    <- vp$cluster_to_purity[as.character(df[[VOTE_COL]])]

df$harmonized_l1 <- df$voted_l1_label
na_voted <- is.na(df$harmonized_l1)
if (any(na_voted)) {
  cat(sprintf("Cells without voted label: %d (fall back to direct override label)\n",
              sum(na_voted)))
  df$harmonized_l1[na_voted] <- df$l1_label[na_voted]
}

cat("\nHarmonized L1 distribution:\n")
print(sort(table(df$harmonized_l1, useNA = "ifany"), decreasing = TRUE))
```

# Save

Pattern source: `l1_harmonize.Rmd:212-300`. Output schema matches the L1
worker brief: `cell_id, ihbca_L1.0_singler, singler_score, singler_delta,
L1.0, L1.0_Long, L1.0_source, vote_purity, leiden_cluster`.

```{r save}
# Long names for override short labels (Adipocyte and Neutrophil are readable as-is)
LONG_NAMES <- c(
  "pDC"          = "Plasmacytoid dendritic cell",
  "PS_Epithelial" = "Patient-specific epithelial",
  "RBC"          = "Red blood cell",
  "Mast"         = "Mast cell"
)

df$harmonized_l1_long <- df$harmonized_l1
for (short in names(LONG_NAMES)) {
  df$harmonized_l1_long[df$harmonized_l1 == short] <- LONG_NAMES[short]
}

# Long -> Short name mapping for L1.0
L1_SHORT <- c(
  "Fibroblast"                            = "Fibro",
  "Basal-myoepithelial"                   = "Basal",
  "Vascular endothelial"                  = "VascEndo",
  "Luminal adaptive secretory precursor"  = "LumSec",
  "Luminal hormone sensing"               = "LumHR",
  "Perivascular"                          = "Peri",
  "T-lymphocyte"                          = "T_cells",
  "B-lymphocyte"                          = "B_cells",
  "Myeloid"                               = "Myeloid",
  "Lymphatic endothelial"                 = "LymphEndo",
  "Lactocyte"                             = "Lacto",
  "Adipocyte"                             = "Adipo",
  "Mast cell"                             = "Mast",
  "Keratinocyte"                          = "Kerat",
  "Plasmacytoid dendritic cell"           = "pDC",
  "Patient-specific epithelial"           = "PS_Epi",
  "Red blood cell"                        = "RBC",
  "Neutrophil"                            = "Neutro"
)

l1_short <- L1_SHORT[df$harmonized_l1_long]
l1_short[is.na(l1_short)] <- df$harmonized_l1_long[is.na(l1_short)]

# Source: cluster_override if the harmonized label came from an override type
l1_source <- ifelse(
  df$harmonized_l1 %in% OVERRIDE_TYPES,
  "cluster_override",
  "singler_vote"
)

out_df <- data.frame(
  cell_id              = df$cell_id,
  ihbca_L1.0_singler   = df$pruned_label,
  singler_score        = round(df$score, 6),
  singler_delta        = round(df$delta_next, 6),
  L1.0                 = l1_short,
  L1.0_Long            = df$harmonized_l1_long,
  L1.0_source          = l1_source,
  vote_purity          = round(df$vote_purity, 6),
  leiden_cluster       = df[[VOTE_COL]],
  stringsAsFactors     = FALSE
)

out_csv <- file.path(OUTDIR,
                     sprintf("harmonized_l1_labels_%s.csv", COMPARTMENT))
write.csv(out_df, out_csv, row.names = FALSE)
cat("Saved:", out_csv, "\n")
cat("Types:", length(unique(out_df$L1.0)), "\n")
cat("Cells:", nrow(out_df), "\n")

cat("\nL1.0 distribution:\n")
print(sort(table(out_df$L1.0), decreasing = TRUE))

cat("\nL1.0_source breakdown:\n")
print(table(out_df$L1.0_source))

# Override detail
cat("\n=== Override Summary ===\n")
for (otype in OVERRIDE_TYPES) {
  long_name  <- LONG_NAMES[otype]
  if (is.na(long_name)) long_name <- otype
  short_name <- L1_SHORT[long_name]
  if (is.na(short_name)) short_name <- otype
  n <- sum(out_df$L1.0_Long == long_name, na.rm = TRUE)
  singler_dist <- sort(
    table(out_df$ihbca_L1.0_singler[out_df$L1.0_Long == long_name]),
    decreasing = TRUE
  )
  cat(sprintf("  %s (%s): %d cells — SingleR called: %s\n",
              long_name, short_name, n,
              paste(sprintf("%s(%d)", names(singler_dist), singler_dist),
                    collapse = ", ")))
}
```

```{r session-info}
cat("\nSeurat:", as.character(packageVersion("Seurat")), "\n")
```
