---
title: "L1 Confusogram — `r params$compartment`"
subtitle: "cluster_annotation vs SingleR L1 raw vs harmonized L1.0"
params:
  compartment: "Epithelial"
  singler_csv: ""       # per-compartment singler_l1_predictions_{compartment}.csv
  harmonized_csv: ""    # per-compartment OR concatenated harmonized_l1_labels.csv
  seurat_rds: ""        # per-compartment Seurat; empty for 'all' mode
  cluster_annotation_csv: ""  # used when seurat_rds is empty (for 'all' mode)
  outdir: ""
  dry_run: false
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 12, fig.height = 10)
suppressPackageStartupMessages({
  library(Seurat)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(pheatmap)
})

COMPARTMENT       <- params$compartment
SINGLER_CSV       <- params$singler_csv
HARMONIZED_CSV    <- params$harmonized_csv
SEURAT_RDS        <- params$seurat_rds
CLUSTER_ANNO_CSV  <- params$cluster_annotation_csv
OUTDIR            <- params$outdir
DRY_RUN           <- isTRUE(params$dry_run)
dir.create(OUTDIR, recursive = TRUE, showWarnings = FALSE)
```

# Parameters

```{r params_echo}
cat("Compartment:          ", COMPARTMENT, "\n")
cat("SingleR CSV:          ", SINGLER_CSV, "\n")
cat("Harmonized CSV:       ", HARMONIZED_CSV, "\n")
cat("Seurat RDS:           ", SEURAT_RDS, "\n")
cat("cluster_annotation CSV:", CLUSTER_ANNO_CSV, "\n")
cat("Output dir:           ", OUTDIR, "\n")
cat("Dry run:              ", DRY_RUN, "\n")
```

# Validation

```{r validate}
# Either seurat_rds OR cluster_annotation_csv must be provided (not both empty)
has_seurat <- nzchar(SEURAT_RDS) && file.exists(SEURAT_RDS)
has_csv    <- nzchar(CLUSTER_ANNO_CSV) && file.exists(CLUSTER_ANNO_CSV)

required_core <- list(
  singler_csv    = SINGLER_CSV,
  harmonized_csv = HARMONIZED_CSV
)

missing_items <- character(0)
for (nm in names(required_core)) {
  f <- required_core[[nm]]
  ok <- file.exists(f)
  cat(sprintf("  [%s] %-20s %s\n", if (ok) "OK " else "MISS", nm, f))
  if (!ok) missing_items <- c(missing_items, nm)
}

cat(sprintf("  [%s] %-20s %s\n",
            if (has_seurat) "OK " else "--",
            "seurat_rds", SEURAT_RDS))
cat(sprintf("  [%s] %-20s %s\n",
            if (has_csv) "OK " else "--",
            "cluster_anno_csv", CLUSTER_ANNO_CSV))

if (!has_seurat && !has_csv) {
  missing_items <- c(missing_items, "seurat_rds_or_cluster_annotation_csv")
}

if (length(missing_items) > 0) {
  stop("Missing required inputs: ", paste(missing_items, collapse = ", "))
}

if (DRY_RUN) {
  cat("\nVALIDATION PASSED (dry run) — exiting before heavy loads\n")
  knitr::knit_exit()
}
```

# Load All Three Label Sets

Pattern source: `annotation_confusogram.Rmd:37-109`. The L1 harmonized CSV
uses the new schema (`pruned_label`, `L1.0`, `L1.0_Long`), so the join
columns are updated accordingly.

```{r load-data}
# 1. cluster_annotation source
if (has_seurat) {
  cat("Loading Seurat for cluster_annotation...\n")
  srt <- readRDS(SEURAT_RDS)
  srt <- UpdateSeuratObject(srt)
  meta <- data.frame(
    cell_id            = colnames(srt),
    cluster_annotation = srt$cluster_annotation,
    stringsAsFactors   = FALSE
  )
  rm(srt); gc()
} else {
  cat("Loading cluster_annotation CSV:", CLUSTER_ANNO_CSV, "\n")
  ca <- read.csv(CLUSTER_ANNO_CSV, stringsAsFactors = FALSE)
  id_col <- intersect(c("cell_id", "X", "barcode"), colnames(ca))[1]
  if (is.na(id_col)) id_col <- colnames(ca)[1]
  meta <- data.frame(
    cell_id            = ca[[id_col]],
    cluster_annotation = ca$cluster_annotation,
    stringsAsFactors   = FALSE
  )
}
cat("  cluster_annotation:", length(unique(meta$cluster_annotation)), "types\n")

# 2. SingleR L1 raw predictions
singler <- read.csv(SINGLER_CSV, stringsAsFactors = FALSE)
cat("  SingleR L1 pruned:",
    length(unique(singler$pruned_label[!is.na(singler$pruned_label)])), "types\n")

# 3. Harmonized L1
harmonized <- read.csv(HARMONIZED_CSV, stringsAsFactors = FALSE)
cat("  Harmonized L1.0:", length(unique(harmonized$L1.0)), "types\n")

# Join all three on cell_id
df <- meta %>%
  left_join(singler[, c("cell_id", "pruned_label", "raw_label")], by = "cell_id") %>%
  left_join(harmonized[, c("cell_id", "L1.0", "L1.0_Long")],      by = "cell_id")

# Restrict the 'all' mode concat to the cells covered by this SingleR run
if (!has_seurat) {
  df <- df %>% filter(!is.na(pruned_label))
}

cat("\nJoined:", nrow(df), "cells\n")
cat("All three labels present:", sum(complete.cases(df[, c("cluster_annotation", "pruned_label", "L1.0")])), "cells\n")

# Short display names (pattern source: annotation_confusogram.Rmd:67-109)
SHORT_NAMES <- c(
  # iHBCA L1 (11 types)
  "Fibroblast"                            = "Fibro",
  "Basal-myoepithelial"                   = "Basal",
  "Vascular endothelial"                  = "VascEndo",
  "Luminal adaptive secretory precursor"  = "LumSec",
  "Luminal hormone sensing"               = "LumHR",
  "Perivascular"                          = "Peri",
  "T-lymphocyte"                          = "T-lymph",
  "B-lymphocyte"                          = "B-lymph",
  "Myeloid"                               = "Myeloid",
  "Lymphatic endothelial"                 = "LymphEndo",
  "Lactocyte"                             = "Lacto",
  # Override types (v2: Adipocyte, Mast, Keratinocyte)
  "Adipocyte"                             = "Adipo",
  "Mast"                                  = "Mast",
  "Mast cell"                             = "Mast",
  "Keratinocyte"                          = "Kerat"
)

shorten <- function(labels) {
  out <- SHORT_NAMES[labels]
  out[is.na(out)] <- labels[is.na(out)]
  out
}
```

# Helpers

Pattern source: `annotation_confusogram.Rmd:118-139`.

```{r diagonal-order-helper}
diagonal_order <- function(mat_norm) {
  dominant_col <- apply(mat_norm, 1, function(x) colnames(mat_norm)[which.max(x)])
  dominant_val <- apply(mat_norm, 1, max)

  col_order <- names(sort(colSums(mat_norm > 0.5), decreasing = TRUE))
  col_order <- unique(c(col_order, colnames(mat_norm)))

  row_df <- data.frame(
    row     = rownames(mat_norm),
    dom_col = dominant_col,
    dom_val = dominant_val,
    stringsAsFactors = FALSE
  )
  row_df$col_rank <- match(row_df$dom_col, col_order)
  row_df <- row_df[order(row_df$col_rank, -row_df$dom_val), ]

  list(row_order = row_df$row, col_order = col_order)
}

plot_confusion <- function(mat_norm, title, fill_color, out_csv) {
  ord <- diagonal_order(mat_norm)
  mat <- mat_norm[ord$row_order, ord$col_order]
  rownames(mat) <- shorten(rownames(mat))
  colnames(mat) <- shorten(colnames(mat))

  mat_display <- matrix(
    ifelse(mat > 0.01, sprintf("%.0f", mat * 100), ""),
    nrow = nrow(mat), dimnames = dimnames(mat)
  )

  pheatmap(
    mat,
    cluster_rows    = FALSE, cluster_cols = FALSE,
    display_numbers = mat_display,
    number_color    = "black",
    fontsize_number = 7,
    fontsize_row    = 9, fontsize_col = 9,
    color           = colorRampPalette(c("white", fill_color))(100),
    breaks          = seq(0, 1, length.out = 101),
    main            = title,
    legend          = FALSE
  )

  write.csv(as.data.frame.matrix(mat_norm), out_csv)
}
```

# Panel A — cluster_annotation -> SingleR L1 raw

Pattern source: `annotation_confusogram.Rmd:142-174`.

```{r panel-a, fig.width=12, fig.height=10}
df_complete <- df %>% filter(!is.na(pruned_label))
xtab_a <- table(df_complete$cluster_annotation, df_complete$pruned_label)
xtab_a_norm <- sweep(xtab_a, 1, rowSums(xtab_a), "/")
xtab_a_norm[is.nan(xtab_a_norm)] <- 0

plot_confusion(
  xtab_a_norm,
  title     = "cluster_annotation -> SingleR L1 (raw)",
  fill_color = "#08519C",
  out_csv    = file.path(OUTDIR, sprintf("confusogram_cluster_vs_singler_%s.csv",
                                         COMPARTMENT))
)
```

# Panel B — SingleR L1 -> Harmonized L1.0

Pattern source: `annotation_confusogram.Rmd:178-204`.

```{r panel-b, fig.width=10, fig.height=10}
xtab_b <- table(df_complete$pruned_label, df_complete$L1.0)
xtab_b_norm <- sweep(xtab_b, 1, rowSums(xtab_b), "/")
xtab_b_norm[is.nan(xtab_b_norm)] <- 0

plot_confusion(
  xtab_b_norm,
  title      = "SingleR L1 -> Harmonized L1.0 (voted + overrides)",
  fill_color = "#B2182B",
  out_csv    = file.path(OUTDIR, sprintf("confusogram_singler_vs_harmonized_%s.csv",
                                         COMPARTMENT))
)
```

# Panel C — cluster_annotation -> Harmonized L1.0

Pattern source: `annotation_confusogram.Rmd:209-235`.

```{r panel-c, fig.width=12, fig.height=10}
xtab_c <- table(df$cluster_annotation, df$L1.0)
xtab_c_norm <- sweep(xtab_c, 1, rowSums(xtab_c), "/")
xtab_c_norm[is.nan(xtab_c_norm)] <- 0

plot_confusion(
  xtab_c_norm,
  title      = "cluster_annotation -> Harmonized L1.0 (end-to-end)",
  fill_color = "#2CA25F",
  out_csv    = file.path(OUTDIR, sprintf("confusogram_cluster_vs_harmonized_%s.csv",
                                         COMPARTMENT))
)
```

# Summary

```{r summary}
cat("=== Annotation Pipeline Summary —", COMPARTMENT, "===\n\n")

cat("cluster_annotation (preprocessing):\n")
print(sort(table(df$cluster_annotation, useNA = "ifany"), decreasing = TRUE))

cat("\nSingleR L1 pruned (iHBCA 11 types):\n")
print(sort(table(df$pruned_label, useNA = "ifany"), decreasing = TRUE))

cat("\nHarmonized L1.0:\n")
print(sort(table(df$L1.0, useNA = "ifany"), decreasing = TRUE))

# Agreement
agree <- mean(df_complete$pruned_label == df_complete$L1.0, na.rm = TRUE)
cat(sprintf("\nSingleR L1 == Harmonized L1.0 (when labels line up): %.1f%%\n",
            100 * agree))

# Override summary
OVERRIDE_TYPES <- c("Adipo", "Mast", "Kerat")
for (ot in OVERRIDE_TYPES) {
  n <- sum(df$L1.0 == ot, na.rm = TRUE)
  if (n > 0) {
    singler_dist <- sort(table(df$pruned_label[df$L1.0 == ot]), decreasing = TRUE)
    cat(sprintf("\n%s (override): %d cells — was assigned by SingleR to:\n", ot, n))
    print(singler_dist)
  }
}
```
