---
title: "Step 10b: Artifact Discovery via Manifold Incoherence"
author: "Spatial HBCA Resource"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    theme: flatly
params:
  PROJECT_ROOT: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 300
)
```

# Overview

**Hypothesis**: Artifacts possess raw UMI counts but lack manifold structure. Under SNN smoothing, their signal collapses (negative delta), whereas real cells stabilize or denoise (delta ≈ 0 or positive).

**Method**:
1. **Cell-level detection**: 1D GMM on Δ = Smooth R² - Raw R²
2. **Cluster-level filtering**: Elbow detection on % artifact enrichment per cluster

**Input**: Step 10a discordance metrics (pearson_r2, N=400)

**Output**: `artifact_barcodes.txt` + `artifact_enriched_clusters.txt` for Steps 10c/10d

---

# Section 0: Load All Data

## 0.1 Libraries

```{r libraries}
suppressPackageStartupMessages({
  library(Seurat)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(mclust)
  library(zoo)
  library(yaml)
  library(here)
  library(ggrepel)
  library(cowplot)
})

theme_set(theme_bw(base_size = 12))
```

## 0.2 Configuration

```{r config}
# PROJECT_ROOT from params (pipeline) or here() (interactive)
PROJECT_ROOT <- params$PROJECT_ROOT
if (PROJECT_ROOT == "") {
  PROJECT_ROOT <- here::here()
}

config_path <- file.path(PROJECT_ROOT, "project/01_Preprocessing/config")
if (!dir.exists(config_path)) {
  stop("PROJECT_ROOT incorrect: ", PROJECT_ROOT)
}

cat("PROJECT_ROOT:", PROJECT_ROOT, "\n\n")

# Load module configs
module_config <- yaml::read_yaml(file.path(config_path, "module_configs.yaml"))

# Load Step 10b config from algorithm_params with defaults
if ("step_10b_artifact_discovery" %in% names(module_config$modules)) {
  cfg <- module_config$modules$step_10b_artifact_discovery$algorithm_params
} else {
  # Fallback defaults (should never be used in production)
  cfg <- list(
    metric = "pearson_r2",
    n_features = "N400",
    conditions = list(raw = "raw_scaled", smooth = "smooth_scaled"),
    clustering = list(
      fine_res = "leiden_scvi_5.0",
      coarse_res = "leiden_scvi_0.3"
    ),
    gmm_params = list(
      min_components = 2,
      max_components = 3,
      model_type = "V",
      posterior_threshold = 0.5
    ),
    coarse_elbow = list(
      search_ceiling_sd_multiplier = 1.0,
      min_cluster_size = 50,
      downsample_threshold = 1000,
      loess_target_neighbors = 50
    ),
    fine_enrichment = list(
      min_enrichment_search = 10.0,
      loess_span_bounds = list(min = 0.2, max = 0.5, neighbors = 10)
    )
  )
}

cat("=== Configuration ===\n")
cat("Metric:", cfg$metric, "| N features:", cfg$n_features, "\n")
cat("Clustering: Fine =", cfg$clustering$fine_res, "| Coarse =", cfg$clustering$coarse_res, "\n")
cat("GMM: G=", cfg$gmm_params$min_components, "-", cfg$gmm_params$max_components,
    " | Model=", cfg$gmm_params$model_type,
    " | Posterior threshold=", cfg$gmm_params$posterior_threshold, "\n", sep = "")
cat("Coarse Elbow: Min cluster size =", cfg$coarse_elbow$min_cluster_size,
    "| SD multiplier =", cfg$coarse_elbow$search_ceiling_sd_multiplier, "\n\n")
```

## 0.3 Load Manifest & Paths

```{r manifest}
manifest_path <- file.path(PROJECT_ROOT, "project/01_Preprocessing/config/preprocessing_manifest.yaml")
preprocessing_manifest <- yaml::read_yaml(manifest_path)

# Step 10a
if (!"step_10a_add_residual_normalized_data" %in% names(preprocessing_manifest)) {
  stop("Step 10a not found in manifest")
}

step_10a <- preprocessing_manifest$step_10a_add_residual_normalized_data
if (step_10a$status != "pass") {
  stop("Step 10a status: ", step_10a$status)
}

# Paths
SEURAT_PATH <- step_10a$outputs$integrated_with_bigsur
DISCORDANCE_BASE <- file.path(dirname(dirname(SEURAT_PATH)), "discordance/metrics")
METADATA_PATH <- file.path(PROJECT_ROOT, "project/01_Preprocessing/outputs/09_IntegrationPreview/metadata/unified_metadata.csv")

cat("=== Input Paths ===\n")
cat("Seurat:", SEURAT_PATH, "\n")
cat("Discordance base:", DISCORDANCE_BASE, "\n")
cat("Metadata:", METADATA_PATH, "\n\n")

# Validate
if (!file.exists(SEURAT_PATH)) stop("Seurat object not found")
if (!file.exists(METADATA_PATH)) stop("Metadata not found")
```

## 0.4 Load Seurat Object & Extract Metadata

```{r load_seurat}
cat("=== Loading Seurat Object ===\n")
cat("Loading from:", SEURAT_PATH, "\n")

seu <- readRDS(SEURAT_PATH)

cat("Cells:", ncol(seu), "\n")
cat("Assays:", paste(names(seu@assays), collapse = ", "), "\n")
cat("Size (GB):", round(object.size(seu) / 1e9, 2), "\n\n")

# Calculate percent.mt if not present
if (!"percent.mt" %in% colnames(seu@meta.data)) {
  cat("Calculating percent.mt...\n")
  seu[["percent.mt"]] <- PercentageFeatureSet(seu, pattern = "^MT-")
  cat("✓ percent.mt calculated\n")
} else {
  cat("✓ percent.mt already present\n")
}
cat("\n")

# Extract QC metrics
qc_cols <- c("nCount_RNA", "nFeature_RNA", "percent.mt")
qc_data <- seu@meta.data %>%
  select(any_of(qc_cols)) %>%
  tibble::rownames_to_column("cell_id")

# Extract UMAP
if ("umap" %in% names(seu@reductions)) {
  umap_coords <- Embeddings(seu, reduction = "umap")
  umap_data <- data.frame(
    cell_id = rownames(umap_coords),
    UMAP_1 = umap_coords[, 1],
    UMAP_2 = umap_coords[, 2]
  )
  cat("✓ UMAP extracted\n")
} else {
  umap_data <- data.frame(cell_id = colnames(seu))
  warning("UMAP not found")
}

# Extract clustering
leiden_cols <- grep("^leiden_scvi_", colnames(seu@meta.data), value = TRUE)
clustering_data <- seu@meta.data %>%
  select(any_of(leiden_cols)) %>%
  tibble::rownames_to_column("cell_id")

cat("✓ QC metrics extracted:", paste(intersect(qc_cols, colnames(qc_data)), collapse = ", "), "\n")
cat("✓ Clustering extracted:", paste(leiden_cols, collapse = ", "), "\n\n")

# Merge Seurat metadata
seurat_metadata <- qc_data %>%
  left_join(umap_data, by = "cell_id") %>%
  left_join(clustering_data, by = "cell_id")

cat("Seurat metadata ready:", nrow(seurat_metadata), "cells ×", ncol(seurat_metadata) - 1, "columns\n")
cat("Seurat object retained in memory for downstream analysis\n\n")
```

## 0.5 Load Discordance Metrics

```{r load_discordance}
cat("=== Loading Discordance Metrics ===\n")

# Construct paths
raw_path <- file.path(DISCORDANCE_BASE, cfg$metric, paste0(cfg$conditions$raw, ".csv"))
smooth_path <- file.path(DISCORDANCE_BASE, cfg$metric, paste0(cfg$conditions$smooth, ".csv"))

if (!file.exists(raw_path)) stop("Raw CSV not found: ", raw_path)
if (!file.exists(smooth_path)) stop("Smooth CSV not found: ", smooth_path)

# Load
raw_df <- read.csv(raw_path, stringsAsFactors = FALSE)
smooth_df <- read.csv(smooth_path, stringsAsFactors = FALSE)

# Extract N400 column and calculate delta
metrics_df <- data.frame(
  cell_id = raw_df$cell_id,
  raw_r2 = raw_df[[cfg$n_features]],
  smooth_r2 = smooth_df[[cfg$n_features]],
  stringsAsFactors = FALSE
)
metrics_df$delta <- metrics_df$smooth_r2 - metrics_df$raw_r2

# Calculate MAD-scaled delta (robust z-score)
delta_median <- median(metrics_df$delta, na.rm = TRUE)
delta_mad <- mad(metrics_df$delta, na.rm = TRUE)
metrics_df$delta_mad_scaled <- (metrics_df$delta - delta_median) / delta_mad

cat("Loaded:", nrow(metrics_df), "cells\n")
cat("Delta range: [", round(min(metrics_df$delta, na.rm = TRUE), 3),
    ", ", round(max(metrics_df$delta, na.rm = TRUE), 3), "]\n", sep = "")
cat("MAD-scaled delta range: [", round(min(metrics_df$delta_mad_scaled, na.rm = TRUE), 2),
    ", ", round(max(metrics_df$delta_mad_scaled, na.rm = TRUE), 2), "]\n\n", sep = "")
```

## 0.6 Load Step 09 Metadata & Merge All

```{r merge_all}
cat("=== Merging All Data ===\n")

# Load Step 09 metadata
step09_metadata <- read.csv(METADATA_PATH, stringsAsFactors = FALSE)
cat("Step 09 metadata:", nrow(step09_metadata), "cells\n")

# Merge: metrics + Seurat metadata + Step 09 metadata
analysis_data <- metrics_df %>%
  left_join(seurat_metadata, by = "cell_id") %>%
  left_join(
    step09_metadata %>% select(-matches("^leiden_scvi_|^UMAP")),
    by = "cell_id"
  )

cat("\n✓ Analysis data ready:", nrow(analysis_data), "cells ×", ncol(analysis_data), "columns\n")

# Verify both clustering resolutions exist
required_clusterings <- c(cfg$clustering$fine_res, cfg$clustering$coarse_res)
missing_clusterings <- setdiff(required_clusterings, colnames(analysis_data))

if (length(missing_clusterings) > 0) {
  stop("Missing required clustering columns: ", paste(missing_clusterings, collapse = ", "))
}

cat("Using fine resolution:", cfg$clustering$fine_res,
    "(", length(unique(analysis_data[[cfg$clustering$fine_res]])), "clusters)\n")
cat("Using coarse resolution:", cfg$clustering$coarse_res,
    "(", length(unique(analysis_data[[cfg$clustering$coarse_res]])), "clusters)\n\n")
```

## 0.7 Setup Output Directories

```{r output_dirs}
OUTPUT_BASE <- file.path(PROJECT_ROOT, "project/01_Preprocessing/outputs/10b_InteractiveExploration")

# Categorical directories (NOT timestamped)
dir.create(file.path(OUTPUT_BASE, "figures"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(OUTPUT_BASE, "exports"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(OUTPUT_BASE, "reports"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(OUTPUT_BASE, "logs"), recursive = TRUE, showWarnings = FALSE)

# Timestamped HTML report (like Step 09)
timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
OUTPUT_HTML <- file.path(OUTPUT_BASE, "reports", paste0("artifact_discovery_", timestamp, ".html"))

cat("Output base:", OUTPUT_BASE, "\n")
cat("HTML report:", OUTPUT_HTML, "\n\n")
```

## 0.8 Exploratory UMAP Visualizations

```{r umap_exploratory, fig.width=15, fig.height=10}
cat("=== Exploratory UMAP Visualizations ===\n\n")

# Prepare data with UMAP coordinates
umap_data <- analysis_data %>%
  filter(!is.na(UMAP_1), !is.na(UMAP_2), !is.na(raw_r2), !is.na(smooth_r2), !is.na(delta))

# 1. UMAP colored by Raw R²
p_umap_raw <- ggplot(umap_data, aes(x = UMAP_1, y = UMAP_2, color = raw_r2)) +
  geom_point(size = 0.3, alpha = 0.6) +
  scale_color_viridis_c(name = "Raw R²", option = "plasma") +
  labs(title = "Raw R² (Raw Scaled)", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
print(p_umap_raw)

# 2. UMAP colored by Smooth R²
p_umap_smooth <- ggplot(umap_data, aes(x = UMAP_1, y = UMAP_2, color = smooth_r2)) +
  geom_point(size = 0.3, alpha = 0.6) +
  scale_color_viridis_c(name = "Smooth R²", option = "plasma") +
  labs(title = "Smooth R² (Smooth Scaled)", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
print(p_umap_smooth)

# 3. UMAP colored by Delta (Smooth - Raw)
p_umap_delta <- ggplot(umap_data, aes(x = UMAP_1, y = UMAP_2, color = delta)) +
  geom_point(size = 0.3, alpha = 0.6) +
  scale_color_viridis_c(
    name = "Δ (Smooth - Raw)",  option = "plasma") +
  labs(title = "Delta (Smooth - Raw)", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
print(p_umap_delta)

# 4. Scatter: Raw vs. Smooth colored by Delta
p_scatter_landscape <- ggplot(umap_data, aes(x = raw_r2, y = smooth_r2, color = delta)) +
  geom_hex(bins = 80) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black", size = 0.8) +
    scale_fill_viridis_c(trans = "log10", name = "Cells") +
  labs(
    title = "Raw vs. Smooth (colored by Δ)",
    x = "Raw R²",
    y = "Smooth R²"
  ) +
  coord_fixed() +
  theme_bw(base_size = 12)
print(p_scatter_landscape)

# 5. Scatter: Delta vs. Raw R²
p_scatter_delta_raw <- ggplot(umap_data, aes(x = raw_r2, y = delta)) +
  geom_hex(bins = 80) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", size = 0.8) +
  scale_fill_viridis_c(trans = "log10", name = "Cells") +
  labs(
    title = "Δ vs. Raw R² (density)",
    x = "Raw R²",
    y = "Δ (Smooth - Raw)"
  ) +
  theme_bw(base_size = 12)
print(p_scatter_delta_raw)

# 6. Scatter: Delta vs. Smooth R²
p_scatter_delta_smooth <- ggplot(umap_data, aes(x = smooth_r2, y = delta)) +
  geom_hex(bins = 80) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", size = 0.8) +
  scale_fill_viridis_c(trans = "log10", name = "Cells") +
  labs(
    title = "Δ vs. Smooth R² (density)",
    x = "Smooth R²",
    y = "Δ (Smooth - Raw)"
  ) +
  theme_bw(base_size = 12)
print(p_scatter_delta_smooth)

# Arrange in grid
library(patchwork)

# Top row: 3 UMAPs
umap_row <- p_umap_raw + p_umap_smooth + p_umap_delta

# Bottom row: 3 scatter plots
scatter_row <- p_scatter_delta_raw + p_scatter_delta_smooth + p_scatter_landscape

# Combined
p_exploratory <- umap_row / scatter_row

print(p_exploratory)

# Save
ggsave(
  file.path(OUTPUT_BASE, "figures/00_exploratory_umaps.png"),
  p_exploratory,
  width = 15,
  height = 10,
  dpi = 300
)

cat("✓ Exploratory UMAPs saved\n\n")
```

---

# Section 1: Diagnostic Landscape

```{r landscape, fig.width=10, fig.height=9}
cat("=== Diagnostic Landscape ===\n\n")

plot_data <- analysis_data %>%
  filter(!is.na(raw_r2), !is.na(smooth_r2), is.finite(raw_r2), is.finite(smooth_r2))

p_landscape <- ggplot(plot_data, aes(x = raw_r2, y = smooth_r2)) +
  geom_hex(bins = 100) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
  scale_fill_viridis_c(trans = "log10", name = "Cells") +
  labs(
    title = "Raw vs Smooth R²",
    subtitle = paste(cfg$metric, "| N =", cfg$n_features),
    x = paste("Raw R² (", cfg$conditions$raw, ")", sep = ""),
    y = paste("Smooth R² (", cfg$conditions$smooth, ")", sep = "")
  ) +
  coord_fixed() +
  theme_bw(base_size = 14)

print(p_landscape)

ggsave(file.path(OUTPUT_BASE, "figures/01_landscape.png"), p_landscape, width = 10, height = 9, dpi = 300)
```

---

# Section 2: GMM Artifact Detection (Cell-Level)

## 2.1 Fit GMM

```{r gmm_fit}
cat("\n=== GMM Fitting ===\n")

gmm_data <- analysis_data %>%
  filter(!is.na(delta), is.finite(delta)) %>%
  pull(delta)

cat("Valid cells:", length(gmm_data), "\n")
cat("Delta - Mean:", round(mean(gmm_data), 4), "| Median:", round(median(gmm_data), 4), "\n\n")

# Fit GMM
gmm_fit <- Mclust(gmm_data,
                  G = cfg$gmm_params$min_components:cfg$gmm_params$max_components,
                  modelNames = cfg$gmm_params$model_type)

cat("GMM Results:\n")
cat("  G =", gmm_fit$G, "| BIC =", round(gmm_fit$bic, 2), "\n")

component_stats <- data.frame(
  component = 1:gmm_fit$G,
  mean = gmm_fit$parameters$mean,
  variance = gmm_fit$parameters$variance$sigmasq,
  proportion = gmm_fit$parameters$pro
)

print(component_stats)
cat("\n")
```

## 2.2 Identify Artifact Component

```{r gmm_classify}
cat("=== Artifact Classification (Global GMM) ===\n")

# Identify artifact component: lowest mean AND must be negative
# Identify artifact component: The component with the lowest mean Delta
# (Artifacts always have lower Delta than real cells, regardless of absolute zero)
artifact_component <- which.min(component_stats$mean)

# Safety Check (Soft)
if (component_stats$mean[artifact_component] > 0) {
  warning(paste0(
    "⚠️  The identified artifact component has a positive mean (", 
    round(component_stats$mean[artifact_component], 3), 
    "). Artifacts typically have negative Delta. Verify plots."
  ))
} else {
  cat("✓ Artifact component identified (Mean < 0)\n")
}

cat("Artifact component:", artifact_component, "\n")
cat("  Mean delta:", round(component_stats$mean[artifact_component], 4), "\n")
cat("  Proportion:", round(component_stats$proportion[artifact_component], 3), "\n\n")

# Classify cells
analysis_data$gmm_classification <- NA_integer_
analysis_data$gmm_posterior_artifact <- NA_real_

valid_idx <- which(!is.na(analysis_data$delta) & is.finite(analysis_data$delta))
analysis_data$gmm_classification[valid_idx] <- gmm_fit$classification
analysis_data$gmm_posterior_artifact[valid_idx] <- gmm_fit$z[, artifact_component]

# Flag artifacts (rename to is_artifact_global for clarity)
analysis_data$is_artifact_global <-
  (analysis_data$gmm_classification == artifact_component) &
  (analysis_data$gmm_posterior_artifact > cfg$gmm_params$posterior_threshold) &
  (analysis_data$delta < 0) &
  !is.na(analysis_data$gmm_posterior_artifact) &
  !is.na(analysis_data$delta)

n_artifacts_global <- sum(analysis_data$is_artifact_global, na.rm = TRUE)
pct_artifacts_global <- 100 * n_artifacts_global / nrow(analysis_data)

cat("Global GMM artifacts:", n_artifacts_global, "(", round(pct_artifacts_global, 2), "%)\n")
cat("Posterior threshold:", cfg$gmm_params$posterior_threshold, "\n")
cat("Delta requirement: < 0\n\n")
```

## 2.3 Visualize GMM

```{r gmm_plot, fig.width=12, fig.height=6}
# Density curves
x_seq <- seq(min(gmm_data), max(gmm_data), length.out = 500)

density_df <- lapply(1:gmm_fit$G, function(k) {
  data.frame(
    x = x_seq,
    density = gmm_fit$parameters$pro[k] *
      dnorm(x_seq, mean = gmm_fit$parameters$mean[k],
            sd = sqrt(gmm_fit$parameters$variance$sigmasq[k])),
    component = paste("Component", k),
    is_artifact = (k == artifact_component)
  )
}) %>% bind_rows()

p_gmm <- ggplot() +
  geom_histogram(data = data.frame(delta = gmm_data), aes(x = delta, y = after_stat(density)),
                 bins = 100, fill = "gray80", alpha = 0.5) +
  geom_line(data = density_df, aes(x = x, y = density, color = component, linetype = is_artifact),
            size = 1.2) +
  scale_linetype_manual(values = c("FALSE" = "solid", "TRUE" = "dashed"), guide = "none") +
  geom_vline(xintercept = component_stats$mean[artifact_component],
             color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "GMM Classification of Delta Distribution",
    subtitle = paste0("G = ", gmm_fit$G, " | Artifact component = ", artifact_component,
                      " (mean = ", round(component_stats$mean[artifact_component], 3), ")"),
    x = "Delta (Smooth R² - Raw R²)",
    y = "Density",
    color = "Component"
  ) +
  theme_bw(base_size = 14)

print(p_gmm)

ggsave(file.path(OUTPUT_BASE, "figures/02_gmm.png"), p_gmm, width = 12, height = 6, dpi = 300)
```

## 2.4 GMM Classification Visualizations

```{r gmm_status_plots}
cat("\n=== GMM Status Visualizations ===\n\n")

# Prepare data with status labels
status_data <- analysis_data %>%
  filter(!is.na(UMAP_1), !is.na(UMAP_2)) %>%
  mutate(
    status_label = ifelse(is_artifact_global, "Artifact", "Non-artifact"),
    status_label = factor(status_label, levels = c("Non-artifact", "Artifact"))
  )

# 1. UMAP colored by GMM status
p_umap_status <- ggplot(status_data, aes(x = UMAP_1, y = UMAP_2, color = status_label)) +
  geom_point(size = 0.3, alpha = 0.6) +
  scale_color_manual(
    name = "GMM Status",
    values = c("Non-artifact" = "gray70", "Artifact" = "red")
  ) +
  labs(title = "UMAP: GMM Classification", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
print(p_umap_status)

# 2. Raw vs. Smooth colored by status
p_scatter_status <- ggplot(status_data %>% filter(!is.na(raw_r2), !is.na(smooth_r2)),
                           aes(x = raw_r2, y = smooth_r2, color = status_label)) +
  geom_point(size = 0.5, alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black", size = 0.8) +
  scale_color_manual(
    name = "GMM Status",
    values = c("Non-artifact" = "gray70", "Artifact" = "red")
  ) +
  labs(
    title = "Raw vs. Smooth (colored by GMM status)",
    x = "Raw R²",
    y = "Smooth R²"
  ) +
  coord_fixed() +
  theme_bw(base_size = 12)
print(p_scatter_status)

# 3. Delta vs. Raw colored by status
p_delta_raw_status <- ggplot(status_data %>% filter(!is.na(raw_r2), !is.na(delta)),
                             aes(x = raw_r2, y = delta, color = status_label)) +
  geom_point(size = 0.5, alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.8) +
  scale_color_manual(
    name = "GMM Status",
    values = c("Non-artifact" = "gray70", "Artifact" = "red")
  ) +
  labs(
    title = "Δ vs. Raw R² (colored by GMM status)",
    x = "Raw R²",
    y = "Δ (Smooth - Raw)"
  ) +
  theme_bw(base_size = 12)
print(p_delta_raw_status)

# Combined status plots
library(patchwork)
p_status_combined <- p_umap_status + p_scatter_status + p_delta_raw_status

print(p_status_combined)

# Save
ggsave(
  file.path(OUTPUT_BASE, "figures/02b_gmm_status_visualizations.png"),
  p_status_combined,
  width = 15,
  height = 5,
  dpi = 300
)

cat("✓ GMM status visualizations saved\n\n")
```

---

# Section 3: Hierarchical Artifact Detection (Coarse → Fine)

## 3.3 Per-Cluster Elbow Detection (DISABLED)

```{r per_cluster_elbow, fig.width=16, fig.height=24}
cat("\n=== Local Elbow Detection: DISABLED ===\n")
cat("Rationale: GMM-only detection provides cleaner separation.\n")
cat("Local elbow was flagging disproportionate cells in healthy clusters.\n")
cat("(22,814 local vs 5,072 GMM with only 912 overlap)\n\n")

# Set placeholder columns for downstream compatibility
analysis_data$is_artifact_local <- FALSE
analysis_data$elbow_delta_coarse <- NA_real_

cat("is_artifact_local = FALSE for all cells\n")
cat("elbow_delta_coarse = NA for all cells\n\n")

# =============================================================================
# COMMENTED OUT: Original per-cluster elbow detection code
# Retained for reference/potential future re-enablement
# =============================================================================

# # 1. Setup
# major_cluster_col <- cfg$clustering$coarse_res
#
# if (!major_cluster_col %in% colnames(analysis_data)) {
#   stop("Clustering column not found.")
# }
#
# # Define Search Bounds
# # Floor: The mean of the global artifact component (center of the debris cloud)
# # Ceiling: 0.1 (just entering the healthy biological variation zone)
# search_floor <- component_stats$mean[artifact_component]
# search_ceiling <- 0.1
#
# cat("Using clustering:", major_cluster_col, "\n")
# cat("Search Range:", round(search_floor, 4), "to", search_ceiling, "\n")
#
# # Get clusters
# clusters <- unique(analysis_data[[major_cluster_col]][!is.na(analysis_data[[major_cluster_col]])])
#
# # 2. Run Detection
# elbow_results_coarse <- lapply(clusters, function(clust) {
#   # Filter to cluster and Bounded Range
#   cluster_cells <- analysis_data %>%
#     filter(!!sym(major_cluster_col) == clust,
#            !is.na(delta),
#            delta >= search_floor,   # Ignore deep tail
#            delta <= search_ceiling) # Ignore healthy plateau
#
#
#   # Rank Data (Ascending: Artifact Mean -> Healthy)
#   fit_data <- cluster_cells %>%
#     arrange(delta) %>%
#     mutate(rank = row_number()) %>%
#     select(rank, delta)
#
#   # --- GEOMETRIC ELBOW (Kneedle) ---
#
#   # 1. Define Chord (Start to End of the bounded region)
#   min_pt <- fit_data[1, ]
#   max_pt <- fit_data[nrow(fit_data), ]
#
#   # 2. Vector of the Chord
#   chord_vec <- c(max_pt$rank - min_pt$rank, max_pt$delta - min_pt$delta)
#
#   # 3. Project vectors from Start to all points
#   vec_x <- fit_data$rank - min_pt$rank
#   vec_y <- fit_data$delta - min_pt$delta
#
#   # 4. Calculate Perpendicular Distance
#   cross_prod <- vec_x * chord_vec[2] - vec_y * chord_vec[1]
#   line_len <- sqrt(sum(chord_vec^2))
#   dist <- abs(cross_prod) / line_len
#
#   # 5. Find Maximum Distance
#   elbow_idx <- which.max(dist)
#
#   # Extract
#   elbow_delta <- fit_data$delta[elbow_idx]
#   elbow_rank <- fit_data$rank[elbow_idx] # Rank within the filtered subset
#
#   data.frame(
#     cluster = clust,
#     elbow_delta = elbow_delta,
#     elbow_rank = elbow_rank,
#     status = "success"
#   )
# }) %>% bind_rows()
#
# cat("Clusters analyzed:", nrow(elbow_results_coarse), "\n")
# cat("Successful fits:", sum(elbow_results_coarse$status == "success"), "\n\n")
#
# # 3. Merge & Flag
# if ("elbow_delta" %in% colnames(analysis_data)) analysis_data$elbow_delta <- NULL
#
# analysis_data <- analysis_data %>%
#   left_join(elbow_results_coarse %>% select(cluster, elbow_delta_coarse = elbow_delta),
#             by = setNames("cluster", major_cluster_col))
#
# analysis_data$is_artifact_local <-
#   !is.na(analysis_data$elbow_delta_coarse) &
#   !is.na(analysis_data$delta) &
#   analysis_data$delta < analysis_data$elbow_delta_coarse
#
# # --- VISUALIZATION ---
#
# # A. Calculate Removal Stats
# removal_stats <- analysis_data %>%
#   filter(!is.na(!!sym(major_cluster_col))) %>%
#   group_by(!!sym(major_cluster_col)) %>%
#   summarise(
#     n_total = n(),
#     n_removed = sum(is_artifact_local, na.rm=TRUE),
#     pct_removed = 100 * n_removed / n_total,
#     elbow_val = unique(elbow_delta_coarse)
#   ) %>%
#   rename(cluster = !!sym(major_cluster_col)) %>%
#   arrange(desc(pct_removed))
#
# # B. Plot Thresholds
# p_thresh <- ggplot(removal_stats, aes(x=reorder(cluster, elbow_val), y=elbow_val)) +
#   geom_col(fill="steelblue") +
#   geom_hline(yintercept = search_floor, linetype="dotted", color="red") +
#   geom_hline(yintercept = search_ceiling, linetype="dotted", color="gray40") +
#   labs(title="Per-Cluster Elbow Thresholds", subtitle="Dotted lines show search range", x="Cluster", y="Delta Cutoff") +
#   theme_bw() + theme(axis.text.x = element_text(angle=45, hjust=1))
#
# # C. Plot Removal Rates
# p_rates <- ggplot(removal_stats, aes(x=reorder(cluster, -pct_removed), y=pct_removed)) +
#   geom_col(fill="firebrick") +
#   labs(title="Percentage Removed per Cluster", x="Cluster", y="% Artifacts") +
#   theme_bw() + theme(axis.text.x = element_text(angle=45, hjust=1))
#
# # D. UMAP Comparison (Provenance)
# umap_removal_data <- analysis_data %>%
#   filter(!is.na(UMAP_1), !is.na(UMAP_2)) %>%
#   mutate(
#     removal_category = case_when(
#       is_artifact_global & is_artifact_local ~ "Both methods",
#       is_artifact_global & !is_artifact_local ~ "GMM only",
#       !is_artifact_global & is_artifact_local ~ "Per-cluster elbow only",
#       TRUE ~ "Not removed"
#     ),
#     removal_category = factor(removal_category, levels = c("Not removed", "GMM only", "Per-cluster elbow only", "Both methods"))
#   )
#
# p_umap_removal <- ggplot(umap_removal_data, aes(x = UMAP_1, y = UMAP_2, color = removal_category)) +
#   geom_point(size = 0.3, alpha = 0.5) +
#   scale_color_manual(
#     name = "Removal Status",
#     values = c(
#       "Not removed" = "gray80",
#       "GMM only" = "orange",
#       "Per-cluster elbow only" = "blue",
#       "Both methods" = "red"
#     )
#   ) +
#   labs(title = "UMAP: Removal Method Comparison", x = "UMAP 1", y = "UMAP 2") +
#   theme_minimal(base_size = 12) +
#   guides(color = guide_legend(override.aes = list(size = 3)))
#
# # E. Elbow Curves (All Clusters Overlaid)
# curve_plot_data <- analysis_data %>%
#   filter(!is.na(!!sym(major_cluster_col)), delta >= search_floor, delta <= search_ceiling) %>%
#   select(cluster = !!sym(major_cluster_col), delta) %>%
#   group_by(cluster) %>%
#   arrange(delta) %>%
#   mutate(
#     rank = row_number(),
#     norm_rank = row_number() / n()
#   ) %>%
#   slice(unique(round(seq(1, n(), length.out = 200))))
#
# plot_elbows <- curve_plot_data %>%
#   inner_join(elbow_results_coarse %>% select(cluster, elbow_delta), by="cluster") %>%
#   group_by(cluster) %>%
#   filter(abs(delta - elbow_delta) == min(abs(delta - elbow_delta))) %>%
#   slice(1)
#
# p_curves <- ggplot() +
#   geom_line(data = curve_plot_data,
#             aes(x = norm_rank, y = delta, color = cluster, group = cluster),
#             alpha = 0.5, size = 0.8) +
#   geom_point(data = plot_elbows,
#              aes(x = norm_rank, y = delta),
#              color = "black", shape = 4, size = 3, stroke = 1.5) +
#   labs(
#     title = "Elbow Detection Curves (All Clusters Overlaid)",
#     subtitle = "Normalized Rank (0-1) vs Delta. 'X' marks the calculated Artifact threshold.",
#     y = "Delta",
#     x = "Normalized Rank (within search window)"
#   ) +
#   theme_bw(base_size = 14) +
#   theme(legend.position = "right") +
#   guides(color = guide_legend(ncol = 1, override.aes = list(size = 2)))
#
# # Combine
# top_row <- cowplot::plot_grid(p_thresh, p_rates, ncol=2)
# final_plot <- cowplot::plot_grid(top_row, p_umap_removal, p_curves, ncol=1, rel_heights = c(1, 1.2, 2))
#
# print(final_plot)
#
# ggsave(file.path(OUTPUT_BASE, "figures/03b_per_cluster_elbows.png"),
#        final_plot, width = 16, height = 24)
#
# cat("Local detection complete. Summary:\n")
# print(head(removal_stats, 5))
```

## 3.2 Union Flag (Total Artifacts)

```{r union_flag}
cat("=== Creating Total Artifact Flag (GMM Only - Local Elbow Disabled) ===\n")

# GMM only (local elbow is disabled, is_artifact_local = FALSE for all cells)
analysis_data$is_artifact_total <-
  analysis_data$is_artifact_global | analysis_data$is_artifact_local

# Summary
n_global <- sum(analysis_data$is_artifact_global, na.rm = TRUE)
n_local <- sum(analysis_data$is_artifact_local, na.rm = TRUE)  # Should be 0
n_total <- sum(analysis_data$is_artifact_total, na.rm = TRUE)

cat("Global GMM artifacts:", n_global, "\n")
cat("Local Elbow artifacts:", n_local, "(disabled)\n")
cat("Total artifacts:     ", n_total, " (",
    round(100 * n_total / nrow(analysis_data), 2), "%)\n\n", sep = "")

# Simplified UMAP (GMM only since local elbow is disabled)
overlap_data <- analysis_data %>%
  filter(!is.na(UMAP_1), !is.na(UMAP_2)) %>%
  mutate(
    status = ifelse(is_artifact_global, "Artifact (GMM)", "Passed"),
    status = factor(status, levels = c("Passed", "Artifact (GMM)"))
  )

p_overlap <- ggplot(overlap_data, aes(x = UMAP_1, y = UMAP_2, color = status)) +
  geom_point(size = 0.3, alpha = 0.5) +
  scale_color_manual(
    name = "Status",
    values = c(
      "Passed" = "gray80",
      "Artifact (GMM)" = "red"
    )
  ) +
  labs(title = "UMAP: GMM Artifact Detection (Local Elbow Disabled)") +
  theme_minimal(base_size = 12) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))

print(p_overlap)

ggsave(
  file.path(OUTPUT_BASE, "figures/03b_gmm_artifacts.png"),
  p_overlap,
  width = 10,
  height = 8,
  dpi = 300
)

cat("Total artifact flag created (GMM only)\n\n")
```

## 3.3 Delta Distribution by Coarse Clusters

```{r delta_by_coarse, fig.width=14, fig.height=11}
cat("=== Delta Distribution by Coarse Clusters ===\n")

coarse_col <- cfg$clustering$coarse_res

# Prepare data
violin_data <- analysis_data %>%
  filter(!is.na(!!sym(coarse_col)), !is.na(delta), is.finite(delta),
         !is.na(delta_mad_scaled), is.finite(delta_mad_scaled)) %>%
  group_by(!!sym(coarse_col)) %>%
  mutate(
    cluster_size = n(),
    median_delta = median(delta, na.rm = TRUE),
    median_delta_mad = median(delta_mad_scaled, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  arrange(median_delta) %>%
  mutate(cluster_label = factor(!!sym(coarse_col),
                                levels = unique(!!sym(coarse_col))))

n_clusters <- length(unique(violin_data[[coarse_col]]))
cat("Coarse clusters:", n_clusters, "\n")

# Violin plot - Raw delta
p_violin_raw <- ggplot(violin_data, aes(x = cluster_label, y = delta, fill = median_delta)) +
  geom_violin(scale = "width", trim = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", size = 0.8) +
  scale_fill_gradient2(
    name = "Median Δ",
    low = "blue", mid = "gray90", high = "red",
    midpoint = 0
  ) +
  labs(
    title = "Delta Distribution by Coarse Clusters (Raw)",
    subtitle = paste0("Clustering: ", coarse_col, " (ordered by median Δ)"),
    x = "Cluster",
    y = "Δ (Smooth - Raw)"
  ) +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_violin_raw)

# Violin plot - MAD-scaled delta
p_violin_mad <- ggplot(violin_data, aes(x = cluster_label, y = delta_mad_scaled, fill = median_delta_mad)) +
  geom_violin(scale = "width", trim = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", size = 0.8) +
  geom_hline(yintercept = c(-3, 3), linetype = "dotted", color = "gray40", size = 0.5) +
  scale_fill_gradient2(
    name = "Median Δ (MAD)",
    low = "blue", mid = "gray90", high = "red",
    midpoint = 0
  ) +
  labs(
    title = "Delta Distribution by Coarse Clusters (MAD-scaled)",
    subtitle = paste0("Clustering: ", coarse_col, " (±3 MAD reference lines)"),
    x = "Cluster",
    y = "Δ (MAD-scaled)"
  ) +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_violin_mad)

# Combined plot
library(patchwork)
p_violin_combined <- p_violin_raw / p_violin_mad

print(p_violin_combined)

# Save
ggsave(
  file.path(OUTPUT_BASE, "figures/03c_delta_by_coarse_clusters.png"),
  p_violin_combined,
  width = 14,
  height = 11,
  dpi = 300
)

cat("✓ Violin plots saved (raw + MAD-scaled)\n\n")
```

## 3.4 Fine-Resolution Cluster Enrichment (Using Union Flag)

```{r fine_enrichment}
cat("\n=== Fine-Resolution Cluster Enrichment Analysis ===\n")
cat("Strategy: Identify trash clusters using TOTAL artifacts (global + local)\n\n")

fine_col <- cfg$clustering$fine_res
cat("Using fine clustering:", fine_col, "\n")

# Calculate enrichment using UNION flag (is_artifact_total)
cluster_stats_fine <- analysis_data %>%
  group_by(!!sym(fine_col)) %>%
  summarise(
    n_cells = n(),
    n_artifacts = sum(is_artifact_total, na.rm = TRUE),  # KEY: Use union flag
    pct_artifact = 100 * n_artifacts / n_cells,
    .groups = "drop"
  ) %>%
  rename(cluster_id = !!sym(fine_col)) %>%
  arrange(desc(pct_artifact))

cat("Fine clusters:", nrow(cluster_stats_fine), "\n")
cat("Enrichment range:", round(min(cluster_stats_fine$pct_artifact), 2), "% -",
    round(max(cluster_stats_fine$pct_artifact), 2), "%\n\n")

cat("Top 10 enriched clusters:\n")
print(head(cluster_stats_fine, 10), row.names = FALSE)
cat("\n")
```

## 3.5 Fine Enrichment Elbow Detection

```{r fine_elbow}
cat("=== Fine Enrichment Elbow Detection ===\n")

# Rank by enrichment (descending: high to low)
cluster_stats_fine <- cluster_stats_fine %>%
  arrange(desc(pct_artifact)) %>%
  mutate(rank = row_number())

# Define search bounds
min_enrichment_threshold <- cfg$fine_enrichment$min_enrichment_search
search_end_idx <- max(which(cluster_stats_fine$pct_artifact >= min_enrichment_threshold))
search_end_idx <- min(search_end_idx, ceiling(nrow(cluster_stats_fine) * 0.5))

if (search_end_idx < 5) {
  # Fallback: use quantile-based threshold
  elbow_threshold_fine <- quantile(cluster_stats_fine$pct_artifact, 0.75, na.rm = TRUE)
  elbow_rank_fine <- min(which(cluster_stats_fine$pct_artifact >= elbow_threshold_fine))
  cat("⚠️  Insufficient search range, using 75th percentile threshold\n")
} else {
  # Extract search region
  search_data <- cluster_stats_fine[1:search_end_idx, ]
  
  # Loess smoothing (adaptive span)
  x_vals <- seq_len(nrow(search_data))
  span_bounds <- cfg$fine_enrichment$loess_span_bounds
  span_value <- min(span_bounds$max, max(span_bounds$min, span_bounds$neighbors / length(x_vals)))
  
  loess_fit <- loess(search_data$pct_artifact ~ x_vals, span = span_value)
  smoothed <- predict(loess_fit, x_vals)
  
  # Calculate derivatives
  first_diff <- diff(smoothed)
  second_diff <- diff(first_diff)
  
  # Find steepest drop (most negative second derivative for descending curve)
  elbow_in_search <- which.min(second_diff) + 2
  elbow_in_search <- min(elbow_in_search, nrow(search_data))
  
  elbow_rank_fine <- search_data$rank[elbow_in_search]
  elbow_threshold_fine <- search_data$pct_artifact[elbow_in_search]
  
  # Store smoothed values
  cluster_stats_fine$pct_smoothed <- NA_real_
  cluster_stats_fine$pct_smoothed[1:search_end_idx] <- smoothed
}

cat("Search range: Top", search_end_idx, "clusters (>=",
    round(min_enrichment_threshold, 2), "% enrichment)\n")
cat("Elbow detected at rank:", elbow_rank_fine, "\n")
cat("Elbow threshold:", round(elbow_threshold_fine, 2), "%\n")
cat("Clusters above elbow:", sum(cluster_stats_fine$pct_artifact >= elbow_threshold_fine), "\n\n")
```

## 3.6 Visualize Fine Enrichment

```{r fine_enrichment_plot, fig.width=14, fig.height=6}
# Rank curve
p_rank_fine <- ggplot(cluster_stats_fine, aes(x = rank, y = pct_artifact)) +
  geom_line(color = "gray40") +
  geom_point(color = "gray40", size = 1.5) +
  geom_line(aes(y = pct_smoothed), color = "blue", size = 1.2, na.rm = TRUE) +
  geom_hline(yintercept = elbow_threshold_fine, color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = elbow_rank_fine, color = "red", linetype = "dotted") +
  annotate("text", x = elbow_rank_fine, y = max(cluster_stats_fine$pct_artifact) * 0.9,
           label = paste0("Elbow\n", round(elbow_threshold_fine, 2), "%"),
           hjust = -0.1, color = "red", size = 4) +
  labs(
    title = "Fine Cluster Enrichment Rank Curve",
    subtitle = paste0("Using ", fine_col, " | Enrichment calculated from TOTAL artifacts"),
    x = "Cluster Rank",
    y = "% Artifact Cells"
  ) +
  theme_bw(base_size = 14)

# Bar plot
cluster_stats_fine$above_elbow <- cluster_stats_fine$pct_artifact >= elbow_threshold_fine

p_bar_fine <- ggplot(cluster_stats_fine, aes(x = reorder(cluster_id, -pct_artifact),
                                             y = pct_artifact, fill = above_elbow)) +
  geom_col() +
  geom_hline(yintercept = elbow_threshold_fine, color = "red", linetype = "dashed", size = 1) +
  scale_fill_manual(values = c("FALSE" = "gray60", "TRUE" = "firebrick"), name = "Above Elbow") +
  labs(
    title = "Fine Cluster Artifact Enrichment",
    x = "Cluster ID (ranked)",
    y = "% Artifact"
  ) +
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6))

p_combined_fine <- cowplot::plot_grid(p_rank_fine, p_bar_fine, ncol = 2)
print(p_combined_fine)

ggsave(file.path(OUTPUT_BASE, "figures/03d_fine_cluster_enrichment.png"),
       p_combined_fine, width = 14, height = 6, dpi = 300)
```

## 3.7 Flag Trash Clusters

```{r flag_trash_clusters}
trash_clusters <- cluster_stats_fine %>%
  filter(pct_artifact >= elbow_threshold_fine) %>%
  pull(cluster_id)

analysis_data$is_trash_cluster <-
  analysis_data[[fine_col]] %in% trash_clusters

n_cells_trash <- sum(analysis_data$is_trash_cluster, na.rm = TRUE)

cat("Flagged trash clusters:", length(trash_clusters), "\n")
cat("Cells in trash clusters:", n_cells_trash,
    "(", round(100 * n_cells_trash / nrow(analysis_data), 2), "%)\n\n")
```

---

# Section 4: Validation

## 4.1 UMAP Validation

```{r umap_validation, fig.width=16, fig.height=10}
umap_data <- analysis_data %>% filter(!is.na(UMAP_1), !is.na(UMAP_2))

# Global artifacts
p1 <- ggplot(umap_data, aes(x = UMAP_1, y = UMAP_2, color = is_artifact_global)) +
  geom_point(size = 0.3, alpha = 0.5) +
  scale_color_manual(values = c("FALSE" = "gray80", "TRUE" = "blue")) +
  labs(title = "Global Artifacts (GMM)", color = "Artifact") +
  coord_fixed() +
  theme_bw()
print(p1)

# Local artifacts
p2 <- ggplot(umap_data, aes(x = UMAP_1, y = UMAP_2, color = is_artifact_local)) +
  geom_point(size = 0.3, alpha = 0.5) +
  scale_color_manual(values = c("FALSE" = "gray80", "TRUE" = "orange")) +
  labs(title = "Local Artifacts (Coarse Elbow)", color = "Artifact") +
  coord_fixed() +
  theme_bw()
print(p2)

# Total artifacts (union)
p3 <- ggplot(umap_data, aes(x = UMAP_1, y = UMAP_2, color = is_artifact_total)) +
  geom_point(size = 0.3, alpha = 0.5) +
  scale_color_manual(values = c("FALSE" = "gray80", "TRUE" = "firebrick")) +
  labs(title = "Total Artifacts (Union)", color = "Artifact") +
  coord_fixed() +
  theme_bw()
print(p3)

# Trash clusters
p4 <- ggplot(umap_data, aes(x = UMAP_1, y = UMAP_2, color = is_trash_cluster)) +
  geom_point(size = 0.3, alpha = 0.5) +
  scale_color_manual(values = c("FALSE" = "gray80", "TRUE" = "darkred")) +
  labs(title = "Trash Clusters (Fine Enrichment)", color = "Trash") +
  coord_fixed() +
  theme_bw()
print(p4)


seu$is_debris <- umap_data$is_artifact_total == TRUE | 
  umap_data$is_trash_cluster == TRUE

UMAPPlot(seu, group.by = "is_debris", shuffle = TRUE, raster = FALSE,
         alpha = 0.4, pt.size = 0.3)
UMAPPlot(seu, split.by = "is_debris", group.by = "HBCATransferredLabels.Kumar_2023", label = TRUE, raster = FALSE,
         alpha = 0.4, pt.size = 0.3) + NoLegend()

p_umap <- cowplot::plot_grid(p1, p2, p3, p4, ncol = 2)
print(p_umap)



ggsave(file.path(OUTPUT_BASE, "figures/04_umap_validation.png"), p_umap,
       width = 16, height = 10, dpi = 300)
```

## 4.2 UMI Comparison

```{r UMI, fig.width=10, fig.height=8}
seu@meta.data$delta <- analysis_data$delta
UMI_data <- seu@meta.data %>%
  filter(!is.na(nCount_RNA), !is.na(delta))

p_UMI <- ggplot(UMI_data, aes(x = nCount_RNA, y = delta)) +
  geom_hex(bins = 80) +
  scale_fill_viridis_c(trans = "log10") +
  labs(
    title = "UMIs vs Delta") +
  theme_bw(base_size = 14) + scale_x_log10()

print(p_UMI)

ggsave(file.path(OUTPUT_BASE, "figures/05_UMI.png"), p_UMI,
       width = 10, height = 8, dpi = 300)

```

## 4.3 QC Metrics

```{r qc, fig.width=14, fig.height=5}
qc_data <- analysis_data %>%
  select(is_artifact_total, nCount_RNA, nFeature_RNA, percent.mt) %>%
  filter(!is.na(is_artifact_total)) %>%
  pivot_longer(cols = c(nCount_RNA, nFeature_RNA, percent.mt), names_to = "metric", values_to = "value")

# Calculate q95 bounds for each metric (global across both groups)
qc_bounds <- qc_data %>%
  group_by(metric) %>%
  summarize(
    q95 = quantile(value, 0.99, na.rm = TRUE),
    .groups = "drop"
  )

# Filter data to q95 range for clearer visualization
qc_data_filtered <- qc_data %>%
  left_join(qc_bounds, by = "metric") %>%
  filter(value <= q95)

p_qc <- ggplot(qc_data_filtered, aes(x = is_artifact_total, y = value, fill = is_artifact_total)) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.1, outlier.shape = NA) +
  # Add q95 horizontal reference line
  geom_hline(data = qc_bounds, aes(yintercept = q95),
             linetype = "dashed", color = "black", linewidth = 0.5, alpha = 0.7) +
  scale_fill_manual(values = c("FALSE" = "gray60", "TRUE" = "firebrick")) +
  facet_wrap(~ metric, scales = "free_y", ncol = 3) +
  labs(title = "QC Metrics: Total Artifacts vs Non-Artifact (filtered to Q99)",
       x = "Artifact (Total)", y = "Value",
       caption = "Dashed line = 95th percentile; values above Q99 excluded for clarity") +
  theme_bw(base_size = 14)

print(p_qc)

ggsave(file.path(OUTPUT_BASE, "figures/06_qc.png"), p_qc, width = 14, height = 5, dpi = 300)
```

---

# Section 5: Export & Provenance

## 5.1 Primary Outputs

```{r export_primary}
cat("\n=== Finalizing Decisions & Provenance ===\n")

# Define resolution columns dynamically
fine_col <- cfg$clustering$fine_res
coarse_col <- cfg$clustering$coarse_res

# 1. Construct Decision Table with Waterfall Provenance
final_decisions <- analysis_data %>%
  mutate(
    # UNION LOGIC: Remove if flagged by ANY method
    to_remove = is_artifact_global | is_artifact_local | is_trash_cluster,
    
    # DETAILED PROVENANCE: Why was it removed?
    # NOTE: Local elbow detection disabled - only GMM and trash cluster reasons remain
    removal_reason = case_when(
      # Hierarchical priority: Most severe first
      is_trash_cluster ~ "TrashCluster_FinePocket",
      is_artifact_global ~ "Global_Incoherence_GMM",
      # is_artifact_global & is_artifact_local ~ "Both_Global_Local",  # Disabled
      # is_artifact_local ~ "Local_Outlier_CoarseElbow",  # Disabled
      TRUE ~ "Pass"
    ),
    
    # Traceability metadata
    gmm_posterior = round(gmm_posterior_artifact, 4),
    coarse_cluster = !!sym(coarse_col),
    coarse_elbow = round(elbow_delta_coarse, 4),
    fine_cluster = !!sym(fine_col)
  ) %>%
  select(
    # Identifiers
    cell_id, sample_id, patient_id,
    # Decision
    to_remove, removal_reason,
    # Metrics
    delta, raw_r2, smooth_r2, delta_mad_scaled,
    # Flags (for debugging)
    is_artifact_global, is_artifact_local, is_trash_cluster,
    # GMM details
    gmm_classification, gmm_posterior,
    # Coarse elbow details
    coarse_cluster, coarse_elbow,
    # Fine cluster details
    fine_cluster,
    # QC comparisons
    nCount_RNA, nFeature_RNA, percent.mt, scDblFinder_score
  )

cat("✓ Decision table ready:", nrow(final_decisions), "cells\n")
cat("  To remove:", sum(final_decisions$to_remove),
    "(", round(100 * sum(final_decisions$to_remove) / nrow(final_decisions), 2), "%)\n\n")

# 2. Export Artifact Barcodes (Primary List for Step 10c/10d)
artifact_barcodes <- final_decisions %>%
  filter(to_remove) %>%
  pull(cell_id)

writeLines(artifact_barcodes, file.path(OUTPUT_BASE, "exports/artifact_barcodes.txt"))
cat("✓ artifact_barcodes.txt (n =", length(artifact_barcodes), ")\n")

# 3. Export Trash Clusters List
writeLines(as.character(trash_clusters),
           file.path(OUTPUT_BASE, "exports/trash_clusters.txt"))
cat("✓ trash_clusters.txt (n =", length(trash_clusters), ")\n")

# 4. Export Manifest Update (for Central Cell Status CSV)
manifest_update <- final_decisions %>%
  transmute(
    cell_id = cell_id,
    sample_id = sample_id,
    patient_id = patient_id,
    step_10b_manifold_pass = !to_remove,  # TRUE = Keep, FALSE = Remove
    step_10b_removal_reason = removal_reason
  )

write.csv(manifest_update,
          file.path(OUTPUT_BASE, "exports/step10b_manifest_update.csv"),
          row.names = FALSE)
cat("✓ step10b_manifest_update.csv (for central manifest integration)\n")

# 5. Export Full Decision Matrix (for debugging)
write.csv(final_decisions,
          file.path(OUTPUT_BASE, "exports/full_decision_matrix.csv"),
          row.names = FALSE)
cat("✓ full_decision_matrix.csv (", nrow(final_decisions), " cells ×",
    ncol(final_decisions), " columns)\n", sep = "")

# 6. Export Coarse Elbow Thresholds - DISABLED (local elbow detection removed)
# if (exists("elbow_results_coarse") && nrow(elbow_results_coarse) > 0) {
#   coarse_thresholds <- elbow_results_coarse %>%
#     filter(status == "success") %>%
#     select(cluster, elbow_delta, elbow_rank, status) %>%
#     rename(!!coarse_col := cluster)
#
#   write.csv(
#     coarse_thresholds,
#     file.path(OUTPUT_BASE, "exports/coarse_elbow_thresholds.csv"),
#     row.names = FALSE
#   )
#   cat("coarse_elbow_thresholds.csv (", nrow(coarse_thresholds), " clusters)\n", sep = "")
# }
cat("coarse_elbow_thresholds.csv: SKIPPED (local elbow disabled)\n")

cat("\n")
```

## 5.2 Summary Report

```{r summary_report}
# Breakdown by removal reason
reason_summary <- final_decisions %>%
  group_by(removal_reason) %>%
  summarise(
    n_cells = n(),
    pct = round(100 * n() / nrow(final_decisions), 2),
    .groups = "drop"
  ) %>%
  arrange(desc(n_cells))

cat("=== Removal Reason Breakdown ===\n")
print(reason_summary, row.names = FALSE)
cat("\n")

# Summary YAML
summary_report <- list(
  timestamp = as.character(Sys.time()),
  config = list(
    metric = cfg$metric,
    n_features = cfg$n_features,
    fine_resolution = cfg$clustering$fine_res,
    coarse_resolution = cfg$clustering$coarse_res
  ),
  gmm_global = list(
    G = gmm_fit$G,
    artifact_component = artifact_component,
    artifact_mean = component_stats$mean[artifact_component],
    posterior_threshold = cfg$gmm_params$posterior_threshold,
    n_flagged = sum(final_decisions$is_artifact_global, na.rm = TRUE)
  ),
  # coarse_elbow_local DISABLED - local elbow detection removed
  coarse_elbow_local = list(
    status = "disabled",
    reason = "Local elbow flagged disproportionate cells in healthy clusters",
    n_flagged = 0
  ),
  fine_enrichment = list(
    total_clusters = nrow(cluster_stats_fine),
    trash_clusters = length(trash_clusters),
    elbow_threshold = round(elbow_threshold_fine, 2),
    n_cells_in_trash = sum(final_decisions$is_trash_cluster, na.rm = TRUE)
  ),
  final_decision = list(
    total_cells = nrow(final_decisions),
    to_remove = sum(final_decisions$to_remove),
    to_keep = sum(!final_decisions$to_remove),
    pct_removed = round(100 * sum(final_decisions$to_remove) / nrow(final_decisions), 2),
    reason_breakdown = setNames(reason_summary$n_cells, reason_summary$removal_reason)
  )
)

write_yaml(summary_report, file.path(OUTPUT_BASE, "exports/summary.yaml"))
cat("✓ summary.yaml saved\n\n")
```

---
