---
title: "Step 13: Filtered Integration Preview - Post-Artifact Removal Assessment"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 10,
  fig.height = 8
)

# Load required libraries
suppressPackageStartupMessages({
  library(Seurat)
  library(ggplot2)
  library(dplyr)
  library(presto)
  library(patchwork)
  library(viridis)
  library(yaml)
})

cat("Libraries loaded\n")
```

# Setup Project Root

```{r project_root}
# Use here package to navigate to preprocessing directory
if (!require("here", quietly = TRUE)) {
  stop("Install 'here' package: install.packages('here')")
}

# PROJECT_ROOT is set by the wrapper from CFG_PROJECT_ROOT. The here::here
# call is a legacy fallback for interactive reruns outside the wrapper and
# is not exercised by the canonical pipeline.
PROJECT_ROOT <- here::here("project", "01_Preprocessing")

cat("Project root:", PROJECT_ROOT, "\n")

# Verify it's the right directory (should have config/ and outputs/)
if (!dir.exists(file.path(PROJECT_ROOT, "config"))) {
  stop("PROJECT_ROOT doesn't have config/ directory. Current: ", PROJECT_ROOT)
}

cat("Project root verified\n")
```

# Load Configuration Files

```{r load_configs}
# Load module configurations
config_path <- file.path(PROJECT_ROOT, "config/module_configs.yaml")
if (!file.exists(config_path)) {
  stop("Config file not found: ", config_path)
}
config <- yaml::read_yaml(config_path)

# Get Step 13 config (uses Step 12 resolutions)
step_13_config <- config$modules$step_13_filtered_integration_preview

if (is.null(step_13_config)) {
  stop("step_13_filtered_integration_preview not found in config$modules")
}

# Get Step 12 config for clustering resolutions
step_12_config <- config$modules$step_12_filtered_scvi_integration

if (is.null(step_12_config)) {
  stop("step_12_filtered_scvi_integration not found in config$modules")
}

cat("Loaded module_configs.yaml\n")

# Load preprocessing manifest
manifest_path <- file.path(PROJECT_ROOT, "config/preprocessing_manifest.yaml")
if (!file.exists(manifest_path)) {
  stop("Preprocessing manifest not found: ", manifest_path)
}
manifest <- yaml::read_yaml(manifest_path)

cat("Loaded preprocessing_manifest.yaml\n")

# Load raw data manifest (up one level from preprocessing)
raw_manifest_path <- file.path(dirname(PROJECT_ROOT), "config/raw_data_manifest.yaml")
if (!file.exists(raw_manifest_path)) {
  stop("Raw data manifest not found: ", raw_manifest_path)
}
raw_manifest <- yaml::read_yaml(raw_manifest_path)

cat("Loaded raw_data_manifest.yaml\n")
```

# Extract Step 12 Outputs

```{r extract_step12}
# CRITICAL DIFFERENCE FROM STEP 09:
# Using Step 12 (filtered scVI) outputs instead of Step 08 (unfiltered)

cat("\n============================================================================\n")
cat("FILTERED INTEGRATION MODE - Step 12 Outputs\n")
cat("============================================================================\n")
cat("Filter: FOUR-WAY (Step 01 + Step 02 + Step 03 + Step 10b)\n")
cat("============================================================================\n\n")

# Step 12 output base
STEP_12_BASE <- file.path(PROJECT_ROOT, "outputs/12_FilteredScviIntegration")

# Validate Step 12 outputs exist
latent_path <- file.path(STEP_12_BASE, "embeddings/latent_50d.csv")
umap_path <- file.path(STEP_12_BASE, "embeddings/umap_2d.csv")
metadata_path <- file.path(STEP_12_BASE, "metadata/cell_metadata_integrated.csv")

# Validate files exist
if (!file.exists(latent_path)) {
  stop("Latent embeddings not found: ", latent_path)
}
if (!file.exists(umap_path)) {
  stop("UMAP coordinates not found: ", umap_path)
}
if (!file.exists(metadata_path)) {
  stop("Cell metadata not found: ", metadata_path)
}

cat("Step 12 file paths validated\n")

# Get default resolution from config
default_resolution <- step_13_config$algorithm_params$default_resolution
cat("Default resolution:", default_resolution, "\n")
```

# Load Step 12 Embeddings

```{r load_embeddings}
cat("\n=== Loading Step 12 (Filtered) Embeddings ===\n\n")

# Load latent embeddings
latent_embedding <- read.csv(latent_path, row.names = 1, check.names = FALSE)
cat("  Latent:", nrow(latent_embedding), "cells x", ncol(latent_embedding), "dims\n")

# Load UMAP coordinates
umap_embedding <- read.csv(umap_path, row.names = 1, check.names = FALSE)
cat("  UMAP:", nrow(umap_embedding), "cells x", ncol(umap_embedding), "dims\n")

# Validate alignment
if (!identical(rownames(latent_embedding), rownames(umap_embedding))) {
  stop("Cell IDs don't match between latent and UMAP!")
}

cat("\nEmbeddings loaded\n")
```

# Define FOUR-WAY Passing Cells

```{r qc_cells}
# All cells in Step 12 embeddings already passed FOUR-WAY QC filter
# (Step 01 UMI + Step 02 MAD + Step 03 Doublet + Step 10b Manifold)
passing_cells <- rownames(latent_embedding)

cat("FOUR-WAY passing cells:", length(passing_cells), "\n")
cat("Filter components:\n")
cat("  - step_01_umi_pass\n")
cat("  - step_02_mad_pass\n")
cat("  - step_03_doublet_pass\n")
cat("  - step_10b_manifold_pass (artifact removal)\n")
```

# Load Step 12 Integrated Metadata

```{r load_step12_metadata}
cat("\n=== Loading Step 12 Cell Metadata ===\n\n")

# Load Step 12 metadata (contains sample_id for each cell)
step_12_cell_metadata <- read.csv(metadata_path, stringsAsFactors = FALSE)
step_12_cell_metadata$cell_id <- step_12_cell_metadata$X

cat("  Loaded Step 12 metadata:", nrow(step_12_cell_metadata), "cells\n")
cat("  Columns:", paste(head(colnames(step_12_cell_metadata), 10), collapse = ", "), "...\n")

# Extract sample list
samples_needed <- unique(step_12_cell_metadata$sample_id)
cat("\n  Samples detected:", length(samples_needed), "\n")

cat("\nStep 12 metadata loaded\n")
```

# Extract Cluster Column Names

```{r define_clustering_columns}
cat("\n=== Extracting Clustering Column Names ===\n\n")

# Extract clustering columns from Step 12 metadata
clustering_columns <- grep("^leiden_scvi_", colnames(step_12_cell_metadata), value = TRUE)

cat("  Found", length(clustering_columns), "clustering columns:\n")
for (col in clustering_columns) {
  cat("    -", col, "\n")
}

# Define default cluster column using leiden_scvi_* format
default_cluster_col <- sprintf("leiden_scvi_%.1f", default_resolution)

# Verify it exists
if (!(default_cluster_col %in% clustering_columns)) {
  stop("Default cluster column '", default_cluster_col, "' not found!\n",
       "  Available: ", paste(clustering_columns, collapse = ", "))
}

cat("\nDefault cluster column:", default_cluster_col, "\n")
```

# Load Step 04 Label Transfer

```{r load_label_metadata}
cat("\n=== Loading Step 04 Cell Type Labels ===\n\n")

# Load label metadata per sample
label_list <- list()

for (sample_id in samples_needed) {
  label_file <- file.path(PROJECT_ROOT, "outputs/04_LabelTransfer/cell_metadata",
                          paste0(sample_id, "_labels.csv"))

  if (!file.exists(label_file)) {
    warning("Label file not found for ", sample_id, ": ", label_file)
    next
  }

  # Load label data
  label_data <- read.csv(label_file, stringsAsFactors = FALSE)

  # First column is ALREADY PREFIXED cell_id
  cell_id_col <- colnames(label_data)[1]
  label_data$cell_id <- label_data[[cell_id_col]]

  # Select only HBCATransferredLabels.Kumar_2023 (reconciled labels)
  label_list[[sample_id]] <- label_data %>%
    select(cell_id, HBCATransferredLabels.Kumar_2023)

  cat("  Loaded", nrow(label_data), "cells from", sample_id, "\n")
}

# Combine all label data
label_metadata <- bind_rows(label_list)
cat("\nTotal label metadata:", nrow(label_metadata), "cells\n")
```

# Load Step 03 Doublet Scores

```{r load_step03_doublets}
cat("\n=== Loading Step 03 Doublet Scores ===\n\n")

# Load doublet scores per sample
doublet_list <- list()

for (sample_id in samples_needed) {
  doublet_file <- file.path(PROJECT_ROOT, "outputs/03_DoubletDetection/metadata",
                            paste0(sample_id, ".csv"))

  if (!file.exists(doublet_file)) {
    warning("Doublet file not found for ", sample_id, ": ", doublet_file)
    next
  }

  # Load doublet data
  doublet_data <- read.csv(doublet_file, stringsAsFactors = FALSE)

  # First column is cell_barcode (already prefixed)
  cell_id_col <- colnames(doublet_data)[1]
  doublet_data$cell_id <- doublet_data[[cell_id_col]]

  # Select only doublet score
  doublet_list[[sample_id]] <- doublet_data %>%
    select(cell_id, scDblFinder_score, doublet_filter_pass)

  cat("  Loaded", nrow(doublet_data), "cells from", sample_id, "\n")
}

# Combine all doublet data
step_03_doublets <- bind_rows(doublet_list)
cat("\nTotal doublet scores:", nrow(step_03_doublets), "cells\n")
```

# Map Anatomical Positions

```{r map_anatomical_positions}
cat("\n=== Mapping Anatomical Positions ===\n\n")

# Build position mapping from raw_data_manifest
position_map <- list()

for (patient_name in names(raw_manifest$patients)) {
  patient <- raw_manifest$patients[[patient_name]]

  for (pos_name in names(patient$positions)) {
    pos <- patient$positions[[pos_name]]
    anatomical <- pos$anatomical

    if (is.null(anatomical)) {
      warning("No anatomical data for ", patient_name, "/", pos_name)
      next
    }

    midline <- anatomical$midline_position

    vertical <- if (length(anatomical$vertical_position) > 1) {
      "Upper+Lower"
    } else {
      anatomical$vertical_position
    }

    horizontal <- if (length(anatomical$horizontal_position) > 1) {
      "Inner+Outer"
    } else {
      anatomical$horizontal_position
    }

    sample_id <- paste0(patient_name, "_", pos_name)
    position_map[[sample_id]] <- list(
      midline_position = midline,
      vertical_position = vertical,
      horizontal_position = horizontal
    )
  }
}

# Convert to dataframe for merging
position_df <- bind_rows(lapply(names(position_map), function(sid) {
  data.frame(
    sample_id = sid,
    midline_position = position_map[[sid]]$midline_position,
    vertical_position = position_map[[sid]]$vertical_position,
    horizontal_position = position_map[[sid]]$horizontal_position,
    stringsAsFactors = FALSE
  )
}))

cat("Mapped", nrow(position_df), "positions\n")
```

# Assemble Unified Metadata

```{r assemble_unified_metadata}
cat("\n=== Assembling Unified Metadata ===\n\n")

# Start with Step 12 integrated metadata (already loaded)
unified_metadata <- step_12_cell_metadata

cat("Starting with Step 12 metadata:", nrow(unified_metadata), "cells\n")

# Merge Step 04 cell type labels
unified_metadata <- unified_metadata %>%
  left_join(label_metadata, by = "cell_id")

cat("After label merge:", nrow(unified_metadata), "cells\n")
cat("  Cells with cell type labels:", sum(!is.na(unified_metadata$HBCATransferredLabels.Kumar_2023)), "\n")

# Merge Step 03 doublet scores
unified_metadata <- unified_metadata %>%
  left_join(step_03_doublets, by = "cell_id")

cat("After doublet score merge:", nrow(unified_metadata), "cells\n")

# Merge anatomical position data
unified_metadata <- unified_metadata %>%
  left_join(position_df, by = "sample_id")

cat("After position merge:", nrow(unified_metadata), "cells\n")

# Save unified metadata
metadata_output <- file.path(PROJECT_ROOT, "outputs/13_FilteredIntegrationPreview/metadata")
dir.create(metadata_output, recursive = TRUE, showWarnings = FALSE)
write.csv(unified_metadata,
          file.path(metadata_output, "unified_metadata.csv"),
          row.names = FALSE)

cat("\nUnified metadata saved\n")
```

# Load Raw Counts Per Sample

```{r load_counts}
cat("\n=== Loading Raw Counts Per Sample ===\n\n")

# Initialize
seurat_list <- list()
manifest_metadata <- data.frame(
  sample_id = character(),
  patient_id = character(),
  position = character(),
  stringsAsFactors = FALSE
)

# Define patients to process
patients <- c("Pat1", "Pat2", "UCI604", "UCI220228")

for (patient_id in patients) {
  patient_data <- raw_manifest$patients[[patient_id]]

  if (is.null(patient_data)) {
    warning("Patient ", patient_id, " not found in raw manifest")
    next
  }

  positions <- names(patient_data$positions)

  for (pos_name in positions) {
    pos_data <- patient_data$positions[[pos_name]]

    if (is.null(pos_data$chromium)) {
      next
    }

    sample_id <- pos_data$chromium$sample_id
    h5_path <- pos_data$chromium$filtered_data_path

    if (is.null(sample_id) || is.null(h5_path)) {
      warning("Missing sample_id or h5_path for ", patient_id, "/", pos_name)
      next
    }

    if (!file.exists(h5_path)) {
      warning("H5 file not found: ", h5_path)
      next
    }

    # Load raw counts
    counts_raw <- Seurat::Read10X_h5(h5_path)

    # Add sample_id prefix to match Step 12 format
    colnames(counts_raw) <- paste0(sample_id, "_", colnames(counts_raw))

    # Filter to FOUR-WAY passing cells
    sample_passing_cells <- passing_cells[grepl(paste0("^", sample_id, "_"), passing_cells)]
    cells_to_keep <- intersect(sample_passing_cells, colnames(counts_raw))

    if (length(cells_to_keep) == 0) {
      warning("No passing cells found for ", sample_id)
      next
    }

    # Subset counts
    counts_filtered <- counts_raw[, cells_to_keep]

    # Create Seurat object
    seurat_obj <- CreateSeuratObject(
      counts = counts_filtered,
      project = sample_id,
      min.cells = 0,
      min.features = 0
    )

    seurat_list[[sample_id]] <- seurat_obj

    manifest_metadata <- rbind(manifest_metadata, data.frame(
      sample_id = sample_id,
      patient_id = patient_id,
      position = pos_name,
      stringsAsFactors = FALSE
    ))
  }
}

cat("\nLoaded", length(seurat_list), "samples\n")
```

# Merge Seurat Objects

```{r merge_objects}
cat("\n=== Merging Seurat Objects ===\n\n")

if (length(seurat_list) == 0) {
  stop("No Seurat objects to merge!")
}

first_obj <- seurat_list[[1]]

if (length(seurat_list) > 1) {
  remaining_objs <- seurat_list[2:length(seurat_list)]

  seurat_merged <- merge(
    x = first_obj,
    y = remaining_objs,
    add.cell.ids = NULL,
    project = "filtered_integrated"
  )
} else {
  seurat_merged <- first_obj
}

cat("Merged:", ncol(seurat_merged), "cells x", nrow(seurat_merged), "genes\n")
```

# Add Patient/Position Metadata

```{r add_metadata}
cat("\n=== Adding Patient/Position Metadata ===\n\n")

extract_sample_id <- function(cell_names) {
  gsub("_[ACGT]+-[0-9]+$", "", cell_names)
}

cell_sample_ids <- extract_sample_id(colnames(seurat_merged))

seurat_merged$patient_id <- manifest_metadata$patient_id[
  match(cell_sample_ids, manifest_metadata$sample_id)
]

seurat_merged$position <- manifest_metadata$position[
  match(cell_sample_ids, manifest_metadata$sample_id)
]

cat("Metadata added\n")
```

# Add Unified Metadata to Seurat Object

```{r add_unified_metadata}
cat("\n=== Adding Unified Metadata to Seurat Object ===\n\n")

seurat_cells <- colnames(seurat_merged)
metadata_cells <- unified_metadata$cell_id

matching_cells <- intersect(seurat_cells, metadata_cells)
cat("  Matching cells:", length(matching_cells), "/", length(seurat_cells), "\n")

cols_to_add <- setdiff(colnames(unified_metadata), c("cell_id", "X", "patient_id", "position", "position_id"))

for (col in cols_to_add) {
  seurat_merged@meta.data[[col]] <- unified_metadata[[col]][
    match(seurat_cells, unified_metadata$cell_id)
  ]
}

cat("\nAdded", length(cols_to_add), "metadata columns to Seurat object\n")
```

# Add Step 12 Embeddings

```{r add_embeddings}
cat("\n=== Adding Step 12 (Filtered) Embeddings ===\n\n")

merged_cells <- colnames(seurat_merged)
latent_subset <- latent_embedding[merged_cells, , drop = FALSE]
umap_subset <- umap_embedding[merged_cells, , drop = FALSE]

seurat_merged[["scvi"]] <- CreateDimReducObject(
  embeddings = as.matrix(latent_subset),
  key = "scvi_",
  assay = "RNA"
)

seurat_merged[["umap"]] <- CreateDimReducObject(
  embeddings = as.matrix(umap_subset),
  key = "UMAP_",
  assay = "RNA"
)

cat("Added scvi (", ncol(latent_subset), "D) and umap (2D) reductions\n", sep = "")
```

# Verify Cluster Assignments

```{r verify_clusters}
cat("\n=== Verifying Cluster Assignments ===\n\n")

missing_cols <- setdiff(clustering_columns, colnames(seurat_merged@meta.data))
if (length(missing_cols) > 0) {
  stop("Missing clustering columns in Seurat metadata: ", paste(missing_cols, collapse = ", "))
}

cat("All", length(clustering_columns), "clustering columns present:\n")
for (col in clustering_columns) {
  n_clusters <- length(unique(seurat_merged@meta.data[[col]]))
  cat("  -", col, ":", n_clusters, "clusters\n")
}

cat("\nDefault cluster column verified:", default_cluster_col, "\n")
```

# Join Layers (CRITICAL for Seurat v5)

```{r join_layers}
cat("\n=== Normalizing and Joining Layers ===\n\n")

gc()

cat("Normalizing data...\n")
seurat_merged <- NormalizeData(seurat_merged, verbose = FALSE)
cat("  Normalization complete\n\n")

gc()

cat("Joining layers...\n")
seurat_merged <- JoinLayers(seurat_merged)
cat("  JoinLayers complete\n\n")

gc()

cat("\nLayers joined successfully\n")
```

# Quick Visualization Test

```{r quick_viz}
cat("\n=== Quick Visualization Test ===\n\n")

# UMAP by patient
p1 <- DimPlot(
  seurat_merged,
  reduction = "umap",
  group.by = "patient_id",
  pt.size = 0.1,
  raster = FALSE
) + ggtitle("UMAP by Patient (FOUR-WAY Filtered)")

print(p1)

# UMAP by cluster (default resolution)
p2 <- DimPlot(
  seurat_merged,
  reduction = "umap",
  group.by = default_cluster_col,
  label = TRUE,
  pt.size = 0.1,
  raster = FALSE
) +
  ggtitle(paste("UMAP by", default_cluster_col, "(Filtered)")) +
  NoLegend()

print(p2)

# UMAP by cell type labels
if ("HBCATransferredLabels.Kumar_2023" %in% colnames(seurat_merged@meta.data)) {
  p3 <- DimPlot(
    seurat_merged,
    reduction = "umap",
    group.by = "HBCATransferredLabels.Kumar_2023",
    label = TRUE,
    repel = TRUE,
    pt.size = 0.1,
    alpha = 0.3,
    raster = FALSE
  ) +
    ggtitle("UMAP by Cell Type (HBCA Kumar 2023) - Filtered") +
    theme(legend.position = "right")

  print(p3)
}

cat("\nVisualizations created successfully\n")
```

# QC Metric Overlays

```{r qc_overlays}
cat("\n=== QC Metric Overlays ===\n\n")

q95_counts <- quantile(seurat_merged$nCount_RNA, 0.95, na.rm = TRUE)
q5_features <- quantile(seurat_merged$nFeature_RNA, 0.05, na.rm = TRUE)
q95_features <- quantile(seurat_merged$nFeature_RNA, 0.95, na.rm = TRUE)

# FeaturePlot: nCount_RNA
p_counts <- FeaturePlot(
  seurat_merged,
  features = "nCount_RNA",
  max.cutoff = q95_counts,
  pt.size = 0.2,
  alpha = 0.3,
  raster = FALSE
) +
  scale_color_viridis(option = "viridis", direction = -1) +
  ggtitle("UMAP: UMI Counts per Cell (Filtered)")

print(p_counts)

# FeaturePlot: nFeature_RNA
p_features <- FeaturePlot(
  seurat_merged,
  features = "nFeature_RNA",
  max.cutoff = q95_features,
  min.cutoff = q5_features,
  order = TRUE,
  raster = FALSE
) +
  scale_color_viridis(option = "viridis") +
  ggtitle("UMAP: Genes Detected per Cell (Filtered)")

print(p_features)

cat("\nQC metric overlays complete\n")
```

# Multi-Resolution Marker Analysis

```{r multi_res_markers, fig.width=14, fig.height=10}
cat("\n=== Multi-Resolution Marker Analysis ===\n\n")

cat("Running marker analysis across", length(clustering_columns), "resolutions...\n\n")

gc()

all_resolution_markers <- list()

for (i in seq_along(clustering_columns)) {
  res_col <- clustering_columns[i]

  cat("---\n")
  cat("Resolution", i, "of", length(clustering_columns), ":", res_col, "\n")

  Idents(seurat_merged) <- seurat_merged@meta.data[[res_col]]

  n_clusters <- length(unique(Idents(seurat_merged)))
  cat("  Clusters:", n_clusters, "\n")

  cat("  Running presto::wilcoxauc()...\n")
  markers <- presto::wilcoxauc(seurat_merged, assay = "data")

  cat("  Found", nrow(markers), "marker tests\n")

  sig_markers <- markers %>%
    filter(padj < 0.05, logFC > 0) %>%
    arrange(group, desc(logFC))

  cat("  Significant markers (padj < 0.05, logFC > 0):", nrow(sig_markers), "\n")

  top_markers <- sig_markers %>%
    group_by(group) %>%
    slice_max(order_by = logFC, n = 5) %>%
    ungroup()

  all_resolution_markers[[res_col]] <- markers

  gc()

  # Generate dotplot
  if (nrow(top_markers) > 0) {
    tryCatch({
      p_dotplot <- DotPlot(
        seurat_merged,
        features = unique(top_markers$feature),
        group.by = res_col
      ) +
        coord_flip() +
        scale_color_viridis(option = "viridis") +
        ggtitle(paste("Top Markers -", res_col, "(Filtered)")) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
              axis.text.y = element_text(size = 6))

      print(p_dotplot)
    }, error = function(e) {
      cat("  Dotplot generation failed:", e$message, "\n")
    })
  }

  cat("\n")
}

cat("Multi-resolution marker analysis complete\n")
```

# Save Per-Resolution Markers

```{r save_markers}
cat("\n=== Saving Per-Resolution Markers ===\n\n")

if (length(all_resolution_markers) == 0) {
  cat("No markers to save - skipping\n")
} else {
  output_dir <- file.path(PROJECT_ROOT, "outputs/13_FilteredIntegrationPreview/markers_by_resolution")
  dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

  for (res_col in names(all_resolution_markers)) {
    res_value <- gsub("leiden_scvi_([0-9.]+)$", "\\1", res_col)
    filename <- paste0("markers_res_", res_value, ".csv")
    filepath <- file.path(output_dir, filename)

    tryCatch({
      write.csv(all_resolution_markers[[res_col]], file = filepath, row.names = FALSE)
      cat("  ", filename, "-", nrow(all_resolution_markers[[res_col]]), "rows\n")
    }, error = function(e) {
      cat("  Failed to save", filename, ":", e$message, "\n")
    })
  }

  # Save combined file
  combined_markers <- bind_rows(
    lapply(names(all_resolution_markers), function(res_col) {
      res_value <- gsub(".*res_([0-9.]+)$", "\\1", res_col)
      all_resolution_markers[[res_col]] %>%
        mutate(resolution = res_value, .before = 1)
    })
  )

  combined_filepath <- file.path(output_dir, "all_markers_all_resolutions.csv")
  write.csv(combined_markers, file = combined_filepath, row.names = FALSE)
  cat("  all_markers_all_resolutions.csv -", nrow(combined_markers), "rows\n")

  cat("\nMarker files saved successfully\n")
}
```

# Save Seurat Object

```{r save_seurat}
cat("\n=== Saving Seurat Object ===\n\n")

seurat_output <- file.path(PROJECT_ROOT, "outputs/13_FilteredIntegrationPreview/seurat_objects")
dir.create(seurat_output, recursive = TRUE, showWarnings = FALSE)

seurat_path <- file.path(seurat_output, "integrated_seurat.rds")
saveRDS(seurat_merged, file = seurat_path)

cat("Seurat object saved\n")
cat("  Path:", seurat_path, "\n")
cat("  Size:", format(file.size(seurat_path) / 1024^3, digits = 2), "GB\n")
cat("  Cells:", ncol(seurat_merged), "\n")
cat("  Genes:", nrow(seurat_merged), "\n")
```

# Summary

```{r summary}
cat("\n=== FILTERED INTEGRATION PREVIEW SUMMARY ===\n\n")

cat("Filter Applied: FOUR-WAY\n")
cat("  - step_01_umi_pass\n")
cat("  - step_02_mad_pass\n")
cat("  - step_03_doublet_pass\n")
cat("  - step_10b_manifold_pass (artifact removal)\n\n")

cat("Seurat Object:\n")
cat("  Cells:", ncol(seurat_merged), "\n")
cat("  Genes:", nrow(seurat_merged), "\n")
cat("  Patients:", length(unique(seurat_merged$patient_id)), "\n")
cat("  Samples:", length(unique(seurat_merged$orig.ident)), "\n")

cat("\nDimensional Reductions:\n")
print(names(seurat_merged@reductions))

cat("\nCluster Columns:\n")
cat(" ", paste(clustering_columns, collapse = ", "), "\n")

cat("\nAll chunks executed successfully!\n")
cat("\nStep 13 improvements over Step 09:\n")
cat("  - Uses FOUR-WAY filtered cells (artifact-free)\n")
cat("  - Loads from Step 12 filtered scVI integration\n")
cat("  - Reduced cell count after Step 10b artifact removal\n")
```
