---
title: "Compartment Integration Report - `r params$compartment` (`r params$config_label`)"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
date: "`r Sys.Date()`"
params:
  project_root: ""
  compartment: ""
  config_label: ""
  upstream_root: ""
  intermediate_dir: ""
  outdir: ""
  dry_run: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 10,
  fig.height = 8
)

suppressPackageStartupMessages({
  library(Seurat)
  library(ggplot2)
  library(dplyr)
  library(presto)
  library(patchwork)
  library(viridis)
  library(Matrix)
  library(jsonlite)
})

# Coerce a leiden column to a factor with numerically-sorted levels. Without
# this, Seurat sees character labels and orders clusters alphabetically
# ("0","1","10","11","2","20",...), which scrambles every downstream heatmap
# column and marker grouping. Falls back to lexical sort if the labels aren't
# all numeric.
order_cluster_factor <- function(x) {
  lvls <- unique(as.character(x))
  num <- suppressWarnings(as.numeric(lvls))
  if (all(!is.na(num))) {
    lvls <- lvls[order(num)]
  } else {
    lvls <- sort(lvls)
  }
  factor(as.character(x), levels = lvls)
}

cat("Libraries loaded\n")
```

# Parameters and Paths

```{r params_echo}
PROJECT_ROOT     <- params$project_root
COMPARTMENT      <- params$compartment
CONFIG_LABEL     <- params$config_label
UPSTREAM_ROOT    <- params$upstream_root
INTERMEDIATE_DIR <- params$intermediate_dir
OUTDIR           <- params$outdir
DRY_RUN          <- isTRUE(params$dry_run)

# Required params must be supplied by the rendering wrapper. Fail loudly
# if any are empty rather than silently substituting a default path.
for (.p in c("project_root", "compartment", "config_label",
             "upstream_root", "intermediate_dir", "outdir")) {
  if (!nzchar(params[[.p]])) {
    stop(sprintf("FATAL: required param '%s' is empty. Render via run/run_compartment_integration_report.sh.", .p))
  }
}

# Intermediate directory written by 08a_export_h5ad_intermediate.py
COUNTS_MTX        <- file.path(INTERMEDIATE_DIR, "counts.mtx.gz")
CELLS_TSV         <- file.path(INTERMEDIATE_DIR, "cells.tsv")
GENES_TSV         <- file.path(INTERMEDIATE_DIR, "genes.tsv")
OBS_CSV           <- file.path(INTERMEDIATE_DIR, "obs.csv")
LATENT_CSV        <- file.path(INTERMEDIATE_DIR, "latent.csv")
UMAP_CSV          <- file.path(INTERMEDIATE_DIR, "umap.csv")
MANIFEST_JSON     <- file.path(INTERMEDIATE_DIR, "manifest.json")

# Assembly cell metadata (whole object; filtered to compartment cells downstream)
CELL_METADATA_CSV <- file.path(UPSTREAM_ROOT, "assembly_step15", "cell_metadata.csv")

# Contamination / cluster_annotation source of truth
CONTAMINATION_CSV <- file.path(UPSTREAM_ROOT, "contamination", "cell_compartments.csv")

cat("============================================================================\n")
cat("Compartment Integration Report\n")
cat("============================================================================\n")
cat("Compartment:      ", COMPARTMENT, "\n")
cat("Config label:     ", CONFIG_LABEL, "\n")
cat("Intermediate dir: ", INTERMEDIATE_DIR, "\n")
cat("Cell metadata:    ", CELL_METADATA_CSV, "\n")
cat("Contamination:    ", CONTAMINATION_CSV, "\n")
cat("Output directory: ", OUTDIR, "\n")
cat("Dry run:          ", DRY_RUN, "\n")
cat("============================================================================\n")
```

# Validation Preamble

Checks that the compartment integration h5ad, assembly metadata, contamination
CSV, and the compartment cell list are all present. If `dry_run` is true, the
report exits here — used for cheap path validation before submitting the
full-scale job.

```{r validate_inputs}
required_files <- list(
  counts_mtx      = COUNTS_MTX,
  cells_tsv       = CELLS_TSV,
  genes_tsv       = GENES_TSV,
  obs_csv         = OBS_CSV,
  latent_csv      = LATENT_CSV,
  umap_csv        = UMAP_CSV,
  manifest_json   = MANIFEST_JSON,
  cell_metadata   = CELL_METADATA_CSV,
  contamination   = CONTAMINATION_CSV
)

missing <- character(0)
for (nm in names(required_files)) {
  f <- required_files[[nm]]
  ok <- file.exists(f)
  cat(sprintf("  [%s] %-15s %s\n", if (ok) "OK " else "MISS", nm, f))
  if (!ok) missing <- c(missing, nm)
}

if (length(missing) > 0) {
  stop("Missing required input files: ", paste(missing, collapse = ", "))
}

if (DRY_RUN) {
  cat("\nVALIDATION PASSED (dry run) - exiting before heavy computation\n")
  knitr::knit_exit()
}
```

# Load Integration Intermediate

The compartment integration h5ad is exported to a flat intermediate directory
by `scripts/08a_export_h5ad_intermediate.py` (run in the scgpt_gpu container).
This Rmd then reads MTX + CSV directly — no reticulate, no Python.

Inputs (under `INTERMEDIATE_DIR`):
- `counts.mtx.gz` — sparse counts (cells x genes; transposed below for Seurat)
- `cells.tsv`, `genes.tsv` — IDs
- `obs.csv` — per-cell obs frame from h5ad
- `latent.csv`, `umap.csv` — embeddings (cell_id keyed)
- `manifest.json` — source path, dimensions, latent/umap key names

```{r load_intermediate}
cat("\n=== Loading integration intermediate ===\n\n")

manifest <- jsonlite::fromJSON(MANIFEST_JSON)
cat("Manifest:\n")
cat("  source_h5ad:", manifest$source_h5ad, "\n")
cat("  shape:", manifest$n_cells, "cells x", manifest$n_genes, "genes\n")
cat("  latent_key:", manifest$latent_key, "\n")
cat("  umap_key:  ", manifest$umap_key, "\n")

cell_ids <- readLines(CELLS_TSV)
gene_ids <- readLines(GENES_TSV)
cat("Loaded", length(cell_ids), "cell IDs and", length(gene_ids), "gene IDs\n")

# Counts: MTX is cells x genes; Seurat wants genes x cells.
counts_mat <- Matrix::readMM(gzfile(COUNTS_MTX))
counts_mat <- as(counts_mat, "CsparseMatrix")
counts_mat <- Matrix::t(counts_mat)
rownames(counts_mat) <- gene_ids
colnames(counts_mat) <- cell_ids
cat("Counts matrix:", nrow(counts_mat), "genes x", ncol(counts_mat), "cells\n")

# obs
obs_df <- read.csv(OBS_CSV, stringsAsFactors = FALSE, check.names = FALSE)
rownames(obs_df) <- obs_df$cell_id
obs_df$cell_id <- NULL
stopifnot(all(rownames(obs_df) == cell_ids))
cat("obs columns (", ncol(obs_df), "):\n  ",
    paste(colnames(obs_df), collapse = ", "), "\n", sep = "")

# Latent embedding
latent_df <- read.csv(LATENT_CSV, stringsAsFactors = FALSE, check.names = FALSE)
stopifnot(all(latent_df$cell_id == cell_ids))
latent_mat <- as.matrix(latent_df[, setdiff(colnames(latent_df), "cell_id"), drop = FALSE])
rownames(latent_mat) <- cell_ids
cat("Latent (", manifest$latent_key, "):", ncol(latent_mat), "dims\n")

# UMAP embedding
umap_df <- read.csv(UMAP_CSV, stringsAsFactors = FALSE, check.names = FALSE)
stopifnot(all(umap_df$cell_id == cell_ids))
umap_mat <- as.matrix(umap_df[, setdiff(colnames(umap_df), "cell_id"), drop = FALSE])
rownames(umap_mat) <- cell_ids
cat("UMAP (", manifest$umap_key, "):", ncol(umap_mat), "dims\n")

# Canonical compartment cell IDs used throughout the report
compartment_cells <- cell_ids
gc()
```

# Load Assembly Cell Metadata

Loads the whole-assembly `cell_metadata.csv` and filters to the compartment's
cells only. This becomes the source for patient_id / sample_id / position and
(if present) pre-computed QC metrics.

```{r load_cell_metadata}
cat("\n=== Loading assembly cell metadata ===\n\n")

cell_meta <- read.csv(CELL_METADATA_CSV, stringsAsFactors = FALSE)

cell_id_col <- intersect(c("cell_id", "X", "barcode"), colnames(cell_meta))[1]
if (is.na(cell_id_col)) cell_id_col <- colnames(cell_meta)[1]
rownames(cell_meta) <- cell_meta[[cell_id_col]]
cat("Whole cell_metadata.csv:", nrow(cell_meta), "rows x", ncol(cell_meta), "cols\n")

# Filter to compartment cells only - critical: QC overlays downstream should
# reflect ONLY these cells, not the whole assembly
cell_meta_comp <- cell_meta[intersect(compartment_cells, rownames(cell_meta)), , drop = FALSE]
cat("Filtered to compartment cells:", nrow(cell_meta_comp), "/", length(compartment_cells), "\n")
cat("Columns:", paste(colnames(cell_meta_comp), collapse = ", "), "\n")
```

# Load Contamination / cluster_annotation

```{r load_contamination}
cat("\n=== Loading contamination CSV ===\n\n")

contam <- read.csv(CONTAMINATION_CSV, stringsAsFactors = FALSE)
contam_id_col <- intersect(c("cell_id", "X", "barcode"), colnames(contam))[1]
if (is.na(contam_id_col)) contam_id_col <- colnames(contam)[1]
rownames(contam) <- contam[[contam_id_col]]
cat("Whole contamination CSV:", nrow(contam), "rows\n")
cat("Columns:", paste(colnames(contam), collapse = ", "), "\n")

contam_comp <- contam[intersect(compartment_cells, rownames(contam)), , drop = FALSE]
cat("Filtered to compartment cells:", nrow(contam_comp), "\n")

expected_contam_cols <- c("cluster_annotation", "compartment", "leiden_1.0")
have    <- intersect(expected_contam_cols, colnames(contam_comp))
missing_contam <- setdiff(expected_contam_cols, colnames(contam_comp))
cat("Have:", paste(have, collapse = ", "), "\n")
if (length(missing_contam) > 0) {
  cat("NOTE: missing (non-fatal):", paste(missing_contam, collapse = ", "), "\n")
}

if ("compartment" %in% colnames(contam_comp)) {
  n_mismatch <- sum(contam_comp$compartment != COMPARTMENT, na.rm = TRUE)
  if (n_mismatch > 0) {
    cat("WARNING:", n_mismatch, "cells have compartment != ", COMPARTMENT, "\n")
  }
}
```

# Load Kumar Consensus Labels

```{r load_kumar}
cat("\n=== Loading per-sample Kumar labels ===\n\n")

# Derive samples from compartment-filtered cell_metadata. The
# compartments_step15/cell_lists/{Compartment}.csv file is a cell-id list
# (not a sample inventory), so we cannot use it directly.
if (!"sample_id" %in% colnames(cell_meta_comp)) {
  stop("cell_meta_comp lacks sample_id column - cannot derive sample list")
}
samples_in_compartment <- sort(unique(cell_meta_comp$sample_id))
cat("Samples in compartment:", length(samples_in_compartment), "\n")

kumar_list   <- list()
missing_kumar <- character(0)

for (sid in samples_in_compartment) {
  kfile <- file.path(UPSTREAM_ROOT, sid, "kumar_labels.csv")
  if (!file.exists(kfile)) {
    missing_kumar <- c(missing_kumar, sid)
    next
  }
  kdf <- read.csv(kfile, stringsAsFactors = FALSE)
  kid_col <- intersect(c("cell_id", "X", "barcode"), colnames(kdf))[1]
  if (is.na(kid_col)) kid_col <- colnames(kdf)[1]
  kdf$cell_id <- kdf[[kid_col]]

  if ("consensus_label" %in% colnames(kdf)) {
    kumar_list[[sid]] <- kdf[, c("cell_id", "consensus_label"), drop = FALSE]
  } else {
    cat("  ", sid, ": no consensus_label column\n")
  }
}

if (length(missing_kumar) > 0) {
  cat("Missing kumar_labels.csv for", length(missing_kumar), "samples:\n  ",
      paste(head(missing_kumar, 10), collapse = ", "),
      if (length(missing_kumar) > 10) " ..." else "", "\n")
}

if (length(kumar_list) > 0) {
  kumar_df <- bind_rows(kumar_list)
  cat("Kumar rows loaded:", nrow(kumar_df), "- unique labels:",
      length(unique(na.omit(kumar_df$consensus_label))), "\n")
} else {
  kumar_df <- NULL
  cat("WARNING: no Kumar labels loaded\n")
}
```

# Load Per-Sample scQC Metrics

Loads `doublet_score` (from `scqc/doublet_scores.csv`) and `smooth_r2` (from
`scqc/discordance_scores.csv`) per sample. These are bare-barcode keyed, so
they get sample-prefixed to match the assembly cell IDs.

```{r load_scqc}
cat("\n=== Loading per-sample scQC metrics ===\n\n")

scqc_list <- list()
missing_scqc <- character(0)

for (sid in samples_in_compartment) {
  doub_file <- file.path(UPSTREAM_ROOT, sid, "scqc", "doublet_scores.csv")
  disc_file <- file.path(UPSTREAM_ROOT, sid, "scqc", "discordance_scores.csv")

  if (!file.exists(doub_file) || !file.exists(disc_file)) {
    missing_scqc <- c(missing_scqc, sid)
    next
  }

  d <- read.csv(doub_file, stringsAsFactors = FALSE)
  s <- read.csv(disc_file, stringsAsFactors = FALSE)

  # Sample-prefix the bare barcodes to match assembly cell IDs
  d$cell_id <- paste0(sid, "_", d$cell_id)
  s$cell_id <- paste0(sid, "_", s$cell_id)

  m <- merge(
    d[, c("cell_id", "doublet_score"), drop = FALSE],
    s[, c("cell_id", "smooth_r2"), drop = FALSE],
    by = "cell_id", all = TRUE
  )
  scqc_list[[sid]] <- m
}

if (length(missing_scqc) > 0) {
  cat("Missing scQC files for", length(missing_scqc), "samples:\n  ",
      paste(head(missing_scqc, 10), collapse = ", "),
      if (length(missing_scqc) > 10) " ..." else "", "\n")
}

if (length(scqc_list) > 0) {
  scqc_df <- bind_rows(scqc_list)
  cat("scQC rows loaded:", nrow(scqc_df),
      "- doublet_score range [", round(range(scqc_df$doublet_score, na.rm = TRUE), 3),
      "], smooth_r2 range [",  round(range(scqc_df$smooth_r2, na.rm = TRUE), 3), "]\n")
} else {
  scqc_df <- NULL
  cat("WARNING: no scQC metrics loaded\n")
}
```

# Load Cluster/Metadata Patches

Optional sidecar CSVs at `${INTERMEDIATE_DIR}/patches/*.csv`. Each must have
`cell_id` as the first column; remaining columns are merged into obs at build
time. This is the lightweight extension point — re-clustering jobs
(`06c_recluster_integration.py`) and other derived-metadata patches drop their
output here without modifying the canonical h5ad or intermediate pieces.

```{r load_patches}
cat("\n=== Loading patches ===\n\n")

patches_dir <- file.path(INTERMEDIATE_DIR, "patches")
patch_df <- NULL

if (dir.exists(patches_dir)) {
  patch_files <- list.files(patches_dir, pattern = "\\.csv$", full.names = TRUE)
  cat("Patches dir:", patches_dir, "(", length(patch_files), "files)\n")

  for (pf in patch_files) {
    pdf <- read.csv(pf, stringsAsFactors = FALSE, check.names = FALSE)
    if (!"cell_id" %in% colnames(pdf)) {
      cat("  SKIP", basename(pf), "- no cell_id column\n")
      next
    }
    new_cols <- setdiff(colnames(pdf), "cell_id")
    cat("  +", basename(pf), ":", length(new_cols), "cols (",
        paste(head(new_cols, 5), collapse = ", "),
        if (length(new_cols) > 5) "..." else "", ")\n")

    if (is.null(patch_df)) {
      patch_df <- pdf
    } else {
      patch_df <- merge(patch_df, pdf, by = "cell_id", all = TRUE)
    }
  }
  if (!is.null(patch_df)) {
    cat("Patches assembled:", nrow(patch_df), "rows x", ncol(patch_df) - 1, "cols\n")
  }
} else {
  cat("No patches directory at", patches_dir, "\n")
}
```

# Build Seurat Object

Constructs a minimal Seurat object from the h5ad counts. Seurat computes
`nCount_RNA` and `nFeature_RNA` at object creation time — because we built
from the compartment-filtered counts matrix, these QC values are intrinsic
to the compartment cells (not inherited from the whole-assembly object).

```{r build_seurat}
cat("\n=== Building Seurat object ===\n\n")

seurat_obj <- CreateSeuratObject(
  counts       = counts_mat,
  project      = paste0("integ_", COMPARTMENT),
  min.cells    = 0,
  min.features = 0
)
cat("Seurat:", ncol(seurat_obj), "cells x", nrow(seurat_obj), "genes\n")

cells <- colnames(seurat_obj)

# Percent mitochondrial - computed on compartment cells
mt_pattern <- "^MT-"
mt_genes <- grep(mt_pattern, rownames(seurat_obj), value = TRUE, ignore.case = TRUE)
if (length(mt_genes) > 0) {
  seurat_obj[["percent_mt"]] <- PercentageFeatureSet(seurat_obj, pattern = mt_pattern)
  cat("Computed percent_mt across", length(mt_genes), "mitochondrial genes\n")
} else {
  cat("No mitochondrial genes matched '", mt_pattern, "'\n", sep = "")
}

# Merge h5ad obs columns (these reflect compartment cells by construction)
obs_aligned <- obs_df[cells, , drop = FALSE]
cols_from_obs <- setdiff(colnames(obs_aligned), colnames(seurat_obj@meta.data))
for (col in cols_from_obs) {
  seurat_obj@meta.data[[col]] <- obs_aligned[[col]]
}
cat("Added", length(cols_from_obs), "columns from h5ad obs\n")

# Assembly metadata (sample_id, patient_id, position, etc.) - already filtered
asm_cols <- setdiff(colnames(cell_meta_comp),
                    c(cell_id_col, colnames(seurat_obj@meta.data)))
for (col in asm_cols) {
  seurat_obj@meta.data[[col]] <- cell_meta_comp[cells, col]
}
cat("Added", length(asm_cols), "columns from assembly cell_metadata\n")

# Contamination columns - force overwrite of the three canonical ones from the
# contamination source of truth
priority_cols <- intersect(c("cluster_annotation", "compartment", "leiden_1.0"),
                           colnames(contam_comp))
other_cols    <- setdiff(colnames(contam_comp),
                         c(contam_id_col, colnames(seurat_obj@meta.data)))
contam_cols   <- unique(c(priority_cols, other_cols))
for (col in contam_cols) {
  seurat_obj@meta.data[[col]] <- contam_comp[cells, col]
}
cat("Added", length(contam_cols), "columns from contamination CSV\n")

# Kumar consensus label
if (!is.null(kumar_df)) {
  m <- match(cells, kumar_df$cell_id)
  seurat_obj$kumar_consensus_label <- kumar_df$consensus_label[m]
  n_matched <- sum(!is.na(seurat_obj$kumar_consensus_label))
  cat("Kumar consensus_label matched:", n_matched, "of", length(cells), "cells\n")
}

# scQC metrics: doublet_score + smooth_r2
if (!is.null(scqc_df)) {
  m <- match(cells, scqc_df$cell_id)
  seurat_obj$doublet_score <- scqc_df$doublet_score[m]
  seurat_obj$smooth_r2     <- scqc_df$smooth_r2[m]
  cat("scQC metrics matched:",
      sum(!is.na(seurat_obj$doublet_score)), "doublet,",
      sum(!is.na(seurat_obj$smooth_r2)),     "smooth_r2 of",
      length(cells), "cells\n")
}

# Patch sidecars (cluster overrides, derived metadata)
if (!is.null(patch_df)) {
  m <- match(cells, patch_df$cell_id)
  patch_cols <- setdiff(colnames(patch_df), "cell_id")
  for (col in patch_cols) {
    seurat_obj@meta.data[[col]] <- patch_df[[col]][m]
  }
  cat("Patch columns merged:", length(patch_cols), "(",
      paste(head(patch_cols, 6), collapse = ", "),
      if (length(patch_cols) > 6) "..." else "", ")\n")
}

# Coerce every leiden-like column to a numerically-ordered factor up front so
# that every downstream consumer (UMAP label order, marker grouping, DoHeatmap
# columns) sees clusters in 0,1,2,...,N order instead of "0","1","10","11","2".
all_leiden_cols <- grep("^leiden", colnames(seurat_obj@meta.data), value = TRUE)
for (lc in all_leiden_cols) {
  seurat_obj@meta.data[[lc]] <- order_cluster_factor(seurat_obj@meta.data[[lc]])
}
if (length(all_leiden_cols) > 0) {
  cat("Ordered", length(all_leiden_cols), "leiden columns:",
      paste(all_leiden_cols, collapse = ", "), "\n")
}

# Identify leiden column for default plotting. Patched columns (from sidecar
# CSVs in patches/) take precedence over the h5ad's native leiden columns —
# the patches are the reason we ran the recluster job in the first place.
leiden_col <- NULL
leiden_candidates <- c(
  "leiden_n30_r1.0", "leiden_n30_r1.5", "leiden_n30_r0.6", "leiden_n30_r2.0",
  "leiden_1.0", "leiden_1_0", "leiden", "clusters"
)
for (cand in leiden_candidates) {
  if (cand %in% colnames(seurat_obj@meta.data)) {
    leiden_col <- cand
    break
  }
}
if (is.null(leiden_col)) {
  cat("NOTE: no leiden column found - default cluster plots will be skipped\n")
} else {
  cat("Leiden column:", leiden_col, "-",
      length(levels(seurat_obj@meta.data[[leiden_col]])), "clusters\n")
}

# Attach latent + UMAP reductions
latent_sub <- latent_mat[cells, , drop = FALSE]
umap_sub   <- umap_mat[cells, , drop = FALSE]

seurat_obj[["scvi"]] <- CreateDimReducObject(
  embeddings = as.matrix(latent_sub),
  key = "scvi_",
  assay = "RNA"
)
seurat_obj[["umap"]] <- CreateDimReducObject(
  embeddings = as.matrix(umap_sub),
  key = "UMAP_",
  assay = "RNA"
)
cat("Added scvi (", ncol(latent_sub), "D) and umap (2D) reductions\n", sep = "")
```

# Normalize and Join Layers

```{r normalize}
cat("\n=== Normalizing and joining layers ===\n\n")

gc()
seurat_obj <- NormalizeData(seurat_obj, verbose = FALSE)
cat("Normalization complete\n")

gc()
seurat_obj <- JoinLayers(seurat_obj)
cat("JoinLayers complete\n")
gc()
```

# Summary Statistics

```{r summary_stats}
cat("\n============================================================================\n")
cat("Summary for", COMPARTMENT, "(", CONFIG_LABEL, ")\n")
cat("============================================================================\n\n")

n_cells_f <- ncol(seurat_obj)
n_genes_f <- nrow(seurat_obj)
cat("Total cells: ", n_cells_f, "\n")
cat("Total genes: ", n_genes_f, "\n")

if ("sample_id" %in% colnames(seurat_obj@meta.data)) {
  cat("\nCells per sample:\n")
  print(sort(table(seurat_obj$sample_id), decreasing = TRUE))
}

if ("patient_id" %in% colnames(seurat_obj@meta.data)) {
  cat("\nCells per patient:\n")
  print(table(seurat_obj$patient_id))
}

if (!is.null(leiden_col)) {
  cat("\nCluster sizes (", leiden_col, "):\n", sep = "")
  print(table(seurat_obj@meta.data[[leiden_col]]))
}

if ("cluster_annotation" %in% colnames(seurat_obj@meta.data)) {
  cat("\ncluster_annotation counts:\n")
  print(sort(table(seurat_obj$cluster_annotation), decreasing = TRUE))
}
```

# UMAP Visualizations {.tabset}

## By Leiden Cluster

```{r umap_leiden, fig.width=14, fig.height=10}
if (!is.null(leiden_col)) {
  Idents(seurat_obj) <- leiden_col
  DimPlot(seurat_obj, reduction = "umap", group.by = leiden_col,
          label = TRUE, pt.size = 0.3, label.size = 4, raster = FALSE) +
    theme(legend.position = "right")
} else {
  cat("No leiden column available\n")
}
```

## By Patient

```{r umap_patient, fig.width=12, fig.height=8}
if ("patient_id" %in% colnames(seurat_obj@meta.data)) {
  DimPlot(seurat_obj, reduction = "umap", group.by = "patient_id",
          pt.size = 0.3, raster = FALSE, shuffle = TRUE)
} else {
  cat("patient_id not available\n")
}
```

## By Sample

```{r umap_sample, fig.width=14, fig.height=10}
if ("sample_id" %in% colnames(seurat_obj@meta.data)) {
  DimPlot(seurat_obj, reduction = "umap", group.by = "sample_id",
          pt.size = 0.2, raster = FALSE) +
    theme(legend.text = element_text(size = 6))
} else {
  cat("sample_id not available\n")
}
```

## By Kumar consensus_label

```{r umap_kumar, fig.width=14, fig.height=10}
if ("kumar_consensus_label" %in% colnames(seurat_obj@meta.data)) {
  DimPlot(seurat_obj, reduction = "umap",
          group.by = "kumar_consensus_label",
          label = TRUE, repel = TRUE,
          pt.size = 0.3, alpha = 0.5, raster = FALSE) +
    NoLegend()
} else {
  cat("Kumar consensus_label not available\n")
}
```

## By cluster_annotation

```{r umap_cluster_annot, fig.width=14, fig.height=10}
if ("cluster_annotation" %in% colnames(seurat_obj@meta.data)) {
  DimPlot(seurat_obj, reduction = "umap",
          group.by = "cluster_annotation",
          label = TRUE, repel = TRUE,
          pt.size = 0.3, alpha = 0.5, raster = FALSE) +
    NoLegend()
} else {
  cat("cluster_annotation not available\n")
}
```

# QC Metric Overlays {.tabset}

QC metrics below are computed from the compartment-filtered counts matrix
(`nCount_RNA`, `nFeature_RNA`, `percent_mt` via `PercentageFeatureSet()`).
Color scales use the 95th percentile of values within this compartment,
so ranges do not leak from the whole-assembly distribution.

## UMI Counts

```{r qc_umi, fig.width=12, fig.height=8}
q95_counts <- quantile(seurat_obj$nCount_RNA, 0.95, na.rm = TRUE)
FeaturePlot(seurat_obj, features = "nCount_RNA",
            max.cutoff = q95_counts, pt.size = 0.2, alpha = 0.5, raster = FALSE) +
  scale_color_viridis(option = "viridis", direction = -1)
```

## Genes Detected

```{r qc_genes, fig.width=12, fig.height=8}
q05_features <- quantile(seurat_obj$nFeature_RNA, 0.05, na.rm = TRUE)
q95_features <- quantile(seurat_obj$nFeature_RNA, 0.95, na.rm = TRUE)
FeaturePlot(seurat_obj, features = "nFeature_RNA",
            min.cutoff = q05_features, max.cutoff = q95_features,
            pt.size = 0.2, alpha = 0.5, raster = FALSE) +
  scale_color_viridis(option = "viridis")
```

## Percent Mitochondrial

```{r qc_mt, fig.width=12, fig.height=8}
if ("percent_mt" %in% colnames(seurat_obj@meta.data)) {
  q95_mt <- quantile(seurat_obj$percent_mt, 0.95, na.rm = TRUE)
  FeaturePlot(seurat_obj, features = "percent_mt",
              max.cutoff = q95_mt,
              pt.size = 0.2, alpha = 0.5, raster = FALSE) +
    scale_color_viridis(option = "magma")
} else {
  cat("percent_mt not computed (no MT- genes in this compartment)\n")
}
```

## Doublet Score

```{r qc_doublet, fig.width=12, fig.height=8}
if ("doublet_score" %in% colnames(seurat_obj@meta.data) &&
    sum(!is.na(seurat_obj$doublet_score)) > 0) {
  q95_d <- quantile(seurat_obj$doublet_score, 0.95, na.rm = TRUE)
  FeaturePlot(seurat_obj, features = "doublet_score",
              max.cutoff = q95_d,
              pt.size = 0.2, alpha = 0.5, raster = FALSE,
              order = TRUE) +
    scale_color_viridis(option = "magma")
} else {
  cat("doublet_score not loaded (check scqc/doublet_scores.csv per sample)\n")
}
```

## Smooth R² (BigSur)

```{r qc_smooth_r2, fig.width=12, fig.height=8}
if ("smooth_r2" %in% colnames(seurat_obj@meta.data) &&
    sum(!is.na(seurat_obj$smooth_r2)) > 0) {
  q05_s <- quantile(seurat_obj$smooth_r2, 0.05, na.rm = TRUE)
  FeaturePlot(seurat_obj, features = "smooth_r2",
              min.cutoff = q05_s,
              pt.size = 0.2, alpha = 0.5, raster = FALSE,
              order = TRUE) +
    scale_color_viridis(option = "viridis")
} else {
  cat("smooth_r2 not loaded (check scqc/discordance_scores.csv per sample)\n")
}
```

# Multi-Resolution Marker Analysis

```{r multi_res_markers, fig.width=14, fig.height=10}
cat("\n=== Multi-Resolution Marker Analysis ===\n\n")

# Pulls in both upstream leiden_* and any patches that wrote columns
# matching leiden_*. Patch columns added via 06c_recluster_integration.py
# show up automatically here.
resolution_cols <- grep("^leiden_", colnames(seurat_obj@meta.data), value = TRUE)
if (length(resolution_cols) == 0 && !is.null(leiden_col)) {
  resolution_cols <- leiden_col
}
cat("Clustering columns:", paste(resolution_cols, collapse = ", "), "\n\n")

# Top markers per cluster: 3 keeps the panel readable.
TOP_N_PER_CLUSTER <- 3

# DoHeatmap downsample size — individual-cell heatmaps stay legible up to a few
# thousand cells per panel; beyond that the per-cell columns collapse to noise.
HEATMAP_CELL_DOWNSAMPLE <- 5000

# Past this many clusters, even DoHeatmap's grouped columns become a pixel
# mess. Fall back to an average-expression tile heatmap (clusters as columns).
TILE_FALLBACK_CLUSTER_THRESHOLD <- 50

all_resolution_markers <- list()

plot_marker_panel <- function(seurat_obj, res_col, top_markers) {
  features <- unique(top_markers$feature)
  n_clusters <- length(unique(seurat_obj@meta.data[[res_col]]))
  Idents(seurat_obj) <- res_col

  if (n_clusters <= TILE_FALLBACK_CLUSTER_THRESHOLD) {
    cells_ds <- sample(colnames(seurat_obj),
                       size = min(HEATMAP_CELL_DOWNSAMPLE, ncol(seurat_obj)))
    p <- DoHeatmap(seurat_obj[, cells_ds],
                   features = features,
                   group.by = res_col,
                   slot     = "data",
                   size     = 3,
                   raster   = FALSE) +
      labs(subtitle = sprintf("%s: %d clusters, top %d markers/cluster (n=%d cells)",
                              res_col, n_clusters, TOP_N_PER_CLUSTER, length(cells_ds)))
    return(p)
  }

  # Tile fallback: average expression heatmap (clusters as columns, features as
  # rows). Used only when cluster count makes per-cell DoHeatmap unreadable.
  avg <- AverageExpression(seurat_obj, features = features,
                           assays = "RNA", slot = "data",
                           verbose = FALSE)$RNA
  avg_z <- t(scale(t(as.matrix(avg))))
  avg_z[is.na(avg_z)] <- 0
  avg_z[avg_z >  3]   <-  3
  avg_z[avg_z < -3]   <- -3

  df <- as.data.frame(as.table(avg_z))
  colnames(df) <- c("feature", "cluster", "z")
  df$feature <- factor(df$feature, levels = features)

  ggplot(df, aes(x = cluster, y = feature, fill = z)) +
    geom_tile() +
    scale_fill_viridis(option = "magma", name = "z(avg expr)") +
    theme_minimal(base_size = 8) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          panel.grid = element_blank()) +
    labs(x = res_col, y = NULL,
         subtitle = sprintf("%s: %d clusters → tile fallback", res_col, n_clusters))
}

for (i in seq_along(resolution_cols)) {
  res_col <- resolution_cols[i]
  cat("---\nResolution", i, "of", length(resolution_cols), ":", res_col, "\n")

  # Numeric-order cluster levels so heatmap columns and marker groups stay
  # ordered 0,1,2,...,N instead of alphabetical 0,1,10,11,2,20.
  seurat_obj@meta.data[[res_col]] <- order_cluster_factor(seurat_obj@meta.data[[res_col]])
  cluster_levels <- levels(seurat_obj@meta.data[[res_col]])
  Idents(seurat_obj) <- seurat_obj@meta.data[[res_col]]
  n_clusters <- length(cluster_levels)
  cat("  Clusters:", n_clusters, "\n")

  markers <- presto::wilcoxauc(seurat_obj, assay = "data")
  cat("  Marker tests:", nrow(markers), "\n")

  sig_markers <- markers %>%
    filter(padj < 0.05, logFC > 0) %>%
    mutate(group = factor(as.character(group), levels = cluster_levels)) %>%
    arrange(group, desc(logFC))
  cat("  Significant (padj<0.05, logFC>0):", nrow(sig_markers), "\n")

  top_markers <- sig_markers %>%
    group_by(group) %>%
    slice_max(order_by = logFC, n = TOP_N_PER_CLUSTER) %>%
    ungroup() %>%
    arrange(group, desc(logFC))

  all_resolution_markers[[res_col]] <- markers
  gc()

  if (nrow(top_markers) > 0) {
    tryCatch({
      print(plot_marker_panel(seurat_obj, res_col, top_markers))
    }, error = function(e) cat("  panel failed:", e$message, "\n"))
  }
  cat("\n")
}

cat("Multi-resolution marker analysis complete\n")
```

# Top Markers at Default Resolution {.tabset}

```{r top_markers_setup}
default_res_col <- if (!is.null(leiden_col)) leiden_col else resolution_cols[1]
cat("Default resolution column:", default_res_col, "\n")

top_markers <- NULL
if (!is.null(default_res_col) && default_res_col %in% names(all_resolution_markers)) {
  top_markers <- all_resolution_markers[[default_res_col]] %>%
    filter(padj < 0.05, logFC > 0) %>%
    group_by(group) %>%
    slice_max(order_by = logFC, n = TOP_N_PER_CLUSTER) %>%
    ungroup()
}
```

## Markers Panel

```{r dotplot_default, fig.width=16, fig.height=12}
if (!is.null(top_markers) && nrow(top_markers) > 0) {
  tryCatch({
    print(plot_marker_panel(seurat_obj, default_res_col, top_markers))
  }, error = function(e) cat("panel error:", e$message, "\n"))
} else {
  cat("No top markers available\n")
}
```

## Marker Table

```{r marker_table}
if (!is.null(top_markers) && nrow(top_markers) > 0) {
  top_markers %>%
    select(group, feature, logFC, auc, padj, pct_in, pct_out) %>%
    arrange(group, desc(logFC)) %>%
    head(100) %>%
    knitr::kable(digits = 3,
                 caption = paste0("Top markers for ", COMPARTMENT, " (", default_res_col, ")"))
}
```

# Cell Type Annotation Marker Analysis

Uses `cluster_annotation` (from the contamination CSV) as the grouping
variable. This catches cases where distinct annotated populations were
merged into a single Leiden cluster at the report's default resolution.

```{r celltype_markers, fig.width=14, fig.height=10}
celltype_markers <- NULL

if ("cluster_annotation" %in% colnames(seurat_obj@meta.data) &&
    sum(!is.na(seurat_obj$cluster_annotation)) > 0) {

  Idents(seurat_obj) <- "cluster_annotation"
  celltype_markers <- presto::wilcoxauc(seurat_obj, assay = "data")
  seurat_obj@misc$celltype_markers <- celltype_markers

  top_ct <- celltype_markers %>%
    filter(auc > 0.6, pct_in > 25) %>%
    group_by(group) %>%
    slice_max(order_by = auc, n = 15) %>%
    arrange(group, desc(auc))

  cat("\nTop 10 markers per cluster_annotation:\n")
  top_ct %>%
    group_by(group) %>%
    slice_head(n = 10) %>%
    select(group, feature, auc, pct_in, pct_out) %>%
    print(n = 50)

  # UMAP by cluster_annotation (restated here alongside markers for panel flow)
  p_ct <- DimPlot(seurat_obj,
                  group.by = "cluster_annotation",
                  raster = FALSE,
                  label = TRUE,
                  repel = TRUE,
                  pt.size = 0.4,
                  alpha = 0.5) + NoLegend()
  print(p_ct)

  # Heatmap markers: top-N per annotation, capped at ~60 total
  n_ct <- length(unique(na.omit(seurat_obj$cluster_annotation)))
  per_type <- ceiling(60 / max(n_ct, 1))

  markers_for_hm <- top_ct %>%
    group_by(group) %>%
    slice_head(n = per_type) %>%
    pull(feature) %>%
    unique()

  cat("\nHeatmap markers:", length(markers_for_hm), "\n")

  if (length(markers_for_hm) > 0) {
    cells_ds <- sample(colnames(seurat_obj), size = min(7500, ncol(seurat_obj)))
    cells_ds_prop <- tryCatch(
      subset(seurat_obj, downsample = 1500) %>% colnames(),
      error = function(e) cells_ds
    )

    tryCatch({
      p_hm <- DoHeatmap(seurat_obj[, cells_ds],
                        features = markers_for_hm,
                        group.by = "cluster_annotation",
                        slot = "data", size = 3)
      print(p_hm)
    }, error = function(e) cat("Random heatmap failed:", e$message, "\n"))

    tryCatch({
      p_hm_prop <- DoHeatmap(seurat_obj[, cells_ds_prop],
                             features = markers_for_hm,
                             group.by = "cluster_annotation",
                             slot = "data", size = 3) + NoLegend()
      print(p_hm_prop)
    }, error = function(e) cat("Proportional heatmap failed:", e$message, "\n"))
  }
} else {
  cat("cluster_annotation not available - skipping cell type marker analysis\n")
}
```

# Cluster Heatmap (Default Resolution)

```{r cluster_heatmap, fig.width=14, fig.height=12}
if (!is.null(default_res_col) && default_res_col %in% names(all_resolution_markers)) {
  def_markers <- all_resolution_markers[[default_res_col]]

  top_cluster <- def_markers %>%
    filter(auc > 0.55, pct_in > 20, logFC > 0) %>%
    group_by(group) %>%
    slice_max(order_by = auc, n = 10) %>%
    arrange(group, desc(auc))

  n_clusters <- length(unique(seurat_obj@meta.data[[default_res_col]]))
  per_cluster <- ceiling(50 / max(n_clusters, 1))

  hm_feats <- top_cluster %>%
    group_by(group) %>%
    slice_head(n = per_cluster) %>%
    pull(feature) %>%
    unique()

  cat("Cluster heatmap features:", length(hm_feats),
      "across", n_clusters, "clusters\n")

  if (length(hm_feats) > 0) {
    cells_hm <- sample(colnames(seurat_obj), size = min(5000, ncol(seurat_obj)))
    Idents(seurat_obj) <- default_res_col
    tryCatch({
      p <- DoHeatmap(seurat_obj[, cells_hm],
                     features = hm_feats,
                     group.by = default_res_col,
                     slot = "data", size = 3)
      print(p)
    }, error = function(e) cat("Cluster heatmap failed:", e$message, "\n"))
  }
}
```

# Save Outputs

```{r save_outputs}
cat("\n=== Saving outputs ===\n\n")

OUT_BASE <- file.path(OUTDIR, COMPARTMENT, CONFIG_LABEL)
dir.create(file.path(OUT_BASE, "markers_by_resolution"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(OUT_BASE, "metadata"),              recursive = TRUE, showWarnings = FALSE)

# Unified metadata
meta_path <- file.path(OUT_BASE, "metadata", "unified_metadata.csv")
write.csv(seurat_obj@meta.data, meta_path, row.names = TRUE)
cat("Metadata saved:", meta_path, "\n")

# Per-resolution markers
for (res_col in names(all_resolution_markers)) {
  fn <- paste0("markers_", res_col, ".csv")
  fp <- file.path(OUT_BASE, "markers_by_resolution", fn)
  write.csv(all_resolution_markers[[res_col]], fp, row.names = FALSE)
  cat("Saved:", fn, "\n")
}

# Combined long-form markers
if (length(all_resolution_markers) > 0) {
  combined <- bind_rows(
    lapply(names(all_resolution_markers), function(r) {
      all_resolution_markers[[r]] %>% mutate(resolution_col = r, .before = 1)
    })
  )
  write.csv(combined,
            file.path(OUT_BASE, "markers_by_resolution", "all_markers_all_resolutions.csv"),
            row.names = FALSE)
  cat("Combined markers saved\n")
}

# cluster_annotation markers
if (!is.null(celltype_markers)) {
  ca_path <- file.path(OUT_BASE, "markers_by_resolution", "markers_cluster_annotation.csv")
  write.csv(celltype_markers, ca_path, row.names = FALSE)
  cat("Annotation markers saved:", ca_path, "\n")
}
```

# Session Info

```{r session_info}
cat("\n============================================================================\n")
cat("Compartment Integration Report -", COMPARTMENT, "(", CONFIG_LABEL, ") - COMPLETE\n")
cat("============================================================================\n")
cat("Cells:  ", ncol(seurat_obj), "\n")
cat("Genes:  ", nrow(seurat_obj), "\n")
cat("Output: ", OUT_BASE, "\n")
cat("============================================================================\n")

sessionInfo()
```
