---
title: "Step 18: Integration Preview — `r params$COMPARTMENT` (harmony_n200 winner)"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
date: "`r Sys.Date()`"
params:
  PROJECT_ROOT: ""
  COMPARTMENT: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 12,
  fig.height = 8
)

suppressPackageStartupMessages({
  library(Seurat)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(presto)
  library(patchwork)
  library(viridis)
  library(knitr)
})

cat("Libraries loaded\n")
```

# Setup

```{r project_root}
PROJECT_ROOT <- params$PROJECT_ROOT
COMPARTMENT  <- params$COMPARTMENT

if (PROJECT_ROOT == "") {
  if (!require("here", quietly = TRUE)) stop("PROJECT_ROOT not provided and 'here' not available")
  PROJECT_ROOT <- here::here()
}

if (COMPARTMENT == "") stop("COMPARTMENT parameter is required")

# Resolve sweep winner from env (set by wrapper.R before render)
get_env <- function(key, default = NULL) {
  val <- Sys.getenv(key, unset = NA_character_)
  if (is.na(val) || val == "") return(default)
  val
}
SWEEP_WINNER <- get_env("CFG_CANONICAL_SWEEP_WINNER")
if (is.null(SWEEP_WINNER)) stop("CFG_CANONICAL_SWEEP_WINNER not set")

STEP_19B_DIR <- get_env("CFG_PREPROCESSING_STEP_19B",
                        file.path(PROJECT_ROOT, "outputs/preprocessing/19b_IntegrationPreview"))

cat("============================================================================\n")
cat("Step 19b: Integration Preview\n")
cat("============================================================================\n")
cat("Compartment:    ", COMPARTMENT, "\n")
cat("Project Root:   ", PROJECT_ROOT, "\n")
cat("Sweep winner:   ", SWEEP_WINNER, "\n")
cat("Step 19b dir:   ", STEP_19B_DIR, "\n")
cat("Integration:   harmony_n200 (winner)\n")
cat("============================================================================\n")
```

# Load Seurat Object and Winner Embeddings

**Adapted from step_21**: instead of loading raw H5 counts + step_20 embeddings, we load
the pre-built compartment Seurat (from `outputs/annotation/l1/`) and inject the
harmony_n200 UMAP + leiden clusters from the sweep winner sidecars.

```{r load_seurat}
SEURAT_RDS <- file.path(PROJECT_ROOT, "outputs/annotation/l1", COMPARTMENT, "seurat.rds")
WINNER_META <- file.path(SWEEP_WINNER, "compartments", COMPARTMENT, "metadata_with_umap.csv")

cat("Seurat RDS:", SEURAT_RDS, "\n")
cat("Winner metadata:", WINNER_META, "\n")

if (!file.exists(SEURAT_RDS)) stop("Seurat RDS not found: ", SEURAT_RDS)
if (!file.exists(WINNER_META)) stop("Winner metadata not found: ", WINNER_META)

seurat_merged <- readRDS(SEURAT_RDS)
seurat_merged <- UpdateSeuratObject(seurat_merged)
cat("\nLoaded Seurat:", ncol(seurat_merged), "cells x", nrow(seurat_merged), "genes\n")

# Inject winner UMAP coordinates and leiden clusters
winner_df <- read.csv(WINNER_META, row.names = 1, stringsAsFactors = FALSE)
cat("Winner metadata:", nrow(winner_df), "cells\n")

cells_common <- intersect(colnames(seurat_merged), rownames(winner_df))
cat("Matched cells:", length(cells_common), "/", ncol(seurat_merged), "\n")

seurat_merged <- subset(seurat_merged, cells = cells_common)

umap_mat <- as.matrix(winner_df[cells_common, c("UMAP_1", "UMAP_2")])
colnames(umap_mat) <- paste0("UMAP_", 1:2)
seurat_merged[["umap"]] <- CreateDimReducObject(
  embeddings = umap_mat, key = "UMAP_", assay = DefaultAssay(seurat_merged)
)

leiden_cols <- grep("^leiden_", colnames(winner_df), value = TRUE)
for (col in leiden_cols) {
  seurat_merged@meta.data[[col]] <- winner_df[rownames(seurat_merged@meta.data), col]
}
cat("Injected leiden columns:", paste(leiden_cols, collapse = ", "), "\n")

# Convert leiden columns to factors with numerically sorted levels
# Prevents lexicographic ordering (0,1,10,11...) in plots
for (col in leiden_cols) {
  vals <- as.character(seurat_merged@meta.data[[col]])
  unique_vals <- unique(vals[!is.na(vals)])
  lvls <- as.character(sort(as.numeric(unique_vals)))
  seurat_merged@meta.data[[col]] <- factor(vals, levels = lvls)
}
cat("Leiden columns converted to numerically sorted factors\n")

# Default resolution for main markers/dotplot view
default_cluster_col <- "leiden_1.0"
cat("Default cluster column:", default_cluster_col, "\n")

# Clustering columns for multi-resolution analysis
clustering_columns <- leiden_cols
cat("Clustering columns:", paste(clustering_columns, collapse = ", "), "\n")

cat("\n✓ Seurat prepared\n")
```

# Normalize and Join Layers

```{r normalize}
cat("\n=== Normalizing and Joining Layers ===\n\n")

gc()
seurat_merged <- NormalizeData(seurat_merged, verbose = FALSE)
cat("Normalization complete\n")

gc()
seurat_merged <- JoinLayers(seurat_merged)
cat("JoinLayers complete\n")

gc()
```

# Summary Statistics

```{r summary_stats}
cat("\n============================================================================\n")
cat("Summary Statistics for", COMPARTMENT, "\n")
cat("============================================================================\n\n")

cat("Total cells:  ", ncol(seurat_merged), "\n")
cat("Total genes:  ", nrow(seurat_merged), "\n")

cat("\nCells by patient:\n")
print(table(seurat_merged$patient_id))

cat("\nCells by sample:\n")
print(table(seurat_merged$sample_id))

if (default_cluster_col %in% colnames(seurat_merged@meta.data)) {
  cat("\nClusters at", default_cluster_col, ":\n")
  print(table(seurat_merged@meta.data[[default_cluster_col]]))
}
```

# UMAP Visualizations {.tabset}

## By Cluster (res 1.0)

```{r umap_clusters, fig.width=14, fig.height=10}
if (default_cluster_col %in% colnames(seurat_merged@meta.data)) {
  Idents(seurat_merged) <- default_cluster_col

  DimPlot(seurat_merged, reduction = "umap", group.by = default_cluster_col,
          label = TRUE, pt.size = 0.3, label.size = 4, raster = FALSE) +
    ggtitle(paste0(COMPARTMENT, " - Clusters (", default_cluster_col, ")")) +
    theme(legend.position = "right")
}
```

## By Cluster (res 0.2)

```{r umap_clusters_02, fig.width=14, fig.height=10}
if ("leiden_0.2" %in% colnames(seurat_merged@meta.data)) {
  DimPlot(seurat_merged, reduction = "umap", group.by = "leiden_0.2",
          label = TRUE, pt.size = 0.3, label.size = 4, raster = FALSE) +
    ggtitle(paste0(COMPARTMENT, " - Clusters (leiden_0.2)")) +
    theme(legend.position = "right")
}
```

## By Patient

```{r umap_patient, fig.width=12, fig.height=8}
DimPlot(seurat_merged, reduction = "umap", group.by = "patient_id",
        pt.size = 0.3, raster = FALSE, shuffle = TRUE) +
  ggtitle(paste0(COMPARTMENT, " - Patient Distribution"))
```

## By Sample

```{r umap_sample, fig.width=14, fig.height=10}
DimPlot(seurat_merged, reduction = "umap", group.by = "sample_id",
        pt.size = 0.2, raster = FALSE) +
  ggtitle(paste0(COMPARTMENT, " - Sample Distribution")) +
  theme(legend.text = element_text(size = 6))
```

## By Cluster Annotation

```{r umap_cluster_annot, fig.width=14, fig.height=10}
if ("cluster_annotation" %in% colnames(seurat_merged@meta.data)) {
  DimPlot(seurat_merged, reduction = "umap",
          group.by = "cluster_annotation",
          label = TRUE, repel = TRUE,
          pt.size = 0.3, alpha = 0.5, raster = FALSE) +
    ggtitle(paste0(COMPARTMENT, " - Cluster Annotations")) +
    NoLegend()
} else {
  cat("cluster_annotation not available\n")
}
```

# Load L1 Labels

```{r load_l1_labels}
# Load harmonized L1 labels from step 19 output and inject into Seurat metadata.
# Priority: (1) L1.0 already in Seurat, (2) harmonized_l1_labels CSV, (3) skip.
#
# Pattern source: cluster_label_hierarchy.py (iHBCA_V1 V1_Annotation) — same
# L1.0 column name and concordance logic used downstream.

L1_COL <- NULL

if ("L1.0" %in% colnames(seurat_merged@meta.data)) {
  L1_COL <- "L1.0"
  cat("L1 labels found in Seurat metadata column 'L1.0'\n")
} else {
  l1_csv <- file.path(PROJECT_ROOT, "outputs/annotation/l1", COMPARTMENT,
                      paste0("harmonized_l1_labels_", COMPARTMENT, ".csv"))
  if (file.exists(l1_csv)) {
    l1_df <- read.csv(l1_csv, stringsAsFactors = FALSE)
    cells_match <- intersect(colnames(seurat_merged), l1_df$cell_id)
    if (length(cells_match) > 0 && "L1.0" %in% colnames(l1_df)) {
      l1_vec <- setNames(rep(NA_character_, ncol(seurat_merged)), colnames(seurat_merged))
      l1_vec[cells_match] <- l1_df$L1.0[match(cells_match, l1_df$cell_id)]
      seurat_merged@meta.data[["L1.0"]] <- l1_vec
      L1_COL <- "L1.0"
      cat(sprintf("L1 labels loaded from CSV: %d / %d cells matched\n",
                  length(cells_match), ncol(seurat_merged)))
      cat(sprintf("  %d unique L1 types: %s\n",
                  length(unique(na.omit(l1_vec))),
                  paste(sort(unique(na.omit(l1_vec))), collapse = ", ")))
    } else {
      cat("WARNING: L1 CSV found but column 'L1.0' missing or no cell overlap\n")
    }
  } else {
    cat("WARNING: No L1 labels found — concordance analysis will be skipped\n")
    cat("  Expected:", l1_csv, "\n")
  }
}
```

# Cluster–Label Concordance Sweep

Compactness metric from iHBCA_V1 `cluster_label_hierarchy.py`: harmonic mean of
cluster purity and label purity (F1 of precision × recall).

- **cluster_purity** = dominant_label_cells / cluster_size — how pure is the cluster
- **label_purity** = dominant_label_cells / total_cells_with_label — how concentrated is the label
- **compactness** = 2 × cp × lp / (cp + lp) — low cp = merge; low lp = split

```{r concordance_sweep, fig.width=12, fig.height=8}
if (!is.null(L1_COL)) {

  # Per cluster: top label, cluster_purity, label_purity, compactness
  # Pattern: cluster_label_hierarchy.py lines 80-112
  compute_compactness <- function(seurat, cluster_col, label_col) {
    cl <- as.character(seurat@meta.data[[cluster_col]])
    lb <- as.character(seurat@meta.data[[label_col]])
    valid <- !is.na(cl) & !is.na(lb)
    cl <- cl[valid]; lb <- lb[valid]
    total_per_label <- table(lb)

    rows <- lapply(unique(cl), function(k) {
      k_mask    <- cl == k
      n_cluster <- sum(k_mask)
      vc        <- sort(table(lb[k_mask]), decreasing = TRUE)
      if (length(vc) == 0) return(NULL)
      top_label <- names(vc)[1]
      top_count <- as.integer(vc[1])
      c_purity  <- top_count / n_cluster
      l_purity  <- top_count / as.integer(total_per_label[top_label])
      compact   <- 2 * c_purity * l_purity / (c_purity + l_purity + 1e-10)
      data.frame(
        cluster        = k,
        n_cells        = n_cluster,
        top_label      = top_label,
        runner_up      = if (length(vc) > 1) names(vc)[2] else NA_character_,
        cluster_purity = round(c_purity, 3),
        label_purity   = round(l_purity, 3),
        compactness    = round(compact, 3),
        stringsAsFactors = FALSE
      )
    })
    do.call(rbind, Filter(Negate(is.null), rows))
  }

  # Sweep all available resolutions
  all_compactness <- list()
  sweep_df <- lapply(leiden_cols, function(col) {
    res <- as.numeric(sub("leiden_", "", col))
    df  <- compute_compactness(seurat_merged, col, L1_COL)
    all_compactness[[col]] <<- df
    data.frame(
      resolution         = res,
      mean_compactness   = mean(df$compactness, na.rm = TRUE),
      median_compactness = median(df$compactness, na.rm = TRUE),
      mean_cluster_purity = mean(df$cluster_purity, na.rm = TRUE),
      mean_label_purity  = mean(df$label_purity, na.rm = TRUE),
      n_clusters         = nrow(df),
      stringsAsFactors   = FALSE
    )
  }) |> do.call(what = rbind) |> (\(x) x[order(x$resolution), ])()

  opt_res  <- sweep_df$resolution[which.max(sweep_df$mean_compactness)]
  def_res  <- as.numeric(sub("leiden_", "", default_cluster_col))

  cat(sprintf("Optimal resolution (max mean compactness): %.1f\n", opt_res))
  cat(sprintf("Default resolution: %.1f\n", def_res))
  cat("\nSweep table:\n")
  print(sweep_df, digits = 3, row.names = FALSE)

  # -- Plot 1: compactness + component purities vs resolution --
  long_df <- sweep_df |>
    select(resolution, mean_compactness, mean_cluster_purity, mean_label_purity) |>
    pivot_longer(-resolution, names_to = "metric", values_to = "value") |>
    mutate(metric = recode(metric,
      mean_compactness    = "Compactness (F1)",
      mean_cluster_purity = "Cluster purity (precision)",
      mean_label_purity   = "Label purity (recall)"
    ))

  p_concordance <- ggplot(long_df, aes(x = resolution, y = value,
                                       color = metric, linetype = metric)) +
    geom_line(linewidth = 1.1) +
    geom_point(size = 2.5) +
    geom_vline(xintercept = def_res, linetype = "dotdash", alpha = 0.4) +
    annotate("text", x = def_res + 0.05, y = 0.98, label = paste0("default (", def_res, ")"),
             hjust = 0, size = 3, color = "grey40") +
    geom_vline(xintercept = opt_res, color = "#c0392b", linetype = "dashed", alpha = 0.7) +
    annotate("text", x = opt_res + 0.05, y = 0.93, label = paste0("optimal (", opt_res, ")"),
             hjust = 0, size = 3, color = "#c0392b") +
    scale_color_manual(values = c("Compactness (F1)" = "#2c7bb6",
                                  "Cluster purity (precision)" = "#4dac26",
                                  "Label purity (recall)"      = "#d7191c")) +
    scale_linetype_manual(values = c("Compactness (F1)" = "solid",
                                     "Cluster purity (precision)" = "dashed",
                                     "Label purity (recall)"      = "dotted")) +
    scale_y_continuous(limits = c(0, 1), labels = function(x) paste0(round(x * 100), "%")) +
    labs(title = paste0(COMPARTMENT, " — Cluster-Label Concordance vs Resolution"),
         x = "Leiden resolution", y = "Mean value across clusters",
         color = NULL, linetype = NULL) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")

  # -- Plot 2: N clusters vs resolution --
  p_nclusters <- ggplot(sweep_df, aes(x = resolution, y = n_clusters)) +
    geom_line(color = "#7b2d8b", linewidth = 1.1) +
    geom_point(color = "#7b2d8b", size = 2.5) +
    geom_vline(xintercept = def_res, linetype = "dotdash", alpha = 0.4) +
    geom_vline(xintercept = opt_res, color = "#c0392b", linetype = "dashed", alpha = 0.7) +
    labs(x = "Leiden resolution", y = "N clusters") +
    theme_minimal(base_size = 11)

  print(p_concordance / p_nclusters + plot_layout(heights = c(3, 1)))

  # -- Per-label label_purity heatmap --
  # For each L1 label, label_purity(label, r) = max label_purity across all
  # clusters where that label is the top label. Answers: "at resolution r, what
  # fraction of label X's cells are in their dominant cluster?"
  # Rows sorted by argmax resolution (early-resolving labels at top).
  leiden_cols_ordered <- leiden_cols[order(as.numeric(sub("leiden_", "", leiden_cols)))]
  sorted_res_vals     <- as.numeric(sub("leiden_", "", leiden_cols_ordered))

  all_labels <- sort(unique(as.character(
    seurat_merged@meta.data[[L1_COL]][!is.na(seurat_merged@meta.data[[L1_COL]])]
  )))

  label_purity_wide <- do.call(cbind, lapply(leiden_cols_ordered, function(col) {
    df <- all_compactness[[col]]
    sapply(all_labels, function(lbl) {
      rows <- df[!is.na(df$top_label) & df$top_label == lbl, ]
      if (nrow(rows) == 0) return(NA_real_)
      max(rows$label_purity, na.rm = TRUE)
    })
  }))
  colnames(label_purity_wide) <- as.character(sorted_res_vals)
  rownames(label_purity_wide) <- all_labels

  best_res_per_label <- setNames(
    apply(label_purity_wide, 1, function(row) {
      if (all(is.na(row))) return(NA_real_)
      as.numeric(colnames(label_purity_wide)[which.max(row)])
    }),
    rownames(label_purity_wide)
  )
  max_lp_per_label <- apply(label_purity_wide, 1, max, na.rm = TRUE)

  label_order        <- order(best_res_per_label, -max_lp_per_label)
  label_purity_wide  <- label_purity_wide[label_order, ]
  best_res_per_label <- best_res_per_label[label_order]
  max_lp_per_label   <- max_lp_per_label[label_order]

  lp_long <- data.frame(
    label      = factor(rep(rownames(label_purity_wide), ncol(label_purity_wide)),
                         levels = rownames(label_purity_wide)),
    resolution = rep(sorted_res_vals, each = nrow(label_purity_wide)),
    lp         = as.vector(label_purity_wide),
    stringsAsFactors = FALSE
  )
  best_df <- data.frame(
    label      = factor(names(best_res_per_label), levels = rownames(label_purity_wide)),
    resolution = best_res_per_label
  )

  p_lp_heatmap <- ggplot(lp_long, aes(x = factor(resolution), y = label, fill = lp)) +
    geom_tile(color = "white", linewidth = 0.3) +
    geom_point(data = best_df, aes(x = factor(resolution), y = label),
               shape = 8, size = 2.5, color = "black", inherit.aes = FALSE) +
    scale_fill_gradient(low = "#f7f7f7", high = "#2166ac", na.value = "grey90",
                        name = "label\npurity", limits = c(0, 1)) +
    labs(
      title    = paste0(COMPARTMENT, " — Per-label label_purity across resolutions"),
      subtitle = paste0("★ = best resolution for that label  |  rows sorted early→late resolver  |",
                        "  red dashed = r_L1 (", opt_res, ")  |  dotdash = default (", def_res, ")"),
      x = "Leiden resolution", y = NULL
    ) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "right",
          axis.text.y     = element_text(size = 8),
          panel.grid      = element_blank())
  print(p_lp_heatmap)

  # -- Per-label argmax summary: which labels require fine resolution? --
  best_summary <- data.frame(
    label  = factor(names(best_res_per_label), levels = rev(names(best_res_per_label))),
    best_r = best_res_per_label,
    max_lp = max_lp_per_label
  )
  p_best_res <- ggplot(best_summary, aes(x = best_r, y = label)) +
    geom_vline(xintercept = opt_res, color = "#c0392b", linetype = "dashed", alpha = 0.6) +
    geom_vline(xintercept = def_res, linetype = "dotdash", alpha = 0.4) +
    geom_point(aes(fill = max_lp, size = max_lp), shape = 21, stroke = 0.5) +
    scale_fill_gradient(low = "#fde0dd", high = "#b30000",
                        name = "max\nlabel purity", limits = c(0, 1)) +
    scale_size_continuous(range = c(2.5, 8), guide = "none") +
    scale_x_continuous(breaks = sorted_res_vals) +
    labs(
      title    = paste0(COMPARTMENT, " — Best resolution per L1 label"),
      subtitle = "Dot fill/size = label_purity at argmax resolution. Labels sorted top→bottom by early→late resolver.",
      x = "Leiden resolution at argmax label_purity", y = NULL
    ) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "right",
          axis.text.y     = element_text(size = 8))
  print(p_best_res)

} else {
  cat("Skipping concordance sweep: no L1 labels available\n")
  all_compactness <- list()
  opt_res <- as.numeric(sub("leiden_", "", default_cluster_col))
}
```

# Resolution × Marker Quality

For each cluster at each resolution, `auc_score_norm` measures how robustly the
cluster is distinguished by its top markers — the normalized trapezoidal integral
of the threshold-swept AUC curve (top 20 genes, thresholds 0.50–0.90).

Ported from iHBCA_V1 `04_cluster_distinctiveness.py:compute_area_scores()`, which
itself ports `calculate_marker_areas` from `scQC/rawPipelineDesignV1.Rmd:2585-2677`.

- `auc_score_norm` ∈ [0,1]: 1 = every top gene is a perfect discriminator
- A cluster **gaining** score at finer resolution = meaningful split
- A cluster **losing** score = over-splitting into noise

```{r auc_score_functions}
# Trapezoid integration (base R, no packages needed)
trapz <- function(x, y) sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)

# Port of 04_cluster_distinctiveness.py:compute_area_scores()
AUC_THRESHOLDS   <- seq(0.50, 0.90, length.out = 41)
LOGFC_THRESHOLDS <- seq(0.00, 4.00, length.out = 81)
N_TOP_MARKERS    <- 20

compute_area_scores <- function(aucs, logfcs) {
  n   <- min(N_TOP_MARKERS, length(aucs))
  top <- order(aucs, decreasing = TRUE)[seq_len(n)]
  a   <- aucs[top]
  l   <- logfcs[top]

  auc_areas   <- sapply(AUC_THRESHOLDS,   function(t) sum(pmax(0, a - t)) / (1 - t))
  logfc_areas <- sapply(LOGFC_THRESHOLDS, function(t) sum(pmax(0, l - t)))

  auc_raw  <- trapz(AUC_THRESHOLDS, auc_areas)
  auc_max  <- n * (max(AUC_THRESHOLDS) - min(AUC_THRESHOLDS))
  list(
    auc_score      = auc_raw,
    auc_score_norm = if (auc_max > 0) auc_raw / auc_max else 0.0,
    logfc_score    = trapz(LOGFC_THRESHOLDS, logfc_areas),
    strong_markers = sum(a > 0.7 & l > 1.0)
  )
}
```

```{r compute_all_markers, cache=FALSE}
# Run presto::wilcoxauc for each leiden resolution; store results in a named
# list consumed by the marker_quality_sweep chunk below.
# Pattern: presto::wilcoxauc per Leiden resolution; auc/logFC ranking.
cat("\n=== Computing markers per resolution ===\n")

seurat_norm <- NormalizeData(seurat_merged, verbose = FALSE)
seurat_norm <- JoinLayers(seurat_norm)

all_resolution_markers <- setNames(
  lapply(leiden_cols, function(res_col) {
    cat(sprintf("  wilcoxauc: %s ... ", res_col))
    Idents(seurat_norm) <- res_col
    mrk <- presto::wilcoxauc(seurat_norm, assay = "data")
    cat(sprintf("%d rows\n", nrow(mrk)))
    # Convert group to numerically-sorted factor to prevent lexicographic
    # ordering (0,1,10,11...) in all downstream plots and tables.
    grp_levels <- as.character(sort(as.numeric(unique(mrk$group))))
    mrk$group  <- factor(mrk$group, levels = grp_levels)
    mrk
  }),
  leiden_cols
)

cat(sprintf("Markers computed for %d resolutions\n", length(all_resolution_markers)))
rm(seurat_norm); gc()
```

```{r marker_quality_sweep, fig.width=14, fig.height=10}
cat("\n=== Resolution × Marker Quality ===\n\n")

if (length(all_resolution_markers) > 0) {

  # Compute auc_score_norm per (resolution, cluster) from existing marker data
  quality_rows <- lapply(names(all_resolution_markers), function(res_col) {
    res  <- as.numeric(sub("leiden_", "", res_col))
    mrk  <- all_resolution_markers[[res_col]]
    cl   <- as.character(seurat_merged@meta.data[[res_col]])

    lapply(unique(mrk$group), function(k) {
      grp_mrk <- mrk[mrk$group == k, ]
      grp_mrk <- grp_mrk[order(grp_mrk$auc, decreasing = TRUE), ]
      if (nrow(grp_mrk) == 0) return(NULL)
      scores <- compute_area_scores(grp_mrk$auc, grp_mrk$logFC)
      top3   <- head(grp_mrk$feature[grp_mrk$auc > 0.5], 3)
      top3   <- c(top3, rep(NA_character_, 3 - length(top3)))
      data.frame(
        resolution     = res,
        cluster        = as.character(k),
        n_cells        = sum(!is.na(cl) & cl == k),
        auc_score_norm = round(scores$auc_score_norm, 4),
        logfc_score    = round(scores$logfc_score, 4),
        strong_markers = scores$strong_markers,
        top_marker_1   = top3[1],
        top_marker_2   = top3[2],
        top_marker_3   = top3[3],
        stringsAsFactors = FALSE
      )
    }) |> do.call(what = rbind)
  }) |> do.call(what = rbind)

  cat(sprintf("Computed auc_score_norm for %d (resolution × cluster) combinations\n",
              nrow(quality_rows)))

  # -- Resolution summary table --
  res_summary <- quality_rows |>
    group_by(resolution) |>
    summarise(
      n_clusters    = n(),
      mean_auc      = round(mean(auc_score_norm,  na.rm = TRUE), 3),
      median_auc    = round(median(auc_score_norm, na.rm = TRUE), 3),
      frac_above_05 = round(mean(auc_score_norm >= 0.5, na.rm = TRUE), 3),
      frac_above_07 = round(mean(auc_score_norm >= 0.7, na.rm = TRUE), 3),
      worst_cluster = cluster[which.min(auc_score_norm)],
      worst_auc     = round(min(auc_score_norm, na.rm = TRUE), 3),
      .groups = "drop"
    ) |>
    arrange(resolution)
  cat("\nResolution summary:\n")
  print(as.data.frame(res_summary), row.names = FALSE)

  # -- Plot: split delta arrows --
  # For each pair of adjacent resolutions, draw parent_auc → child_auc as a
  # segment per split. Color = gain (green) / loss (red) / neutral (grey).
  # Child node size ∝ n_cells.
  # Interpretation: green rising lines = split is meaningful (biology resolves);
  #   red falling lines = over-splitting into noise (merge back).
  sorted_res  <- sort(unique(quality_rows$resolution))
  sorted_cols <- paste0("leiden_", sorted_res)
  sorted_cols <- sorted_cols[sorted_cols %in% leiden_cols]

  auc_lookup <- setNames(
    lapply(sorted_res, function(r) {
      qr <- quality_rows[abs(quality_rows$resolution - r) < 0.001, ]
      setNames(qr$auc_score_norm, as.character(qr$cluster))
    }),
    as.character(sorted_res)
  )

  edges <- do.call(rbind, lapply(seq_len(length(sorted_cols) - 1), function(i) {
    coarse_col <- sorted_cols[i]
    fine_col   <- sorted_cols[i + 1]
    coarse_res <- sorted_res[i]
    fine_res   <- sorted_res[i + 1]

    cc    <- as.character(seurat_merged@meta.data[[coarse_col]])
    fc    <- as.character(seurat_merged@meta.data[[fine_col]])
    valid <- !is.na(cc) & !is.na(fc)

    tab        <- table(fine = fc[valid], coarse = cc[valid])
    parent_map <- apply(tab, 1, function(row) names(which.max(row)))

    p_auc <- auc_lookup[[as.character(coarse_res)]]
    c_auc <- auc_lookup[[as.character(fine_res)]]

    rows <- lapply(names(parent_map), function(child) {
      parent <- parent_map[[child]]
      pa <- p_auc[parent]; ca <- c_auc[child]
      if (is.na(pa) || is.na(ca)) return(NULL)
      data.frame(
        transition = paste0(coarse_res, " → ", fine_res),
        coarse_res = coarse_res,
        parent     = parent,
        child      = child,
        parent_auc = pa,
        child_auc  = ca,
        delta      = ca - pa,
        n_cells    = sum(valid & fc == child, na.rm = TRUE),
        stringsAsFactors = FALSE
      )
    })
    do.call(rbind, Filter(Negate(is.null), rows))
  }))

  if (!is.null(edges) && nrow(edges) > 0) {
    edges$direction <- ifelse(edges$delta >  0.03, "gain",
                       ifelse(edges$delta < -0.03, "loss", "neutral"))
    edges$transition <- factor(
      edges$transition,
      levels = unique(edges$transition[order(edges$coarse_res)])
    )

    p_splits <- ggplot(edges) +
      geom_segment(aes(x = 0, xend = 1,
                       y = parent_auc, yend = child_auc,
                       color = direction),
                   linewidth = 0.8, alpha = 0.75) +
      geom_point(aes(x = 0, y = parent_auc),
                 color = "grey30", size = 2.5) +
      geom_point(aes(x = 1, y = child_auc,
                     size = n_cells, color = direction), alpha = 0.85) +
      geom_hline(yintercept = 0.5, linetype = "dashed",
                 color = "grey40", alpha = 0.6) +
      scale_color_manual(
        values = c(gain = "#2ca25f", loss = "#e34a33", neutral = "grey60"),
        name   = "Split direction"
      ) +
      scale_size_continuous(range = c(1.5, 6), guide = "none") +
      scale_x_continuous(breaks = c(0, 1), labels = c("parent", "child"),
                         expand = expansion(mult = 0.2)) +
      scale_y_continuous(limits = c(0, 1)) +
      facet_wrap(~ transition, nrow = 1) +
      labs(
        title    = paste0(COMPARTMENT, " — Split quality: auc_score_norm parent → child"),
        subtitle = "Green = gaining score (split justified) | Red = losing score (over-splitting) | size ∝ n_cells | dashed = 0.5",
        x = NULL, y = "auc_score_norm"
      ) +
      theme_minimal(base_size = 10) +
      theme(legend.position  = "bottom",
            strip.text       = element_text(face = "bold", size = 9),
            axis.text.x      = element_text(face = "bold", size = 9))
    print(p_splits)
  } else {
    cat("Split delta plot not available: only one resolution present\n")
  }

  # Summary table: low-quality clusters at default resolution
  low_quality <- quality_rows[
    quality_rows$resolution == as.numeric(sub("leiden_", "", default_cluster_col)) &
    quality_rows$auc_score_norm < 0.5, ]
  cat(sprintf("\nLow-quality clusters at %s (auc_score_norm < 0.5): %d\n",
              default_cluster_col, nrow(low_quality)))
  if (nrow(low_quality) > 0) print(low_quality[order(low_quality$auc_score_norm), ],
                                    row.names = FALSE)

  # Save for downstream use (L2 review Rmd)
  quality_out <- file.path(STEP_19B_DIR, COMPARTMENT, "markers_by_resolution",
                            "marker_quality_sweep.csv")
  write.csv(quality_rows, quality_out, row.names = FALSE)
  cat(sprintf("Saved: %s\n", quality_out))

} else {
  cat("Skipping: all_resolution_markers not populated\n")
  quality_rows <- NULL
}
```

# Split & Merge Analysis {.tabset}

For each cluster at the chosen resolution, compactness decomposes into:

- **Low cluster_purity → merge**: multiple L1 types coexist in one cluster
- **Low label_purity → split**: one L1 type is fragmented across multiple clusters

Threshold for flagging: compactness < 0.6 (same stability threshold as iHBCA_V1).
Marker investigation uses `presto::wilcoxauc()` on cell subsets — clusters for splits,
L1 labels for merges.

```{r split_merge_setup}
COMPACTNESS_THRESHOLD <- 0.6  # below = unstable (split or merge)
SPLIT_MERGE_TOP_N    <- 5     # max cases to investigate per category

# Use optimal resolution for split/merge analysis (fall back to default)
sm_col <- paste0("leiden_", opt_res)
if (!sm_col %in% leiden_cols) sm_col <- default_cluster_col
cat(sprintf("Split/merge analysis at resolution: %s\n", sm_col))

compact_df <- if (sm_col %in% names(all_compactness)) {
  all_compactness[[sm_col]]
} else if (!is.null(L1_COL)) {
  compute_compactness(seurat_merged, sm_col, L1_COL)
} else {
  NULL
}

# Per-label summary: for each L1 type, find its best-case label_purity
# (fraction of its cells concentrated in its dominant cluster)
if (!is.null(compact_df)) {
  cl  <- as.character(seurat_merged@meta.data[[sm_col]])
  lb  <- as.character(seurat_merged@meta.data[[L1_COL]])
  total_per_label <- table(lb[!is.na(lb)])

  label_summary <- lapply(names(total_per_label), function(lbl) {
    rows <- compact_df[compact_df$top_label == lbl, ]
    if (nrow(rows) == 0) return(NULL)
    best_row <- rows[which.max(rows$label_purity), ]
    data.frame(
      label              = lbl,
      total_cells        = as.integer(total_per_label[lbl]),
      best_cluster       = best_row$cluster,
      best_cluster_purity = best_row$cluster_purity,
      best_label_purity  = best_row$label_purity,
      compactness        = best_row$compactness,
      n_dominant_clusters = nrow(rows),
      stringsAsFactors   = FALSE
    )
  }) |> do.call(what = rbind) |> (\(x) x[order(x$best_label_purity), ])()

  merge_cases <- compact_df[compact_df$cluster_purity < COMPACTNESS_THRESHOLD, ]
  merge_cases <- merge_cases[order(merge_cases$cluster_purity), ]
  split_cases <- label_summary[label_summary$best_label_purity < COMPACTNESS_THRESHOLD, ]

  cat(sprintf("\nMerge candidates (cluster_purity < %.1f): %d clusters\n",
              COMPACTNESS_THRESHOLD, nrow(merge_cases)))
  cat(sprintf("Split candidates (label_purity  < %.1f): %d labels\n",
              COMPACTNESS_THRESHOLD, nrow(split_cases)))
}
```

## Confusion Matrix

```{r confusion_heatmap, fig.width=14, fig.height=10}
if (!is.null(compact_df) && !is.null(L1_COL)) {
  cl <- as.character(seurat_merged@meta.data[[sm_col]])
  lb <- as.character(seurat_merged@meta.data[[L1_COL]])
  valid <- !is.na(cl) & !is.na(lb)

  tab <- as.data.frame(table(cluster = cl[valid], label = lb[valid]))
  tab_norm <- tab |>
    group_by(label) |>
    mutate(frac = Freq / sum(Freq)) |>
    ungroup()

  cl_levels <- as.character(sort(as.numeric(unique(cl[valid]))))
  lb_levels <- names(sort(table(lb[valid]), decreasing = TRUE))
  tab_norm$cluster <- factor(tab_norm$cluster, levels = cl_levels)
  tab_norm$label   <- factor(tab_norm$label,   levels = lb_levels)

  p_conf <- ggplot(tab_norm, aes(x = cluster, y = label, fill = frac)) +
    geom_tile(color = "white", linewidth = 0.2) +
    scale_fill_viridis(option = "magma", direction = -1,
                       labels = function(x) paste0(round(x * 100), "%"),
                       name = "% of label\nin cluster") +
    labs(title = paste0(COMPARTMENT, " — L1 Label × Cluster (", sm_col, ")"),
         subtitle = "Color = fraction of each L1 label's cells in each cluster (row-normalized)",
         x = "Leiden cluster", y = "L1 label") +
    theme_minimal(base_size = 9) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
          axis.text.y = element_text(size = 7))
  print(p_conf)
} else {
  cat("Confusion matrix not available: missing L1 labels or compactness data\n")
}
```

## Confusion Matrices (All Resolutions)

Row-normalized (same as above) across every leiden resolution. L1 label order is
fixed by decreasing total cell count so rows are comparable across panels.

```{r confusion_all_resolutions, fig.width=18, fig.height=6}
if (!is.null(L1_COL)) {
  lb_all  <- as.character(seurat_merged@meta.data[[L1_COL]])
  lb_levels <- names(sort(table(lb_all[!is.na(lb_all)]), decreasing = TRUE))

  all_conf <- lapply(leiden_cols, function(col) {
    res <- as.numeric(sub("leiden_", "", col))
    cl  <- as.character(seurat_merged@meta.data[[col]])
    valid <- !is.na(cl) & !is.na(lb_all)
    tab <- as.data.frame(table(cluster = cl[valid], label = lb_all[valid]))
    tab |>
      group_by(label) |>
      mutate(frac = Freq / sum(Freq), resolution = res) |>
      ungroup()
  }) |> do.call(what = rbind)

  # Numeric cluster ordering per resolution
  all_conf$cluster <- factor(
    all_conf$cluster,
    levels = as.character(sort(as.numeric(unique(all_conf$cluster))))
  )
  all_conf$label <- factor(all_conf$label, levels = lb_levels)
  all_conf$res_label <- paste0("res ", all_conf$resolution)
  all_conf$res_label <- factor(all_conf$res_label,
                                levels = paste0("res ", sort(unique(all_conf$resolution))))

  p_all <- ggplot(all_conf, aes(x = cluster, y = label, fill = frac)) +
    geom_tile(color = "white", linewidth = 0.15) +
    scale_fill_viridis(option = "magma", direction = -1,
                       labels = function(x) paste0(round(x * 100), "%"),
                       name = "% of label") +
    facet_wrap(~ res_label, scales = "free_x", nrow = 1) +
    labs(title = paste0(COMPARTMENT, " — L1 Label × Cluster across all resolutions"),
         x = "Leiden cluster", y = NULL) +
    theme_minimal(base_size = 8) +
    theme(axis.text.x  = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 6),
          axis.text.y  = element_text(size = 7),
          strip.text   = element_text(size = 9, face = "bold"),
          panel.spacing = unit(0.3, "lines"),
          legend.position = "right")

  print(p_all)
} else {
  cat("Skipping: no L1 labels available\n")
}
```

## Merges

Clusters where cluster_purity < `r COMPACTNESS_THRESHOLD`: the dominant L1 label accounts for
fewer than `r paste0(round(COMPACTNESS_THRESHOLD*100), "%")` of cells. These clusters contain
multiple L1 types — either genuine intermediates or integration artifacts.

```{r merge_cases, fig.width=12, fig.height=6}
if (!is.null(compact_df) && nrow(merge_cases) > 0) {
  cat(sprintf("%d merge-candidate clusters (showing top %d by lowest cluster_purity):\n\n",
              nrow(merge_cases), SPLIT_MERGE_TOP_N))
  print(head(merge_cases, 20), row.names = FALSE)

  top_merges <- head(merge_cases, SPLIT_MERGE_TOP_N)

  for (i in seq_len(nrow(top_merges))) {
    k <- top_merges$cluster[i]
    cp <- top_merges$cluster_purity[i]
    ru <- top_merges$runner_up[i]
    tl <- top_merges$top_label[i]

    cat(sprintf("\n--- Cluster %s (cluster_purity=%.2f, top=%s, runner_up=%s) ---\n",
                k, cp, tl, ifelse(is.na(ru), "none", ru)))

    # Subset to this cluster's cells, use L1 label as group
    cluster_cells <- colnames(seurat_merged)[
      !is.na(seurat_merged@meta.data[[sm_col]]) &
      as.character(seurat_merged@meta.data[[sm_col]]) == k
    ]
    if (length(cluster_cells) < 20) {
      cat("  Skipping (< 20 cells)\n"); next
    }
    sub_seu <- seurat_merged[, cluster_cells]
    lbl_in_cluster <- as.character(sub_seu@meta.data[[L1_COL]])
    present_labels  <- names(which(table(lbl_in_cluster) >= 10))
    if (length(present_labels) < 2) {
      cat("  Skipping (< 2 labels with ≥10 cells)\n"); next
    }

    sub_seu@meta.data[["_merge_label"]] <- ifelse(
      lbl_in_cluster %in% present_labels, lbl_in_cluster, NA_character_)
    sub_seu <- subset(sub_seu, cells = colnames(sub_seu)[!is.na(sub_seu@meta.data[["_merge_label"]])])
    Idents(sub_seu) <- "_merge_label"

    tryCatch({
      mrk <- presto::wilcoxauc(sub_seu, assay = "data") |>
        filter(padj < 0.05, logFC > 0, auc > 0.6) |>
        group_by(group) |>
        slice_max(order_by = auc, n = 5) |>
        ungroup()
      if (nrow(mrk) == 0) { cat("  No significant markers\n"); next }
      top_feats <- unique(mrk$feature)
      p <- DotPlot(sub_seu, features = top_feats, group.by = "_merge_label") +
        coord_flip() + scale_color_viridis(option = "viridis") +
        ggtitle(paste0("Cluster ", k, " — merge markers (", sm_col, ")")) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
      print(p)
    }, error = function(e) cat("  Marker error:", e$message, "\n"))
  }
} else {
  cat(sprintf("No merge candidates at threshold %.1f\n", COMPACTNESS_THRESHOLD))
}
```

## Splits

L1 labels where best_label_purity < `r COMPACTNESS_THRESHOLD`: even the dominant cluster captures
fewer than `r paste0(round(COMPACTNESS_THRESHOLD*100), "%")` of this label's cells. These types
are fragmented across clusters — potential sub-populations or integration artifacts.

```{r split_cases, fig.width=12, fig.height=6}
if (!is.null(compact_df) && nrow(split_cases) > 0) {
  cat(sprintf("%d split-candidate labels (showing all, top %d investigated):\n\n",
              nrow(split_cases), SPLIT_MERGE_TOP_N))
  print(split_cases, row.names = FALSE)

  top_splits <- head(split_cases, SPLIT_MERGE_TOP_N)

  for (i in seq_len(nrow(top_splits))) {
    lbl    <- top_splits$label[i]
    lp     <- top_splits$best_label_purity[i]
    n_dom  <- top_splits$n_dominant_clusters[i]

    cat(sprintf("\n--- %s (label_purity=%.2f, dominant in %d clusters) ---\n",
                lbl, lp, n_dom))

    # Subset to cells with this L1 label; group by cluster
    label_cells <- colnames(seurat_merged)[
      !is.na(seurat_merged@meta.data[[L1_COL]]) &
      as.character(seurat_merged@meta.data[[L1_COL]]) == lbl
    ]
    if (length(label_cells) < 20) {
      cat("  Skipping (< 20 cells)\n"); next
    }
    sub_seu <- seurat_merged[, label_cells]
    cl_in_label <- as.character(sub_seu@meta.data[[sm_col]])
    # Keep only top-2 clusters by cell count to focus the comparison
    cl_counts   <- sort(table(cl_in_label), decreasing = TRUE)
    top2_cl     <- names(cl_counts)[seq_len(min(2, length(cl_counts)))]
    if (length(top2_cl) < 2) { cat("  Skipping (< 2 clusters)\n"); next }

    sub_seu <- subset(sub_seu, cells = colnames(sub_seu)[cl_in_label %in% top2_cl])
    sub_seu@meta.data[["_split_cluster"]] <- as.character(
      sub_seu@meta.data[[sm_col]])
    Idents(sub_seu) <- "_split_cluster"

    cat(sprintf("  Comparing clusters %s (%d cells) vs %s (%d cells)\n",
                top2_cl[1], as.integer(cl_counts[top2_cl[1]]),
                top2_cl[2], as.integer(cl_counts[top2_cl[2]])))

    tryCatch({
      mrk <- presto::wilcoxauc(sub_seu, assay = "data") |>
        filter(padj < 0.05, logFC > 0, auc > 0.6) |>
        group_by(group) |>
        slice_max(order_by = auc, n = 5) |>
        ungroup()
      if (nrow(mrk) == 0) { cat("  No significant markers\n"); next }
      top_feats <- unique(mrk$feature)
      p <- DotPlot(sub_seu, features = top_feats, group.by = "_split_cluster") +
        coord_flip() + scale_color_viridis(option = "viridis") +
        ggtitle(paste0(lbl, " — split markers (top-2 clusters, ", sm_col, ")")) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
      print(p)
    }, error = function(e) cat("  Marker error:", e$message, "\n"))
  }
} else {
  cat(sprintf("No split candidates at threshold %.1f\n", COMPACTNESS_THRESHOLD))
}
```

# QC Metric Overlays {.tabset}

## UMI Counts

```{r qc_umi, fig.width=12, fig.height=8}
q95_counts <- quantile(seurat_merged$nCount_RNA, 0.95, na.rm = TRUE)

FeaturePlot(seurat_merged, features = "nCount_RNA",
            max.cutoff = q95_counts, pt.size = 0.2, alpha = 0.5, raster = FALSE) +
  scale_color_viridis(option = "viridis", direction = -1) +
  ggtitle(paste0(COMPARTMENT, " - UMI Counts per Cell"))
```

## Genes Detected

```{r qc_genes, fig.width=12, fig.height=8}
q95_features <- quantile(seurat_merged$nFeature_RNA, 0.95, na.rm = TRUE)

FeaturePlot(seurat_merged, features = "nFeature_RNA",
            max.cutoff = q95_features, pt.size = 0.2, alpha = 0.5, raster = FALSE) +
  scale_color_viridis(option = "viridis") +
  ggtitle(paste0(COMPARTMENT, " - Genes Detected per Cell"))
```

# Multi-Resolution Marker Analysis

```{r marker_analysis, fig.width=14, fig.height=10}
cat("\n=== Multi-Resolution Marker Analysis ===\n\n")

gc()

all_resolution_markers <- list()

for (i in seq_along(clustering_columns)) {
  res_col <- clustering_columns[i]

  cat("---\n")
  cat("Resolution", i, "of", length(clustering_columns), ":", res_col, "\n")

  Idents(seurat_merged) <- seurat_merged@meta.data[[res_col]]

  n_clusters <- length(unique(Idents(seurat_merged)))
  cat("  Clusters:", n_clusters, "\n")

  cat("  Running presto::wilcoxauc()...\n")
  markers <- presto::wilcoxauc(seurat_merged, assay = "data")

  cat("  Found", nrow(markers), "marker tests\n")

  sig_markers <- markers %>%
    filter(padj < 0.05, logFC > 0) %>%
    arrange(group, desc(logFC))

  cat("  Significant markers (padj < 0.05, logFC > 0):", nrow(sig_markers), "\n")

  top_markers <- sig_markers %>%
    group_by(group) %>%
    slice_max(order_by = logFC, n = 5) %>%
    ungroup()

  all_resolution_markers[[res_col]] <- markers

  gc()

  if (nrow(top_markers) > 0) {
    tryCatch({
      p_dotplot <- DotPlot(
        seurat_merged,
        features = unique(top_markers$feature),
        group.by = res_col
      ) +
        coord_flip() +
        scale_color_viridis(option = "viridis") +
        ggtitle(paste0(COMPARTMENT, " - Top Markers (", res_col, ")")) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
              axis.text.y = element_text(size = 6))

      print(p_dotplot)
    }, error = function(e) {
      cat("  Dotplot generation failed:", e$message, "\n")
    })
  }

  cat("\n")
}

cat("Multi-resolution marker analysis complete\n")
```

# Top Markers by Cluster {.tabset}

```{r top_markers_setup}
if (default_cluster_col %in% names(all_resolution_markers)) {
  default_markers <- all_resolution_markers[[default_cluster_col]]

  top_markers <- default_markers %>%
    filter(padj < 0.05, logFC > 0) %>%
    group_by(group) %>%
    slice_max(order_by = logFC, n = 5) %>%
    ungroup()
}
```

## DotPlot

```{r dotplot, fig.width=16, fig.height=12}
if (exists("top_markers") && nrow(top_markers) > 0) {
  Idents(seurat_merged) <- default_cluster_col

  tryCatch({
    DotPlot(seurat_merged, features = unique(top_markers$feature),
            group.by = default_cluster_col) +
      coord_flip() +
      scale_color_viridis(option = "viridis") +
      ggtitle(paste0(COMPARTMENT, " - Top Markers (", default_cluster_col, ")")) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
            axis.text.y = element_text(size = 7))
  }, error = function(e) {
    cat("DotPlot error:", e$message, "\n")
  })
}
```

## Marker Table

```{r marker_table}
if (exists("top_markers") && nrow(top_markers) > 0) {
  top_markers %>%
    select(group, feature, logFC, auc, padj, pct_in, pct_out) %>%
    arrange(group, desc(logFC)) %>%
    head(100) %>%
    knitr::kable(digits = 3,
                 caption = paste0("Top markers for ", COMPARTMENT,
                                  " (", default_cluster_col, ")"))
}
```

# Cell Type Annotation Analysis

```{r celltype_analysis, fig.width=14, fig.height=10}
if ("cluster_annotation" %in% colnames(seurat_merged@meta.data)) {

  cat("\n=== Cell Type Annotation Analysis ===\n\n")

  p_celltype <- DimPlot(seurat_merged,
          group.by = "cluster_annotation",
          raster = FALSE,
          label = TRUE,
          repel = TRUE,
          pt.size = 0.4,
          alpha = 0.5) +
    ggtitle(paste0(COMPARTMENT, " - Cell Type Annotations")) +
    NoLegend()

  print(p_celltype)

  Idents(seurat_merged) <- "cluster_annotation"
  celltype_markers <- presto::wilcoxauc(seurat_merged, assay = "data")

  top_celltype_markers <- celltype_markers %>%
    filter(auc > 0.6, pct_in > 25) %>%
    group_by(group) %>%
    slice_max(order_by = auc, n = 15) %>%
    arrange(group, desc(auc))

  cat("\n=== TOP 10 MARKERS PER CELL TYPE ===\n")
  top_celltype_markers %>%
    group_by(group) %>%
    slice_head(n = 10) %>%
    select(group, feature, auc, pct_in, pct_out) %>%
    print(n = 50)

  n_celltypes <- length(unique(seurat_merged$cluster_annotation))
  markers_per_type <- ceiling(60 / n_celltypes)

  markers_for_heatmap <- top_celltype_markers %>%
    group_by(group) %>%
    slice_head(n = markers_per_type) %>%
    pull(feature) %>%
    unique()

  cat("\nPlotting", length(markers_for_heatmap), "markers across cell types\n")

  cells_downsample <- sample(colnames(seurat_merged), size = min(7500, ncol(seurat_merged)))
  cells_downsample_prop <- subset(seurat_merged, downsample = 1500) %>% colnames()

  if (length(markers_for_heatmap) > 0) {
    tryCatch({
      p_heatmap <- DoHeatmap(seurat_merged[, cells_downsample],
                             features = markers_for_heatmap,
                             group.by = "cluster_annotation",
                             slot = "data", size = 3) +
        ggtitle(paste0(COMPARTMENT, " - Cell Type Markers (Random Sample)"))
      print(p_heatmap)
    }, error = function(e) {
      cat("  Random heatmap failed:", e$message, "\n")
    })

    tryCatch({
      p_heatmap_prop <- DoHeatmap(seurat_merged[, cells_downsample_prop],
                                  features = markers_for_heatmap,
                                  group.by = "cluster_annotation",
                                  slot = "data", size = 3) +
        ggtitle(paste0(COMPARTMENT, " - Cell Type Markers (Proportional Sample)")) +
        NoLegend()
      print(p_heatmap_prop)
    }, error = function(e) {
      cat("  Proportional heatmap failed:", e$message, "\n")
    })
  }

} else {
  cat("\nNOTE: cluster_annotation not found in Seurat metadata — skipping.\n")
}
```

# Cluster Heatmaps (All Resolutions)

Top markers per cluster (auc > 0.55, pct_in > 20, logFC > 0) shown as DoHeatmap
for each leiden resolution. Each resolution produces two panels:

- **Random sample** (up to 5,000 cells): representative density
- **Proportional sample** (1,500 cells per cluster): equal visual weight per cluster

```{r cluster_heatmaps_all, results='asis', fig.width=16, fig.height=16}
for (res_col in leiden_cols) {
  res <- as.numeric(sub("leiden_", "", res_col))

  # Emit a TOC-visible H2 section header per resolution
  cat(sprintf("\n\n## Resolution %s\n\n", res_col))

  if (!res_col %in% names(all_resolution_markers)) {
    cat("No markers computed for this resolution.\n\n"); next
  }

  res_markers <- all_resolution_markers[[res_col]]
  cl_vals <- seurat_merged@meta.data[[res_col]]
  n_clusters <- length(unique(cl_vals[!is.na(cl_vals)]))
  markers_per_cluster <- max(3, ceiling(80 / n_clusters))

  top_feats <- res_markers %>%
    filter(auc > 0.55, pct_in > 20, logFC > 0) %>%
    group_by(group) %>%
    slice_max(order_by = auc, n = markers_per_cluster) %>%
    pull(feature) %>%
    unique()

  cat(sprintf("**%d clusters | %d markers (≤%d per cluster)**\n\n",
              n_clusters, length(top_feats), markers_per_cluster))

  # auc_score_norm summary for this resolution (from quality_rows if available)
  if (!is.null(quality_rows)) {
    qr <- quality_rows[abs(quality_rows$resolution - res) < 0.001, ]
    if (nrow(qr) > 0) {
      cat(sprintf("auc_score_norm — mean: %.3f | median: %.3f | fraction ≥ 0.5: %.0f%%\n\n",
                  mean(qr$auc_score_norm, na.rm = TRUE),
                  median(qr$auc_score_norm, na.rm = TRUE),
                  100 * mean(qr$auc_score_norm >= 0.5, na.rm = TRUE)))
      # Per-cluster auc table (worst first)
      qr_display <- qr[order(qr$auc_score_norm), c("cluster", "n_cells",
                                                     "auc_score_norm", "strong_markers",
                                                     "top_marker_1", "top_marker_2", "top_marker_3")]
      print(knitr::kable(qr_display, digits = 3,
                         caption = paste0("auc_score_norm per cluster — res ", res)))
      cat("\n\n")
    }
  }

  if (length(top_feats) == 0) {
    cat("No markers passing thresholds (auc > 0.55, pct_in > 20, logFC > 0) — skipping heatmaps.\n\n")
    next
  }

  cells_rand <- sample(colnames(seurat_merged), size = min(5000, ncol(seurat_merged)))
  cells_prop <- subset(seurat_merged, downsample = 1500) %>% colnames()
  Idents(seurat_merged) <- res_col

  tryCatch({
    p_rand <- DoHeatmap(seurat_merged[, cells_rand],
                        features = top_feats,
                        group.by = res_col,
                        slot = "data", size = 3) +
      ggtitle(paste0(COMPARTMENT, " — ", res_col, " — Random Sample")) +
      NoLegend()
    print(p_rand)
  }, error = function(e) cat("Random heatmap failed:", e$message, "\n\n"))

  tryCatch({
    p_prop <- DoHeatmap(seurat_merged[, cells_prop],
                        features = top_feats,
                        group.by = res_col,
                        slot = "data", size = 3) +
      ggtitle(paste0(COMPARTMENT, " — ", res_col, " — Proportional Sample")) +
      NoLegend()
    print(p_prop)
  }, error = function(e) cat("Proportional heatmap failed:", e$message, "\n\n"))

  gc()
}
```

# Save Outputs

```{r save_outputs}
cat("\n=== Saving Outputs ===\n\n")

OUTPUT_BASE <- file.path(STEP_19B_DIR, COMPARTMENT)
dir.create(file.path(OUTPUT_BASE, "markers_by_resolution"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(OUTPUT_BASE, "metadata"), recursive = TRUE, showWarnings = FALSE)

# Save metadata
METADATA_PATH <- file.path(OUTPUT_BASE, "metadata/unified_metadata.csv")
write.csv(seurat_merged@meta.data, METADATA_PATH, row.names = TRUE)
cat("Metadata saved:", METADATA_PATH, "\n")

# Save per-resolution markers
for (res_col in names(all_resolution_markers)) {
  filename <- paste0("markers_", gsub("\\.", "", res_col), ".csv")
  filepath <- file.path(OUTPUT_BASE, "markers_by_resolution", filename)
  write.csv(all_resolution_markers[[res_col]], file = filepath, row.names = FALSE)
  cat("Markers saved:", filename, "\n")
}

# Save cell type markers if available
if (exists("celltype_markers") && !is.null(celltype_markers)) {
  filepath <- file.path(OUTPUT_BASE, "markers_by_resolution/markers_cluster_annotation.csv")
  write.csv(celltype_markers, filepath, row.names = FALSE)
  cat("Cell type markers saved\n")
}

cat("\n✓ Outputs saved to:", OUTPUT_BASE, "\n")
```

```{r session_info}
cat("\nSeurat:", as.character(packageVersion("Seurat")), "\n")
cat("presto:", as.character(packageVersion("presto")), "\n")
```
