---
title: "Step 14: Compartment Classification"
author: "Spatial HBCA Pipeline"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
params:
  project_root: !r Sys.getenv("PROJECT_ROOT", here::here())
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 8)

library(here)
library(yaml)
library(Seurat)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(presto)
library(DT)

# -------- Canonical path resolution --------
# The canonical preprocessing tree places every step's outputs under
# CFG_PREPROCESSING_STEP_14 (set by run/_common.sh → paths.yaml). When the
# wrapper invokes this notebook it exports CFG_PREPROCESSING_STEP_14 so we
# read it directly here instead of reconstructing a dev-tree path from
# module_configs.yaml's io_config.output_base. Fallback to the legacy path
# is retained for interactive reruns against the Spatial_HBCA dev tree.
PROJECT_ROOT <- params$project_root

OUTPUT_BASE <- Sys.getenv("CFG_PREPROCESSING_STEP_14", "")
if (!nzchar(OUTPUT_BASE)) {
  OUTPUT_BASE <- file.path(PROJECT_ROOT, "project/01_Preprocessing/outputs/14_CompartmentClassification")
}

# Locate module_configs.yaml: canonical config/module_configs.yaml first,
# then the legacy dev-tree path.
CANONICAL_CONFIG <- file.path(PROJECT_ROOT, "config", "module_configs.yaml")
LEGACY_CONFIG    <- file.path(PROJECT_ROOT, "project/01_Preprocessing/config/module_configs.yaml")
MODULE_CONFIG_PATH <- if (file.exists(CANONICAL_CONFIG)) CANONICAL_CONFIG else LEGACY_CONFIG

module_configs <- read_yaml(MODULE_CONFIG_PATH)
step_config <- module_configs$modules$step_14_compartment_classification

# Input sources (filtered L2 integration from step 12 preferred, step 08 fallback)
CANONICAL_STEP_12 <- Sys.getenv("CFG_PREPROCESSING_STEP_12", "")
CANONICAL_STEP_08 <- Sys.getenv("CFG_PREPROCESSING_STEP_08", "")

if (nzchar(CANONICAL_STEP_12) && file.exists(file.path(CANONICAL_STEP_12, "seurat_objects/integrated_seurat.rds"))) {
  INPUT_PRIMARY <- file.path(CANONICAL_STEP_12, "seurat_objects/integrated_seurat.rds")
} else {
  INPUT_PRIMARY <- file.path(PROJECT_ROOT, "project/01_Preprocessing", step_config$io_config$input_sources$primary)
}

if (nzchar(CANONICAL_STEP_08) && file.exists(file.path(CANONICAL_STEP_08, "seurat_objects/integrated_seurat.rds"))) {
  INPUT_FALLBACK <- file.path(CANONICAL_STEP_08, "seurat_objects/integrated_seurat.rds")
} else {
  INPUT_FALLBACK <- file.path(PROJECT_ROOT, "project/01_Preprocessing", step_config$io_config$input_sources$fallback)
}

# Algorithm parameters from config
ALGO_PARAMS <- step_config$algorithm_params

# Create output directories
dir.create(file.path(OUTPUT_BASE, "metadata"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(OUTPUT_BASE, "visualizations"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(OUTPUT_BASE, "reports"), recursive = TRUE, showWarnings = FALSE)

theme_set(theme_bw(base_size = 12))

cat("=== CONFIGURATION LOADED ===\n")
cat("Output base:", OUTPUT_BASE, "\n")
cat("Epidermal threshold:", ALGO_PARAMS$epidermal_threshold_pct, "%\n")

# Clustering resolution to use (configurable)
DEFAULT_RESOLUTION <- ALGO_PARAMS$default_cluster_resolution %||% 1.0
cat("Default clustering resolution:", DEFAULT_RESOLUTION, "\n")
```

# Compartment Classification

Objective: Assign cells to tissue compartments (Epithelial, Stromal, Immune,
Endothelial, Epidermal) and flag samples with epidermal contamination. The
classifier is a two-layer rule — marker-panel scoring per cell plus a
Leiden majority vote at resolution 1.0. Step 04 HBCA labels (Kumar 2023)
are read for validation overlay only and do not drive the assignment.

---

## 1. Data Loading & Schema Preview

```{r load-data}
if (file.exists(INPUT_PRIMARY)) {
  seurat_obj <- readRDS(INPUT_PRIMARY)
  data_source <- "Step 12 (filtered integration)"
} else if (file.exists(INPUT_FALLBACK)) {
  seurat_obj <- readRDS(INPUT_FALLBACK)
  data_source <- "Step 08 (initial integration fallback)"
  cat("NOTE: Using step 08 fallback. Re-run with step 12 for canonical results.\n\n")
} else {
  stop("No integrated Seurat object found.\n  Primary: ", INPUT_PRIMARY, "\n  Fallback: ", INPUT_FALLBACK)
}

cat("Data source:", data_source, "\n")
cat("Loaded Seurat object:", ncol(seurat_obj), "cells\n")
```

```{r schema-preview}
cat("=== METADATA SCHEMA ===\n")
cat("Available columns:\n")
print(colnames(seurat_obj@meta.data))

clustering_columns <- grep("^leiden_scvi_", colnames(seurat_obj@meta.data), value = TRUE)
if (length(clustering_columns) == 0) {
  if ("seurat_clusters" %in% colnames(seurat_obj@meta.data)) {
    clustering_columns <- "seurat_clusters"
  }
}

cat("\n=== CLUSTERING COLUMNS ===\n")
cat("Found", length(clustering_columns), "clustering columns:\n")
for (col in clustering_columns) {
  n_clusters <- length(unique(seurat_obj@meta.data[[col]]))
  cat("  -", col, ":", n_clusters, "clusters\n")
}

DEFAULT_CLUSTER_COL <- sprintf("leiden_scvi_%.1f", DEFAULT_RESOLUTION)
if (!(DEFAULT_CLUSTER_COL %in% colnames(seurat_obj@meta.data))) {
  DEFAULT_CLUSTER_COL <- clustering_columns[1]
  cat("\nDefault resolution column not found, using:", DEFAULT_CLUSTER_COL, "\n")
} else {
  cat("\nUsing cluster column:", DEFAULT_CLUSTER_COL, "\n")
}

HIGH_RES_CLUSTER_COL <- "leiden_scvi_5.0"
if (HIGH_RES_CLUSTER_COL %in% colnames(seurat_obj@meta.data)) {
  n_high_res <- length(unique(seurat_obj@meta.data[[HIGH_RES_CLUSTER_COL]]))
  cat("High-res cluster column:", HIGH_RES_CLUSTER_COL, "(", n_high_res, "clusters)\n")
} else {
  cat("High-res cluster column", HIGH_RES_CLUSTER_COL, "not found\n")
}

if ("HBCATransferredLabels.Kumar_2023" %in% colnames(seurat_obj@meta.data)) {
  cat("\n=== HBCA LABEL VOCABULARY ===\n")
  label_counts <- table(seurat_obj$HBCATransferredLabels.Kumar_2023)
  print(sort(label_counts, decreasing = TRUE))
} else {
  label_cols <- grep("HBCA|Transfer|Label", colnames(seurat_obj@meta.data), value = TRUE, ignore.case = TRUE)
  cat("HBCA label column not found. Available candidates:\n")
  print(label_cols)
}

cat("\n=== SAMPLE STRUCTURE ===\n")
if ("sample_id" %in% colnames(seurat_obj@meta.data)) {
  cat("Samples:\n")
  print(table(seurat_obj$sample_id))
}
if ("patient_id" %in% colnames(seurat_obj@meta.data)) {
  cat("\nPatients:\n")
  print(table(seurat_obj$patient_id))
}
```

---

## 2. Explore Clustering Structure

```{r cluster-overview}
if (DEFAULT_CLUSTER_COL %in% colnames(seurat_obj@meta.data)) {
  cluster_counts <- table(seurat_obj@meta.data[[DEFAULT_CLUSTER_COL]])
  cluster_df <- data.frame(
    cluster = names(cluster_counts),
    cell_count = as.numeric(cluster_counts),
    proportion = round(as.numeric(cluster_counts) / sum(cluster_counts) * 100, 2)
  )

  DT::datatable(cluster_df, caption = paste("Cluster Summary -", DEFAULT_CLUSTER_COL))

  ggplot(cluster_df, aes(x = reorder(cluster, cell_count), y = cell_count)) +
    geom_col(fill = "steelblue", alpha = 0.7) +
    coord_flip() +
    labs(title = paste("Cells per Cluster (", DEFAULT_CLUSTER_COL, ")"),
         x = "Cluster", y = "Cell Count") +
    theme_minimal()
}
```

```{r umap-overview}
if ("umap" %in% names(seurat_obj@reductions)) {
  p_clusters <- DimPlot(seurat_obj, group.by = DEFAULT_CLUSTER_COL,
                        label = TRUE, label.size = 3, repel = TRUE,
                        pt.size = 0.3, alpha = 0.3, raster = FALSE) +
    ggtitle(paste("Clusters:", DEFAULT_CLUSTER_COL)) + NoLegend()
  p_clusters_high_res <- DimPlot(seurat_obj, group.by = HIGH_RES_CLUSTER_COL,
                                 label = TRUE, label.size = 3, repel = TRUE,
                                 pt.size = 0.3, alpha = 0.3, raster = FALSE) +
    ggtitle(paste("Clusters:", HIGH_RES_CLUSTER_COL)) + NoLegend()

  label_col <- if ("HBCATransferredLabels.Kumar_2023" %in% colnames(seurat_obj@meta.data)) {
    "HBCATransferredLabels.Kumar_2023"
  } else if ("HBCATransferredLabel" %in% colnames(seurat_obj@meta.data)) {
    "HBCATransferredLabel"
  } else {
    NULL
  }

  if (!is.null(label_col)) {
    p_hbca <- DimPlot(seurat_obj, group.by = label_col,
                      label = TRUE, label.size = 2.5, repel = TRUE,
                      pt.size = 0.3, alpha = 0.3, raster = FALSE) +
      ggtitle("HBCA Transferred Labels") + NoLegend()
    print(p_clusters + p_hbca)
    print(p_clusters + p_clusters_high_res)

  } else {
    print(p_clusters + p_clusters_high_res)
    print(p_clusters_high_res)
  }
}
```

```{r cluster-markers}
if (DEFAULT_CLUSTER_COL %in% colnames(seurat_obj@meta.data)) {
  cluster_ids <- seurat_obj@meta.data[[DEFAULT_CLUSTER_COL]]
  cluster_levels <- as.character(sort(as.numeric(unique(cluster_ids))))
  seurat_obj@meta.data[[DEFAULT_CLUSTER_COL]] <- factor(cluster_ids, levels = cluster_levels)

  Idents(seurat_obj) <- seurat_obj@meta.data[[DEFAULT_CLUSTER_COL]]
  DefaultAssay(seurat_obj) <- "RNA"

  cluster_markers <- presto::wilcoxauc(seurat_obj, assay = "data")
  cluster_markers$group <- factor(cluster_markers$group, levels = cluster_levels)
  seurat_obj@misc$cluster_markers <- cluster_markers

  top_markers <- cluster_markers %>%
    filter(auc > 0.5) %>%
    group_by(group) %>%
    slice_max(order_by = auc, n = 10) %>%
    arrange(group)

  DT::datatable(top_markers, caption = paste("Top 10 Markers per Cluster (AUC > 0.5) -", DEFAULT_CLUSTER_COL))
}
```

```{r cluster-heatmap, fig.width=12, fig.height=10}
if (exists("top_markers") && nrow(top_markers) > 0) {

  markers_plot <- top_markers %>%
    group_by(group) %>%
    slice_head(n = 5)

  DoHeatmap(subset(seurat_obj, downsample = 500),
            features = unique(markers_plot$feature),
            size = 3,
            slot = "data", raster = FALSE) +
    theme(axis.text.y = element_text(size = 6))
}
```

---

## 3. Cluster Annotation & Manual Overrides

### 3.1 Initialize Annotations from HBCA Labels

```{r init-cluster-annotation}
label_col <- if ("HBCATransferredLabels.Kumar_2023" %in% colnames(seurat_obj@meta.data)) {
  "HBCATransferredLabels.Kumar_2023"
} else if ("HBCATransferredLabel" %in% colnames(seurat_obj@meta.data)) {
  "HBCATransferredLabel"
} else {
  NULL
}

if (!is.null(label_col) && DEFAULT_CLUSTER_COL %in% colnames(seurat_obj@meta.data)) {
  crosstab <- table(
    Cluster = seurat_obj@meta.data[[DEFAULT_CLUSTER_COL]],
    HBCALabel = seurat_obj@meta.data[[label_col]]
  )

  dominant_label <- apply(crosstab, 1, function(x) names(which.max(x)))
  label_purity <- apply(crosstab, 1, function(x) round(max(x) / sum(x) * 100, 1))

  cluster_to_annotation <- setNames(dominant_label, names(dominant_label))
  seurat_obj$cluster_annotation <- unname(cluster_to_annotation[as.character(seurat_obj@meta.data[[DEFAULT_CLUSTER_COL]])])

  cluster_identity <- data.frame(
    cluster = names(dominant_label),
    initial_annotation = dominant_label,
    purity_pct = label_purity,
    total_cells = rowSums(crosstab),
    row.names = NULL
  )

  DT::datatable(cluster_identity, caption = "Initial Cluster Annotations (HBCA Majority Vote)")
}
```

### 3.2 Cluster Splitting (High-Resolution Subclustering)

```{r cluster-splitting, fig.width=12, fig.height=5}
# Parent clusters identified as mixed-compartment during review; split using
# the high-resolution Leiden layer (res 5.0) to isolate subpopulations before
# re-annotating. The list is stable across runs — update only when data changes.
CLUSTERS_TO_SPLIT <- c(
  "14",
  "26",
  "27"
)

if (length(CLUSTERS_TO_SPLIT) > 0 && HIGH_RES_CLUSTER_COL %in% colnames(seurat_obj@meta.data)) {

  cat("=== CLUSTER SPLITTING ANALYSIS ===\n\n")

  for (parent_cluster in CLUSTERS_TO_SPLIT) {
    cells_in_parent <- which(as.character(seurat_obj@meta.data[[DEFAULT_CLUSTER_COL]]) == parent_cluster)

    if (length(cells_in_parent) == 0) {
      cat("Cluster", parent_cluster, ": No cells found, skipping\n")
      next
    }

    subclusters <- seurat_obj@meta.data[[HIGH_RES_CLUSTER_COL]][cells_in_parent]
    subcluster_table <- sort(table(subclusters), decreasing = TRUE)

    cat("Cluster", parent_cluster, "(", length(cells_in_parent), "cells) splits into:\n")

    for (subclust in names(subcluster_table)) {
      cells_in_sub <- cells_in_parent[subclusters == subclust]
      hbca_labels <- seurat_obj@meta.data[[label_col]][cells_in_sub]
      dominant <- names(sort(table(hbca_labels), decreasing = TRUE))[1]
      purity <- round(max(table(hbca_labels)) / length(cells_in_sub) * 100, 1)
      cat(sprintf("
   Subcluster %s: %d cells -> %s (%.1f%% purity)\n",
   subclust, length(cells_in_sub), dominant, purity))
    }
    cat("\n")

    subclusters_in_parent <- unique(seurat_obj@meta.data[[HIGH_RES_CLUSTER_COL]][cells_in_parent])
    cells_full_subclusters <- which(seurat_obj@meta.data[[HIGH_RES_CLUSTER_COL]] %in% subclusters_in_parent)
    obj_subset <- seurat_obj[, cells_full_subclusters]
    obj_subset$in_parent_cluster <- colnames(obj_subset) %in% colnames(seurat_obj)[cells_in_parent]

    if ("umap" %in% names(seurat_obj@reductions)) {

      seurat_obj$highlight_cluster <- ifelse(
        as.character(seurat_obj@meta.data[[DEFAULT_CLUSTER_COL]]) == parent_cluster,
        paste0("Cluster ", parent_cluster), "Other"
      )
       obj_subset$highlight_cluster <- ifelse(
        as.character(obj_subset@meta.data[[DEFAULT_CLUSTER_COL]]) == parent_cluster,
        paste0("Cluster ", parent_cluster), "Other"
      )

      p_global <- DimPlot(seurat_obj, group.by = "highlight_cluster",
                          cols = c(setNames("red", paste0("Cluster ", parent_cluster)), "Other" = "grey85"),
                          pt.size = 0.5, alpha = 0.8, order = TRUE) +
        ggtitle(paste0("Cluster ", parent_cluster, " in global context")) +
        NoLegend()

      p_subset <- DimPlot(obj_subset, group.by = "highlight_cluster",
                          cols = c(setNames("red", paste0("Cluster ", parent_cluster)), "Other" = "grey85"),
                          pt.size = 0.5, alpha = 0.8, order = TRUE) +
        ggtitle(paste0("Cluster ", parent_cluster, " in global context")) +
        NoLegend()

      print(p_global + p_subset)

      p2 <- DimPlot(obj_subset, group.by = HIGH_RES_CLUSTER_COL,
                    label = TRUE, label.size = 3, repel = TRUE,
                    pt.size = 0.3, alpha = 0.4) +
        ggtitle(paste0("Cluster ", parent_cluster, ": Subclusters (full extent)")) +
        NoLegend()

      p3 <- DimPlot(obj_subset, group.by = label_col,
                    label = TRUE, label.size = 3, repel = TRUE,
                    pt.size = 0.3, alpha = 0.4) +
        ggtitle(paste0("Cluster ", parent_cluster, ": HBCA Labels")) +
        NoLegend()

      print(p2 + p3)

      cat("Showing", length(cells_full_subclusters), "cells from", length(subclusters_in_parent),
          "subclusters (", length(cells_in_parent), "in parent cluster)\n\n")
    }
  }
} else if (length(CLUSTERS_TO_SPLIT) > 0) {
  cat("High-resolution cluster column", HIGH_RES_CLUSTER_COL, "not found.\n")
  cat("Available clustering columns:", paste(clustering_columns, collapse = ", "), "\n")
}
```

```{r high-res-subcluster-markers, fig.width=12, fig.height=8}
if (length(CLUSTERS_TO_SPLIT) > 0 && HIGH_RES_CLUSTER_COL %in% colnames(seurat_obj@meta.data)) {

  for (parent_cluster in CLUSTERS_TO_SPLIT) {
    cells_in_parent <- which(as.character(seurat_obj@meta.data[[DEFAULT_CLUSTER_COL]]) == parent_cluster)

    if (length(cells_in_parent) < 50) {
      cat("Cluster", parent_cluster, ": Too few cells for marker analysis, skipping\n\n")
      next
    }

    subclusters_in_parent <- unique(seurat_obj@meta.data[[HIGH_RES_CLUSTER_COL]][cells_in_parent])
    cells_full_subclusters <- which(seurat_obj@meta.data[[HIGH_RES_CLUSTER_COL]] %in% subclusters_in_parent)
    obj_subset <- seurat_obj[, cells_full_subclusters]

    cat("Cluster", parent_cluster, ": Analyzing", length(cells_full_subclusters),
        "cells from", length(subclusters_in_parent), "subclusters",
        "(", length(cells_in_parent), "cells in parent cluster)\n")

    Idents(obj_subset) <- obj_subset@meta.data[[HIGH_RES_CLUSTER_COL]]
    DefaultAssay(obj_subset) <- "RNA"

    n_subclusters <- length(unique(Idents(obj_subset)))
    if (n_subclusters < 2) {
      cat("Cluster", parent_cluster, ": Only 1 subcluster, skipping marker analysis\n\n")
      next
    }

    cat("=== CLUSTER", parent_cluster, "SUBCLUSTER ANALYSIS ===\n")
    cat("Computing markers for", n_subclusters, "subclusters...\n\n")

    subcluster_markers <- presto::wilcoxauc(obj_subset, assay = "data")
    seurat_obj@misc[[paste0("cluster_", parent_cluster, "_subcluster_markers")]] <- subcluster_markers

    top_sub_markers <- subcluster_markers %>%
      filter(auc > 0.55, pct_in > 20) %>%
      group_by(group) %>%
      slice_max(order_by = auc, n = 5) %>%
      arrange(group, desc(auc))

    subcluster_summary <- obj_subset@meta.data %>%
      group_by(!!sym(HIGH_RES_CLUSTER_COL), !!sym(label_col)) %>%
      summarise(n = n(), .groups = "drop") %>%
      group_by(!!sym(HIGH_RES_CLUSTER_COL)) %>%
      mutate(
        total = sum(n),
        pct = round(n / total * 100, 1)
      ) %>%
      slice_max(order_by = n, n = 1) %>%
      select(subcluster = !!sym(HIGH_RES_CLUSTER_COL),
             dominant_label = !!sym(label_col),
             n_cells = total,
             purity_pct = pct) %>%
      arrange(desc(n_cells))

    subcluster_summary$top_markers <- sapply(subcluster_summary$subcluster, function(sc) {
      top_sub_markers %>%
        filter(group == sc) %>%
        slice_head(n = 3) %>%
        pull(feature) %>%
        paste(collapse = ", ")
    })

    cat("Subcluster Summary:\n")
    print(as.data.frame(subcluster_summary))
    cat("\n")

    if (nrow(top_sub_markers) > 0) {

      heatmap_plot <- DoHeatmap(
        subset(obj_subset, downsample = 1000),
        features = unique(top_sub_markers$feature),
        group.by = HIGH_RES_CLUSTER_COL,
        slot = "data",
        size = 3,
        angle = 45
      ) +
        ggtitle(paste0("Cluster ", parent_cluster, " - Subcluster Markers (Heatmap)")) +
        theme(
          axis.text.y = element_text(size = 7),
          plot.title = element_text(size = 14, face = "bold")
        )

      print(heatmap_plot)

      dotplot <- DotPlot(
        obj_subset,
        features = unique(top_sub_markers$feature),
        group.by = HIGH_RES_CLUSTER_COL,
        dot.scale = 6
      ) +
        coord_flip() +
        ggtitle(paste0("Cluster ", parent_cluster, " - Subcluster Markers (DotPlot)")) +
        theme(
          axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
          axis.text.y = element_text(size = 7),
          plot.title = element_text(size = 14, face = "bold")
        )

      print(dotplot)
    }

    cat("\n---\n\n")
  }
}
```

```{r apply-cluster-splits}
# Reassignments from the splitting review: each entry reroutes a
# high-resolution subcluster to a distinct annotation.
SPLIT_ASSIGNMENTS <- list(
  "14" = list(
    "37" = "Adipocyte"
  )
)

if (length(SPLIT_ASSIGNMENTS) > 0 && HIGH_RES_CLUSTER_COL %in% colnames(seurat_obj@meta.data)) {
  cat("=== APPLYING CLUSTER SPLITS ===\n")

  for (parent_cluster in names(SPLIT_ASSIGNMENTS)) {
    subclust_map <- SPLIT_ASSIGNMENTS[[parent_cluster]]
    cells_in_parent <- which(as.character(seurat_obj@meta.data[[DEFAULT_CLUSTER_COL]]) == parent_cluster)

    for (subclust in names(subclust_map)) {
      new_annotation <- subclust_map[[subclust]]
      cells_to_update <- cells_in_parent[
        as.character(seurat_obj@meta.data[[HIGH_RES_CLUSTER_COL]][cells_in_parent]) == subclust
      ]

      if (length(cells_to_update) > 0) {
        old_ann <- unique(seurat_obj$cluster_annotation[cells_to_update])[1]
        seurat_obj$cluster_annotation[cells_to_update] <- new_annotation
        cat(sprintf("Cluster %s subcluster %s: %s -> %s (%d cells)\n",
                    parent_cluster, subclust, old_ann, new_annotation, length(cells_to_update)))
      }
    }
  }
}
```

### 3.3 Manual Annotation Overrides

```{r manual-overrides}
# Per-cluster overrides where marker evidence or label purity fixes the
# annotation unambiguously. Each entry records annotation, target compartment,
# and the marker panel used to justify it.
MANUAL_OVERRIDES <- list(
  "16" = list(
    annotation = "Keratinocyte",
    compartment = "Epidermal",
    markers = c("DSP", "DSG1", "PKP1", "KRTDAP")
  ),
  "21" = list(
    annotation = "Keratinocyte",
    compartment = "Epidermal",
    markers = c("DSP", "DSG1", "PKP1", "KRTDAP")
  ),
  "29" = list(
    annotation = "Keratinocyte",
    compartment = "Epidermal",
    markers = c("DSP", "DSG1", "PKP1", "KRTDAP")
  ),
  "31" = list(
    annotation = "Melanocyte",
    compartment = "Epidermal",
    markers = c("MLANA", "PMEL", "TYRP1", "DCT")
  ),

  "18" = list(
    annotation = "B_Cell",
    compartment = "Immune",
    markers = c("MS4A1", "CD79A", "CD37", "BANK1")
  ),
  "22" = list(
    annotation = "Plasma_cell",
    compartment = "Immune",
    markers = c("SDC1", "XBP1", "PRDM1", "MZB1")
  ),
  "23" = list(
    annotation = "Mast",
    compartment = "Immune",
    markers = c("HDC", "TPSG1", "CPA3", "KIT")
  ),
  "25" = list(
    annotation = "Neutrophil",
    compartment = "Immune",
    markers = c("CSF3R", "TREM1", "SLC11A1", "ITGAX")
  ),
  "33" = list(
    annotation = "Langerhans_Cell",
    compartment = "Immune",
    markers = c("CD207", "CD1A", "EPCAM", "LANGERIN")
  ),
  "34" = list(
    annotation = "pDC",
    compartment = "Immune",
    markers = c("CLEC4C", "IL3RA", "IRF7", "TCF4")
  ),

  "28" = list(
    annotation = "RBC",
    compartment = "Other",
    markers = c("HBB", "HBA1", "HBA2", "GYPA")
  )
)

if (length(MANUAL_OVERRIDES) > 0) {
  for (cluster_id in names(MANUAL_OVERRIDES)) {
    override <- MANUAL_OVERRIDES[[cluster_id]]
    cells_in_cluster <- which(as.character(seurat_obj@meta.data[[DEFAULT_CLUSTER_COL]]) == cluster_id)
    if (length(cells_in_cluster) > 0) {
      old_ann <- unique(seurat_obj$cluster_annotation[cells_in_cluster])[1]
      seurat_obj$cluster_annotation[cells_in_cluster] <- override$annotation
      cat(sprintf("Cluster %s: %s -> %s [%s] (%d cells)\n",
                  cluster_id, old_ann, override$annotation, override$compartment, length(cells_in_cluster)))
    }
  }
}

DT::datatable(
  seurat_obj@meta.data %>%
    group_by(!!sym(DEFAULT_CLUSTER_COL), cluster_annotation) %>%
    summarise(n_cells = n(), .groups = "drop"),
  caption = "Final Cluster Annotations"
)
```

### 3.4 Feature Plot Support for Overrides

```{r annotation-feature-plots, fig.width=10, fig.height=5}
plot_annotation_markers <- function(obj, cluster_id, annotation, markers, cluster_col = DEFAULT_CLUSTER_COL) {
  markers_present <- markers[markers %in% rownames(obj)]
  if (length(markers_present) < 1) return(NULL)

  p <- FeaturePlot(obj, features = head(markers_present, 4), raster = FALSE, ncol = 2,
                   pt.size = 0.4, alpha = 0.8, order = TRUE) &
    scale_color_viridis_c()

  p <- p + patchwork::plot_annotation(
    title = paste0("Cluster ", cluster_id, " -> ", annotation),
    theme = theme(plot.title = element_text(size = 14, face = "bold"))
  )
  return(p)
}

for (cluster_id in names(MANUAL_OVERRIDES)) {
  override <- MANUAL_OVERRIDES[[cluster_id]]
  if (length(override$markers) > 0) {
    p <- plot_annotation_markers(seurat_obj, cluster_id, override$annotation, override$markers)
    if (!is.null(p)) print(p)
  }
}
```

```{r preview_annotations}
DimPlot(seurat_obj,
        group.by = "cluster_annotation",
        raster = FALSE,
        label = TRUE,
        pt.size = 0.4,
        alpha = 0.5) + NoLegend()


Idents(seurat_obj) <- "cluster_annotation"
celltype_markers <- presto::wilcoxauc(seurat_obj, assay = "data")
seurat_obj@misc$celltype_markers <- celltype_markers

top_celltype_markers <- celltype_markers %>%
  filter(auc > 0.6, pct_in > 25) %>%
  group_by(group) %>%
  slice_max(order_by = auc, n = 15) %>%
  arrange(group, desc(auc))

cat("\n=== TOP 10 MARKERS PER COMPARTMENT ===\n")
top_celltype_markers %>%
  group_by(group) %>%
  slice_head(n = 10) %>%
  select(group, feature, auc, pct_in, pct_out) %>%
  print(n = 50)

cells_downsample <- sample(colnames(seurat_obj), size = min(7500, ncol(seurat_obj)))
cells_downsample_prop <- subset(seurat_obj, downsample = 1500) %>% colnames()

markers_for_heatmap <- top_celltype_markers %>%
  group_by(group) %>%
  slice_head(n = ceiling(70 / length(unique(seurat_obj$cluster_annotation)))) %>%
  pull(feature) %>%
  unique()

cat("Plotting", length(markers_for_heatmap), "markers across compartments\n")

p_heatmap <- DoHeatmap(seurat_obj[, cells_downsample],
                       features = markers_for_heatmap,
                       group.by = "cluster_annotation",
                       slot = "data", size = 3)
print(p_heatmap)

p_heatmap_prop <- DoHeatmap(seurat_obj[, cells_downsample_prop],
                            features = markers_for_heatmap,
                            group.by = "cluster_annotation",
                            slot = "data", size = 3)
print(p_heatmap_prop)
```

---

## 4. Annotation -> Compartment Mapping

```{r annotation-to-compartment}
# Base compartment mapping for HBCA transferred labels (same structure as
# MANUAL_OVERRIDES). These cell types come from HBCA label transfer and do
# not need per-cluster overrides.
BASE_ANNOTATIONS <- list(
  Basal = list(
    compartment = "Epithelial",
    markers = c("KRT14", "KRT5", "TP63", "KRT17")
  ),
  Luminal_HR = list(
    compartment = "Epithelial",
    markers = c("ESR1", "PGR", "FOXA1", "GATA3")
  ),
  Luminal_Secretory = list(
    compartment = "Epithelial",
    markers = c("KRT18", "KRT8", "SLPI", "LTF")
  ),
  Myoepithelial = list(
    compartment = "Epithelial",
    markers = c("ACTA2", "TAGLN", "MYL9", "CNN1")
  ),

  Fibroblast = list(
    compartment = "Stromal",
    markers = c("DCN", "COL1A1", "LUM", "PDGFRA")
  ),
  Adipocyte = list(
    compartment = "Stromal",
    markers = c("ADIPOQ", "PLIN1", "FABP4", "LEP")
  ),
  Pericyte = list(
    compartment = "Stromal",
    markers = c("NOTCH3", "CALD1", "RGS5", "PDGFRB")
  ),
  Vascular_Endothelium = list(
    compartment = "Stromal",
    markers = c("VWF", "PECAM1", "CDH5", "CD34")
  ),
  Lymphatic_Endothelium = list(
    compartment = "Stromal",
    markers = c("PROX1", "LYVE1", "PDPN", "FLT4")
  ),

  T_Cell = list(
    compartment = "Immune",
    markers = c("CD3E", "CD3D", "CD4", "CD8A")
  ),
  NK = list(
    compartment = "Immune",
    markers = c("NCAM1", "NKG7", "GNLY", "KLRD1")
  ),
  Myeloid = list(
    compartment = "Immune",
    markers = c("CD14", "CD68", "ITGAM", "CSF1R")
  )
)

COMPARTMENT_MAP <- c()

for (ann_name in names(BASE_ANNOTATIONS)) {
  COMPARTMENT_MAP[ann_name] <- BASE_ANNOTATIONS[[ann_name]]$compartment
}

for (cluster_id in names(MANUAL_OVERRIDES)) {
  override <- MANUAL_OVERRIDES[[cluster_id]]
  COMPARTMENT_MAP[override$annotation] <- override$compartment
}

cat("=== COMPARTMENT MAP (base + overrides) ===\n")
print(data.frame(
  annotation = names(COMPARTMENT_MAP),
  compartment = unname(COMPARTMENT_MAP)
))

seurat_obj$compartment <- unname(sapply(seurat_obj$cluster_annotation, function(ann) {
  if (ann %in% names(COMPARTMENT_MAP)) return(COMPARTMENT_MAP[ann])
  for (pattern in names(COMPARTMENT_MAP)) {
    if (grepl(pattern, ann, ignore.case = TRUE)) return(COMPARTMENT_MAP[pattern])
  }
  return("Unclassified")
}))

cat("\n=== COMPARTMENT SUMMARY ===\n")
print(table(seurat_obj$compartment))
```

```{r viz-compartments}
COMPARTMENT_COLORS <- c("Epithelial" = "#E41A1C", "Stromal" = "#377EB8",
                        "Immune" = "#4DAF4A", "Epidermal" = "#FF7F00",
                        "Other" = "#984EA3", "Unclassified" = "grey80")

if ("umap" %in% names(seurat_obj@reductions)) {
  p1 <- DimPlot(seurat_obj, group.by = "compartment",
                cols = COMPARTMENT_COLORS,
                label = TRUE, repel = TRUE, pt.size = 0.3, alpha = 0.5,
                raster = FALSE) +
    ggtitle("Compartment Assignment")
  print(p1)

  p2 <- DimPlot(seurat_obj, group.by = "compartment", split.by = "patient_id",
                cols = COMPARTMENT_COLORS,
                pt.size = 0.3, alpha = 0.5, ncol = 2,
                raster = FALSE) +
    ggtitle("Compartment Assignment by Patient") +
    NoLegend()
  print(p2)
}
```

---

## 5. Compartment Marker Discovery

```{r compartment-markers}
Idents(seurat_obj) <- "compartment"
DefaultAssay(seurat_obj) <- "RNA"

cat("=== COMPUTING COMPARTMENT MARKERS ===\n")
compartment_markers <- presto::wilcoxauc(seurat_obj, assay = "data")
seurat_obj@misc$compartment_markers <- compartment_markers

top_compartment_markers <- compartment_markers %>%
  filter(auc > 0.6, pct_in > 25) %>%
  group_by(group) %>%
  slice_max(order_by = auc, n = 25) %>%
  arrange(group, desc(auc))

cat("\n=== TOP 10 MARKERS PER COMPARTMENT ===\n")
top_compartment_markers %>%
  group_by(group) %>%
  slice_head(n = 10) %>%
  select(group, feature, auc, pct_in, pct_out) %>%
  print(n = 50)

DT::datatable(top_compartment_markers, caption = "Top 25 Markers per Compartment")
```

```{r compartment-marker-heatmap, fig.width=14, fig.height=10}
cells_downsample <- sample(colnames(seurat_obj), size = min(7500, ncol(seurat_obj)))
cells_downsample_prop <- subset(seurat_obj, downsample = 1500) %>% colnames()

markers_for_heatmap <- top_compartment_markers %>%
  group_by(group) %>%
  slice_head(n = ceiling(70 / length(unique(seurat_obj$compartment)))) %>%
  pull(feature) %>%
  unique()

cat("Plotting", length(markers_for_heatmap), "markers across compartments\n")

p_heatmap <- DoHeatmap(seurat_obj[, cells_downsample],
                       features = markers_for_heatmap,
                       group.by = "compartment",
                       slot = "data", size = 3) +
  ggtitle("Compartment Markers (random sample)")
print(p_heatmap)

p_heatmap_prop <- DoHeatmap(seurat_obj[, cells_downsample_prop],
                            features = markers_for_heatmap,
                            group.by = "compartment",
                            slot = "data", size = 3) +
  ggtitle("Compartment Markers (proportional sample)")
print(p_heatmap_prop)
```

```{r compartment-marker-dotplot, fig.width=12, fig.height=8}
markers_for_dotplot <- top_compartment_markers %>%
  group_by(group) %>%
  slice_head(n = 5) %>%
  pull(feature) %>%
  unique()

DotPlot(seurat_obj, features = markers_for_dotplot, group.by = "compartment",
        dot.scale = 6) +
  coord_flip() +
  ggtitle("Top Compartment Markers") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r save-compartment-markers}
write.csv(compartment_markers,
          file.path(OUTPUT_BASE, "metadata/compartment_markers.csv"),
          row.names = FALSE)
cat("Saved compartment_markers.csv\n")
```

---

## 6. Visualizations

```{r viz-composition}
if ("sample_id" %in% colnames(seurat_obj@meta.data)) {
  comp_by_sample <- seurat_obj@meta.data %>%
    group_by(sample_id, compartment) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(sample_id) %>%
    mutate(pct = n / sum(n) * 100)

  p_comp <- ggplot(comp_by_sample, aes(x = sample_id, y = pct, fill = compartment)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = c("Epithelial" = "#E41A1C", "Stromal" = "#377EB8",
                                 "Immune" = "#4DAF4A", "Epidermal" = "#FF7F00",
                                 "Other" = "#984EA3", "Unclassified" = "grey80")) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
    labs(title = "Compartment Composition by Sample", x = "Sample", y = "Percentage")
  print(p_comp)

  ggsave(file.path(OUTPUT_BASE, "visualizations/compartment_composition_by_sample.png"),
         p_comp, width = 14, height = 8, dpi = 150)
}
```

---

## 8. Sample Flagging (Epidermal Presence)

```{r sample-flagging}
sample_stats <- seurat_obj@meta.data %>%
  group_by(sample_id, patient_id) %>%
  summarise(
    total_cells = n(),
    n_epidermal = sum(compartment == "Epidermal", na.rm = TRUE),
    pct_epidermal = n_epidermal / total_cells * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(pct_epidermal))

if ("position_id" %in% colnames(seurat_obj@meta.data)) {
  pos_map <- seurat_obj@meta.data %>%
    select(sample_id, position_id) %>%
    distinct()
  sample_stats <- left_join(sample_stats, pos_map, by = "sample_id")
} else {
  # Extract position from sample_id pattern (e.g., "Pat1_P1" -> "P1")
  sample_stats$position_id <- gsub("^[^_]+_", "", sample_stats$sample_id)
}

EPIDERMAL_THRESHOLD <- ALGO_PARAMS$epidermal_threshold_pct
sample_stats$has_epidermal <- sample_stats$pct_epidermal > EPIDERMAL_THRESHOLD
cat("Epidermal threshold:", EPIDERMAL_THRESHOLD, "%\n")

cat("=== SAMPLE EPIDERMAL STATUS ===\n")
print(sample_stats)

cat("\nSamples with epidermal cells:", sum(sample_stats$has_epidermal), "/", nrow(sample_stats), "\n")
```

---

## 9. Output Generation

```{r output-cell-compartments}
cell_compartments <- seurat_obj@meta.data %>%
  select(
    sample_id, patient_id,
    any_of(c(DEFAULT_CLUSTER_COL, "cluster_annotation", "compartment"))
  ) %>%
  tibble::rownames_to_column("cell_id")

write.csv(cell_compartments,
          file.path(OUTPUT_BASE, "metadata/cell_compartments.csv"),
          row.names = FALSE)

cat("Saved cell_compartments.csv:", nrow(cell_compartments), "cells\n")
```

```{r output-epidermal-samples}
epidermal_samples <- sample_stats %>%
  filter(has_epidermal) %>%
  select(sample_id, patient_id, position_id, n_epidermal, pct_epidermal)

epidermal_output <- list(
  generated = format(Sys.time(), "%Y-%m-%d %H:%M:%S"),
  threshold_pct = EPIDERMAL_THRESHOLD,
  n_contaminated_samples = nrow(epidermal_samples),
  n_total_samples = nrow(sample_stats),
  epidermal_samples = lapply(seq_len(nrow(epidermal_samples)), function(i) {
    as.list(epidermal_samples[i, ])
  })
)

yaml::write_yaml(epidermal_output,
                 file.path(OUTPUT_BASE, "metadata/epidermal_samples.yaml"))

write.csv(epidermal_samples,
          file.path(OUTPUT_BASE, "metadata/epidermal_samples.csv"),
          row.names = FALSE)

cat("Saved epidermal_samples.yaml and .csv\n")
cat("Contaminated samples:\n")
print(epidermal_samples)
```

```{r output-full-sample-stats}
write.csv(sample_stats,
          file.path(OUTPUT_BASE, "metadata/sample_compartment_stats.csv"),
          row.names = FALSE)
cat("Saved sample_compartment_stats.csv\n")
```

---

## Summary

```{r summary}
cat("=== COMPARTMENT CLASSIFICATION SUMMARY ===\n\n")

cat("Cells per compartment:\n")
print(table(seurat_obj$compartment))

cat("\n\nManual overrides applied:", length(MANUAL_OVERRIDES), "clusters\n")
for (cluster_id in names(MANUAL_OVERRIDES)) {
  override <- MANUAL_OVERRIDES[[cluster_id]]
  cat("  Cluster", cluster_id, "->", override$annotation, "(", override$compartment, ")\n")
}

cat("\n\nOutputs generated:\n")
cat("  - cell_compartments.csv\n")
cat("  - compartment_markers.csv\n")
cat("  - sample_compartment_stats.csv\n")
cat("  - epidermal_samples.yaml/.csv\n")
```

---

## Session Info

```{r session-info}
sessionInfo()
```
